---
title: "1.2_Gene_expression_analysis"
author: "Fay"
date: '2022-07-27'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1.  Gene expression in the laboratory infections - Heatmap

```{r load_libraries, echo=FALSE, include = FALSE}
library(pheatmap)
library(tidyverse)
library(matrixStats)
library(tidyr)
library(janitor)
library(tibble)
library(corrplot)
library(RColorBrewer)
library(ggplot2)
library(dplyr)
library(missMDA) 
library(FactoMineR)
library(PerformanceAnalytics)
library(factoextra)
library(MASS)
library(reshape2)
library(cowplot)
library(lmtest)
if(any(grepl("package:plyr", search()))) detach("package:plyr") else message("plyr not loaded")
```

```{r}
field <- read.csv("output_data/gene_expression/data_products/field_imputed_gene_expression.csv")

Genes_field   <- c("IFNy",  "CXCR3", "IL.6", #"GBP2", "IL.12", "IRG6",
                     "IL.10", "IL.13", "IL1RN",
                     "CXCR3", "CASP1", "CXCL9", 
                     "IDO1", "IRGM1", "MPO", "MUC2", "MUC5AC", "MYD88", 
                     "NCR1", "PRF1", "RETNLB", "SOCS1", "TICAM1", "TNF")
```

```{r}
#filter out weird ifn y values
field <- field %>% 
  filter(!IFNy >=40)

# turn the data frame into a matrix and transpose it. We want to have each cell 
 # type as a row name 
 gene <- t(as.matrix(field %>% dplyr::select(c(Mouse_ID, all_of(Genes_field)))))
 
 #switch the matrix back to a data frame format
 gene <- as.data.frame(gene)
 
 # turn the first row into column names
 gene %>%
     row_to_names(row_number = 1) -> heatmap_data
 
 table(rowSums(is.na(heatmap_data)) == nrow(heatmap_data))
 
 # turn the columns to numeric other wise the heatmap function will not work
 heatmap_data[] <- lapply(heatmap_data, function(x) as.numeric(as.character(x)))
 
 # remove columns with only NAs 
 heatmap_data <- Filter(function(x)!all(is.na(x)), heatmap_data) 
 
 #remove rows with only Nas
 heatmap_data <-  heatmap_data[, colSums(is.na(heatmap_data)) != 
                                 nrow(heatmap_data)]
 
 ### Prepare the annotation data frame for the heatmap
 
annotation_df <- as_tibble(field) %>%
  dplyr::group_by(Mouse_ID) %>%
  dplyr::select(c("Mouse_ID", "Sex", "HI")) %>%
  dplyr::filter(Mouse_ID %in% colnames(heatmap_data))
  
annotation_df <- unique(annotation_df)
 
annotation_df <- as.data.frame(unique(annotation_df)) %>%
  dplyr::select(-Mouse_ID)


### Prepare the annotation columns for the heatmap
rownames(annotation_df) <- annotation_df$Mouse_ID
# Match the row names to the heatmap data frame
rownames(annotation_df) <- colnames(heatmap_data)

heatmap_data <- as.matrix(heatmap_data)
```


Heatmap on gene expression data: 

```{r pheatmap_genes, echo = FALSE}


pheatmap(heatmap_data, annotation_col = annotation_df, scale = "row")
```

# Heatmap with EIm mc
```{r}


# turn the data frame into a matrix and transpose it. We want to have each cell 
 # type as a row name 
 gene <- t(as.matrix(field %>% 
                       filter(!MC.Eimeria == "NA") %>%
                       dplyr::select(c(Mouse_ID, all_of(Genes_field)))))
 
 #switch the matrix back to a data frame format
 gene <- as.data.frame(gene)
 
 # turn the first row into column names
 gene %>%
     row_to_names(row_number = 1) -> heatmap_data
 
 table(rowSums(is.na(heatmap_data)) == nrow(heatmap_data))
 
 # turn the columns to numeric other wise the heatmap function will not work
 heatmap_data[] <- lapply(heatmap_data, function(x) as.numeric(as.character(x)))
 
 # remove columns with only NAs 
 heatmap_data <- Filter(function(x)!all(is.na(x)), heatmap_data) 
 
 #remove rows with only Nas
 heatmap_data <-  heatmap_data[, colSums(is.na(heatmap_data)) != 
                                 nrow(heatmap_data)]
 
 ### Prepare the annotation data frame for the heatmap
 field$MC.Eimeria <- as.character(field$MC.Eimeria)

 annotation_df <- as_tibble(field) %>%
  filter(!MC.Eimeria == "NA") %>%
  dplyr::group_by(Mouse_ID) %>%
  dplyr::select(c("Mouse_ID", "MC.Eimeria")) %>%
  dplyr::filter(Mouse_ID %in% colnames(heatmap_data))
  
annotation_df <- unique(annotation_df)
 
annotation_df <- as.data.frame(unique(annotation_df)) %>%
  dplyr::select(-Mouse_ID)


### Prepare the annotation columns for the heatmap
rownames(annotation_df) <- annotation_df$Mouse_ID
# Match the row names to the heatmap data frame
rownames(annotation_df) <- colnames(heatmap_data)


heatmap_data <- as.data.frame(heatmap_data)



```

```{r}

pheatmap(heatmap_data, annotation_col = annotation_df, scale = "row")
```
# Heatmap with eimeria species
```{r}

# turn the data frame into a matrix and transpose it. We want to have each cell 
 # type as a row name 
 gene <- t(as.matrix(field %>% 
                       filter(!eimeriaSpecies == "NA") %>%
                       dplyr::select(c(Mouse_ID, all_of(Genes_field)))))
 
 #switch the matrix back to a data frame format
 gene <- as.data.frame(gene)
 
 # turn the first row into column names
 gene %>%
     row_to_names(row_number = 1) -> heatmap_data
 
 table(rowSums(is.na(heatmap_data)) == nrow(heatmap_data))
 
 # turn the columns to numeric other wise the heatmap function will not work
 heatmap_data[] <- lapply(heatmap_data, function(x) as.numeric(as.character(x)))
 
 # remove columns with only NAs 
 heatmap_data <- Filter(function(x)!all(is.na(x)), heatmap_data) 
 
 #remove rows with only Nas
 heatmap_data <-  heatmap_data[, colSums(is.na(heatmap_data)) != 
                                 nrow(heatmap_data)]
 
 annotation_df <- as_tibble(field) %>%
  filter(!MC.Eimeria == "NA") %>%
  dplyr::group_by(Mouse_ID) %>%
  dplyr::select(c("Mouse_ID", "eimeriaSpecies")) %>%
  dplyr::filter(Mouse_ID %in% colnames(heatmap_data))
  
annotation_df <- unique(annotation_df)
 
annotation_df <- as.data.frame(unique(annotation_df)) %>%
  dplyr::select(-Mouse_ID)


### Prepare the annotation columns for the heatmap
rownames(annotation_df) <- annotation_df$Mouse_ID
# Match the row names to the heatmap data frame
rownames(annotation_df) <- colnames(heatmap_data)


heatmap_data <- as.data.frame(heatmap_data)


```
```{r}

pheatmap(heatmap_data, annotation_col = annotation_df, scale = "row")
```

```{r histogram_genes, echo = FALSE, message=FALSE}
f <- field %>%
  pivot_longer(cols=  all_of(Genes_field), names_to = "Genes", 
               values_to = "expression")

f %>% 
  filter(!expression >= 40) %>%
  ggplot(aes(x = expression, color = eimeriaSpecies)) + 
  geom_histogram(echo = FALSE)
```

```{r gene_correlations, echo = FALSE}
gene_correlation <- field %>% 
  ungroup() %>%
  dplyr::select(all_of(Genes_field))

# draw correlation between the genes
gene_correlation <- as.matrix(cor(gene_correlation, use="pairwise.complete.obs"))
# load the function to calculate the p value for correlations
source("r_scripts/functions/p_value_for_correlations.R")
# matrix of the p-value of the correlatio
p.mat <- cor.mtest(gene_correlation)
```


## Corrplot of correlations 

Here is a corrplot of the correlations between the genes.
I am using the non-normalized genes

```{r corrplot_correlations_genes, echo = FALSE}
corrplot(gene_correlation, 
         method = "circle",  #method of the plot, "color" would show colour gradient
         tl.col = "black", tl.srt=45, #colour of labels and rotation
         col = brewer.pal(n = 8, name ="RdYlBu"), #colour of matrix
         order="hclust", #hclust reordering
         p.mat = p.mat, sig.level = 0.01, insig = "blank",
         addCoef.col = 'black',
         number.cex=0.5) #Add significance level to the correlogram
#remove the values that are insignificant
```
```{r}
res.pca <- PCA(field[,all_of(Genes_field)], scale.unit = TRUE, graph = FALSE)

```

```{r, echo = FALSE}
# extract pc scores for first two component and add to dat dataframe

  
field$pc1 <- res.pca$ind$coord[, 1] # indexing the first column
field$pc2 <- res.pca$ind$coord[, 2]  # indexing the second column


#We also need to extract the data for the variable contributions to each of the pc axes.
pca.vars <- res.pca$var$coord %>% data.frame
pca.vars$vars <- rownames(pca.vars)
pca.vars.m <- melt(pca.vars, id.vars = "vars")
source("r_scripts/functions/circle_fun.R")
circ <- circleFun(c(0,0),2,npoints = 500)
```


Biplot of the  gene pca


```{r biplot_pca_genes}
#Now we can make our initial plot of the PCA.
field %>% 
  pivot_longer(cols = all_of(Genes_field), names_to = "Gene", values_to = "gene_expression")  %>%
  ggplot(aes(x = pc1, y = pc2, color = eimeriaSpecies, shape = eimeriaSpecies)) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_vline(xintercept = 0, lty = 2) +
  geom_point(alpha = 0.8) +
  stat_ellipse(geom="polygon", aes(fill = eimeriaSpecies), alpha = 0.2, show.legend = FALSE,
               level = 0.95) +
  theme_minimal() +
  theme(panel.grid = element_blank(), panel.border = element_rect(fill= "transparent")) 
  
```


```{r correlations_genes_dimensions, echo = FALSE}
#It’s possible to use the function corrplot() [corrplot package] to highlight the most contributing variables for each dimension:
var.contrib <- res.pca$var$contrib
corrplot(var.contrib, is.corr=FALSE) 
```

The function fviz_contrib() [factoextra package] can be used to draw a bar plot of variable contributions. If your data contains many variables, you can decide to show only the top contributing variables. The R code below shows the top 10 variables contributing to the principal components:


```{r contr_var_pc_genes, echo = FALSE}
# Contributions of variables to PC1
fviz_contrib(res.pca, choice = "var", axes = 1, top = 18)
```


```{r}
# Contributions of variables to PC2
fviz_contrib(res.pca, choice = "var", axes = 2, top = 18)
```


```{r contr_var_pc1_2_genes, echo = FALSE}
fviz_contrib(res.pca, choice = "var", axes = 1:2, top = 18)
```
The red dashed line on the graph above indicates the expected average contribution. If the contribution of the variables were uniform, the expected value would be 1/length(variables) = 1/10 = 10%. For a given component, a variable with a contribution larger than this cutoff could be considered as important in contributing to the component.

Note that, the total contribution of a given variable, on explaining the variations retained by two principal components, say PC1 and PC2, is calculated as contrib = [(C1 * Eig1) + (C2 * Eig2)]/(Eig1 + Eig2), where

C1 and C2 are the contributions of the variable on PC1 and PC2, respectively
Eig1 and Eig2 are the eigenvalues of PC1 and PC2, respectively. Recall that eigenvalues measure the amount of variation retained by each PC.
In this case, the expected average contribution (cutoff) is calculated as follow: As mentioned above, if the contributions of the 10 variables were uniform, the expected average contribution on a given PC would be 1/10 = 10%. The expected average contribution of a variable for PC1 and PC2 is : [(10* Eig1) + (10 * Eig2)]/(Eig1 + Eig2)



```{r pca_contribution_genes, echo = FALSE}
#The most important (or, contributing) variables can be highlighted on the correlation plot as follow:
fviz_pca_var(res.pca, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")
             )
```

To visualize the contribution of individuals to the first two principal components:

```{r contr_individuals_genes, echo = FALSE}
# Total contribution on PC1 and PC2
fviz_contrib(res.pca, choice = "ind", axes = 1:2)
```

PCA + Biplot combination

```{r pca_biplot_genes, echo = FALSE}
fviz_pca_biplot(res.pca, 
                col.ind = field$eimeriaSpecies, palette = "jco", 
                addEllipses = TRUE, label = "var",
                col.var = "black", repel = TRUE,
                legend.title = "Parasite in the challenge infection") 
```
 In the following example, we want to color both individuals and variables by groups. The trick is to use pointshape = 21 for individual points. This particular point shape can be filled by a color using the argument fill.ind. The border line color of individual points is set to “black” using col.ind. To color variable by groups, the argument col.var will be used.



