---
title: "Gene_expression_analysis"
author: "Fay"
date: '2022-05-18'
output:
  pdf_document: 
    fig_width: 12
    fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```
GAPDH	HKG				
B-actin	HKG				
Ppia	HKG				
Ppip	HKG				
CDC42	HKG	susceptible to DNA contamination			
					
Housekeeping genes selected: GAPDH and PPIB					
					
Relm-b	mucosal defense factor (goblet cells)				
Muc2	the major secretory mucin within the gastrointestinal tract				
TFF3	mucosal defense factor (goblet cells)				
Muc5ac	similar to MUC2, produced by surface goblet cells				
NKp46	NK marker				
F4/80	macropahge marker (distinguish by immune response trend)				
Mpo	myeloperoxidase in Neutrophils				
MyD88	TLR protein, NF-kB IRAK protein, inflammation marker by TLR			MyD88-Dependent Pathway	
caspase-1	inflammasome marker (IL-1b and IL-18 production)				
IL-1Ra	natuaral IL-1b antagonist for infection control (if not increase in Tregs is seen)				
CXCL9,	immune cell migration marker + Th1 activator (confirm FACS)				
CXCR3,	CXCL9 and CXCL11 receptor				
IL-6	TNF inhibitor, 				
IL-12ra	T-cell marker Th1				
IFN-y	compare with IFN-y producing cells and IFN-y ELISAs, should correlate wit PRF1, NKp46 and F4/80. One of these cell types just have to be doing the job!				
IRG6A	autonomous cell defense (opsonization)				
TNF-a	upregulated in eimeria but not well explained. Could be present and driving infection where IFN-y isn't				
IL-17	in case IFN-y isn't coming up but pathogenicity is				
TRIF	Type I IFN production			TRIF Dependent Pathway	
Socs1	JAK/STAT signaling pathway, proinflammatory regulating + T-cell differentiation, could explain severity				
IDO1	DC, monocyte and MC protein regulating T-cell activity				
Prf1	perforin, should be dominant in primary infections, but must be correlated between T-cell and NK cell expresions				
CD56	CD56bright = more cytokine producing NKs, CD56dim = more direct cytotoxic killing				
IL-4					
IL-13					
IL-10					


## 1.  Gene expression in the laboratory infections - Heatmap

```{r load_libraries, echo=FALSE, include = FALSE}
library(pheatmap)
library(tidyverse)
library(matrixStats)
library(tidyr)
library(janitor)
library(tibble)
library(corrplot)
library(RColorBrewer)
library(ggplot2)
library(dplyr)
library(missMDA) 
library(FactoMineR)
library(PerformanceAnalytics)
library(factoextra)
library(MASS)
library(reshape2)
library(cowplot)
library(lmtest)
library(optimx)
if(any(grepl("package:plyr", search()))) detach("package:plyr") else message("plyr not loaded")

```


#document gene normalization
https://www.youtube.com/watch?v=tgp4bbnj-ng
paper: Licak Analysis of relative gene expression data using real time quantitative
Add the normalized Gene expression to the rest of the data 


```{r Data_preparation_heatmap, echo=FALSE, iclude = FALSE, message=FALSE}
### Import the data
Challenge <- read.csv("https://raw.githubusercontent.com/derele/Eimeria_Lab/master/data_products/Challenge_infections.csv")

# Vector for selecting genes
Genes <- c("IFNy", "CXCR3_bio", "IL.6", "IL.10", "IL.13","IL1RN", 
           "CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", "MUC2", "MUC5AC", "MYD88", 
           "NCR1", "PRF1", "RETNLB", "SOCS1", "TICAM1", "TNF")

Genes_refe <- c("IFNy", "CXCR3_bio", "IL.6", "IL.10", "IL.13", "IL1RN", 
           "CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", "MUC2", "MUC5AC", "MYD88", 
           "NCR1", "PRF1", "RETNLB", "SOCS1", "TICAM1", "TNF", "PPIB")



Challenge <- Challenge %>%
    dplyr::mutate(Parasite_primary = case_when(
        primary_infection == "E64" ~ "E_ferrisi",
        primary_infection == "E88" ~ "E_falciformis",
        primary_infection == "Eflab" ~ "E_falciformis",
        primary_infection == "E139" ~ "E_ferrisi",
        primary_infection == "UNI" ~ "uninfected",
        TRUE ~ ""))

Challenge <- Challenge %>%
    dplyr::mutate(Parasite_challenge = case_when(    
        challenge_infection == "E64" ~ "E_ferrisi",
        challenge_infection == "E88" ~ "E_falciformis",
        challenge_infection == "Eflab" ~ "E_falciformis",
        challenge_infection == "E139" ~ "E_ferrisi",
        challenge_infection == "UNI" ~ "uninfected",
        TRUE ~ ""))

Challenge <- Challenge %>%
  dplyr::mutate(infection_history = case_when(
    Parasite_primary == "uninfected" & Parasite_challenge == "uninfected" ~ "uninfected",
    Parasite_primary == "uninfected" & Parasite_challenge == "E_ferrisi" ~ "uninfected_ferrisi",
    Parasite_primary == "uninfected" & Parasite_challenge == "E_falciformis" ~ "uninfected_falciformis",
    Parasite_primary == "E_falciformis" & Parasite_challenge == "E_falciformis" ~ "falciformis_falciformis",
    Parasite_primary == "E_falciformis" & Parasite_challenge == "E_ferrisi" ~ "falciformis_ferrisi",
    Parasite_primary == "E_falciformis" & Parasite_challenge == "uninfected" ~ "falciformis_uninfected",
    Parasite_primary == "E_ferrisi" & Parasite_challenge == "E_falciformis" ~ "ferrisi_falciformis",
    Parasite_primary == "E_ferrisi" & Parasite_challenge == "E_ferrisi" ~ "ferrisi_ferrisi",
    Parasite_primary == "E_ferrisi" & Parasite_challenge == "uninfected" ~ "ferrisi_uninfected",
        TRUE ~ ""))

### Add the variable end weight (relative weight at day of sacrifice)
# start by adding the variable dpi_max which inficates the last day of each mouse
Challenge <- Challenge %>% 
  dplyr::filter(!weight == "NA") %>%
  dplyr::group_by(EH_ID, infection) %>%
  dplyr::mutate(dpi_max = max(dpi))

#somehow case when dplyr ways didn't work for me and this is the only solution 
#that is functional

#let's filter for the challenge mice
chal <- Challenge %>% filter(infection == "challenge")

#now only select the rows where the dpi is equal to the dpi max for each mouse
chal <- chal[chal$dpi == chal$dpi_max, ] 

#now we can easily add the variable end weight to each mouse (which in now equal
#to the weight on the dpi = dpi_max)
chal <- chal %>% dplyr::mutate(end_rel_weight = (weight/weight_dpi0) * 100)


#let'repeat for the prim 


#let's filter for the challenge mice
prim <- Challenge %>% filter(infection == "primary")

#now only select the rows where the dpi is equal to the dpi max for each mouse
prim <- prim[prim$dpi == prim$dpi_max, ] 

#now we can easily add the variable end weight to each mouse (which in now equal
#to the weight on the dpi = dpi_max)
prim <- prim %>% 
  dplyr::mutate(end_rel_weight = (weight/weight_dpi0) * 100)

c <- rbind(chal, prim)

#now jon it to the challenge infections
c %>% 
  dplyr::select(EH_ID, end_rel_weight) %>%
  right_join(Challenge) -> Challenge

Challenge <- unique(Challenge)

gene <- as_tibble(Challenge) %>%
  dplyr::filter(infection == "challenge", dpi == dpi_max) %>%
  dplyr::group_by(EH_ID) %>%
  dplyr::select(c(EH_ID, Eim_MC, all_of(Genes_refe)))

gene <- unique(gene)

gene1 <- gene

```

https://www.youtube.com/watch?v=tgp4bbnj-ng

If you don't need a graph with absolute values of expressions and just relative comparison then you can choose one of the samples as a calibrator (control sample). This control sample will have the value of 1 and the rest of the samples will be compared relative to this one. I would choose as control the sample with the lowest expression.

https://www.researchgate.net/post/How-can-I-analyze-qPCR-data-without-control-sample

```{r normalize_gene_expression}
#create calibrator samples for the analysis
# create new columns with the minimum of each gene
# which I will use as the calibrator sample
#create a function which gives you the PPIB value for the minimum value of a gene
Dct_Calibrator <- function(x) {
 h <-  gene[which(gene[[x]] == min(gene[[x]], na.rm = TRUE)), ][,c(x, "PPIB")]
 h[[x]] - h[["PPIB"]]
}


#Apply the function to all genes
Calibrator_values <- as.data.frame(unlist(lapply(Genes, Dct_Calibrator)))

Calibrator_values <- cbind(Calibrator_values, as.data.frame(Genes))


Calibrator_values <- Calibrator_values %>%
  rename(Dct.calibrator = "unlist(lapply(Genes, Dct_Calibrator))")

Calibrator_values <- Calibrator_values %>% pivot_wider(names_from = Genes, values_from = Dct.calibrator)



gene <- gene %>%
  dplyr::mutate(IFNy_N = 2^- ((IFNy - PPIB) - Calibrator_values[["IFNy"]]),
                CXCR3_bio_N  = 2^- ((CXCR3_bio - PPIB)- Calibrator_values[["CXCR3_bio"]]),
                IL.6_N = 2^-((IL.6 - PPIB)- Calibrator_values[["IL.6"]]),
                IL.10_N = 2^-((IL.10 - PPIB)- Calibrator_values[["IL.10"]]),
                IL.13_N = 2^-((IL.13 - PPIB)- Calibrator_values[["IL.13"]]),
                IL1RN_N = 2^-((IL1RN - PPIB)- Calibrator_values[["IL1RN"]]),
                CASP1_N = 2^-((CASP1 - PPIB)- Calibrator_values[["CASP1"]]),
                CXCL9_N  = 2^-((CXCL9 - PPIB)- Calibrator_values[["CXCL9"]]),
                IDO1_N = 2^-((IDO1 - PPIB)- Calibrator_values[["IDO1"]]),
                IRGM1_N = 2^-(( IRGM1 - PPIB)- Calibrator_values[["IRGM1"]]),
                MPO_N  =  2^-((MPO - PPIB) - Calibrator_values[["MPO"]]),
                MUC2_N = 2^-((MUC2 - PPIB)- Calibrator_values[["MUC2"]]),
                MUC5AC_N = 2^-((MUC5AC - PPIB)- Calibrator_values[["MUC5AC"]]),
                MYD88_N  = 2^-((MYD88 - PPIB)- Calibrator_values[["MYD88"]]),
                NCR1_N  = 2^-((NCR1 - PPIB)- Calibrator_values[["NCR1"]]),
                PRF1_N = 2^-((PRF1 - PPIB)- Calibrator_values[["PRF1"]]),
                RETNLB_N =  2^-((RETNLB - PPIB)- Calibrator_values[["RETNLB"]]),
                SOCS1_N = 2^-((SOCS1 - PPIB)- Calibrator_values[["SOCS1"]]),
                TICAM1_N  = 2^-((TICAM1 - PPIB)- Calibrator_values[["TICAM1"]]),
                TNF_N = 2^-((TNF - PPIB) - Calibrator_values[["TNF"]]))
           

#create a new vector for selecting genes
genes_N <- c("IFNy_N", "CXCR3_bio_N", "IL.6_N", "IL.10_N", "IL.13_N", "IL1RN_N", 
           "CASP1_N", "CXCL9_N", "IDO1_N", "IRGM1_N", "MPO_N", "MUC2_N", "MUC5AC_N", "MYD88_N", 
           "NCR1_N", "PRF1_N", "RETNLB_N", "SOCS1_N", "TICAM1_N", "TNF_N")


```

```{r}

# turn the data frame into a matrix and transpose it. We want to have each cell 
 # type as a row name 
 gene <- t(as.matrix(gene1 %>% dplyr::select(c(EH_ID, all_of(Genes)))))
 
 #switch the matrix back to a data frame format
 gene <- as.data.frame(gene)
 
 # turn the first row into column names
 gene %>%
     row_to_names(row_number = 1) -> heatmap_data
 
 table(rowSums(is.na(heatmap_data)) == nrow(heatmap_data))

 # turn the columns to numeric other wise the heatmap function will not work
 heatmap_data[] <- lapply(heatmap_data, function(x) as.numeric(as.character(x)))

 # remove columns with only NAs 
 heatmap_data <- Filter(function(x)!all(is.na(x)), heatmap_data) 
 
 #remove rows with only Nas
 heatmap_data <-  heatmap_data[, colSums(is.na(heatmap_data)) != nrow(heatmap_data)]


 ### Prepare the annotation data frame for the heatmap

 
annotation_df <- as_tibble(Challenge) %>%
  dplyr::filter(infection == "challenge", dpi == dpi_max) %>%
  dplyr::group_by(EH_ID) %>%
  dplyr::select(c("EH_ID", "Parasite_challenge", "infection_history","mouse_strain", 
                  "max_WL", "delta", "hybrid_status")) %>%
  dplyr::filter(EH_ID %in% colnames(heatmap_data))
  
annotation_df <- unique(annotation_df)
 

annotation_df <- as.data.frame(unique(annotation_df)) %>%
  dplyr::select(-EH_ID)

### Prepare the annotation columns for the heatmap
rownames(annotation_df) <- annotation_df$EH_ID


# Match the row names to the heatmap data frame
rownames(annotation_df) <- colnames(heatmap_data)

```

Heatmap on gene expression data: 

```{r pheatmap_genes, echo = FALSE}

pheatmap(heatmap_data, annotation_col = annotation_df, scale = "row")

```

## 2. Correlations between the genes 


```{r gene_correlations, echo = FALSE}
gene_correlation <- gene1 %>% 
  ungroup() %>%
  dplyr::select(all_of(Genes))

# draw correlation between the genes
gene_correlation <- as.matrix(cor(gene_correlation, use="pairwise.complete.obs"))

# load the function to calculate the p value for correlations
source("r_scripts/functions/p_value_for_correlations.R")

# matrix of the p-value of the correlatio
p.mat <- cor.mtest(gene_correlation)
```


## Corrplot of correlations 

Here is a corrplot of the correlations between the genes.
I am using the non-normalized genes

```{r corrplot_correlations_genes, echo = FALSE}

corrplot(gene_correlation, 
         method = "circle",  #method of the plot, "color" would show colour gradient
         tl.col = "black", tl.srt=45, #colour of labels and rotation
         col = brewer.pal(n = 8, name ="RdYlBu"), #colour of matrix
         order="hclust", #hclust reordering
         p.mat = p.mat, sig.level = 0.01, insig = "blank",
         addCoef.col = 'black',
         number.cex=0.5) #Add significance level to the correlogram
#remove the values that are insignificant
```

```{r gene_expression_intensity, echo = FALSE, message = FALSE}

g <-  as_tibble(Challenge) %>%
  dplyr::filter(infection == "challenge", dpi == dpi_max) %>%
  dplyr::group_by(EH_ID)%>%
  dplyr::filter(EH_ID %in% colnames(heatmap_data)) %>%
    pivot_longer(cols = all_of(Genes), names_to = "Gene", values_to = "gene_expression")

g <- unique(g)

g1 <- as_tibble(Challenge) %>%
  dplyr::filter(infection == "challenge", dpi == dpi_max) %>%
  dplyr::group_by(EH_ID)%>%
  dplyr::filter(EH_ID %in% colnames(heatmap_data)) %>%
  dplyr::select(all_of(Genes)) %>%
  ungroup() %>%
  dplyr::select(-EH_ID)

g1 <- unique(g1) 

g %>%
    filter(!Parasite_challenge == "uninfected", Eim_MC == "TRUE") %>%
    ggplot(aes(x = delta, y = gene_expression, color = Parasite_challenge)) +
    geom_jitter() +
    facet_wrap(~ Gene, scales = "free") +
    theme_light() +
    labs(x = "Delta Ct, Infection intensity", y = "Gene expression",
         title = "Gene expression in response to infection intensity") +
    theme_bw()

```




```{r gene_expression_box, echo = FALSE, message = FALSE}
g %>%
  ggplot(aes(x = Parasite_challenge, y = gene_expression, color = Parasite_challenge)) + 
  geom_boxplot() +
  geom_jitter() +
  facet_wrap(~ Gene) +
  theme_bw() +
    labs(x = "Infection groups, E64 = E. ferrisi, E88 = E.falciformis, 
         UNI = Uninfected", y = "Gene expression",
         title = "Gene expression in response to infection group") 
```

```{r histogram_genes, echo = FALSE, message=FALSE}
g  %>%
  ggplot(aes(x = gene_expression, color = challenge_infection)) + 
  geom_histogram(echo = FALSE)
```




It is possible to compute a pca with missing data using the package missMDA. 
The missMDA package is dedicated to missing values in exploratory  multivariate data analysis: single imputation/multiple imputation, etc. 

Following the tutorial of the package author: Francois Husson: https://www.youtube.com/watch?v=OOM8_FH6_8o

## 3. PCA
#### Handling missing data in a pca:
Bad methods: removing individuals with missing data or replacing missing data with the mean (default setting in many packages).


```{r, echo = FALSE, echo = FALSE, warning=FALSE}
#Make a scatter plot matrix showing the correlation coefficients between variables and the significance levels : the package PerformanceAnalytics is required.

chart.Correlation(g1, histogram=TRUE, pch=19)

```

We will now continue by using an iterative pca to impute missing data 
 A. Initialization: impute using the mean
 B. Step lampda:
  # a. do pca on imputed data table S dimensions retained
  # b. missing data imputed using pca
  # c. means (and standard deviations) updated
 C. Iterate the estimation and imputation steps (until convergence)
(convergence: the act of converging and especially moving toward union or uniformity)

Overfitting is a common problem due to believing too much in links between variables. 
--> regularized iterative PCA (This version is what is being implented in missMDA)
This is a way of taking less risk when imputing the missing data. 
The algorithm estimates the missing data values with values that have no influence
on the PCA results, i.e., no influence on the coordinates of the individals or variables. 


```{r pca_gene, echo = FALSE}

## 1. Estimate the number of dimensions necessary to impute the dataset 
# with the estim_ncpPCA
nb <- estim_ncpPCA(g1, scale = TRUE) 
# Each observed value is removed, then imputed using ncp components
# The difference between the ovserved value and the imputed value is calculated 
# ncp = 5

## 2. Imputation of the missing values using the function impute PCA
comp <- imputePCA(g1, ncp = 5, scale = TRUE) #impute the table
#head(comp$completeObs) # resulting imputed data table

# save the imputed file 
imputed_gene <- as.data.frame(comp$completeObs)
# this estimate should be better than simply using the mean, beacause we've taken into account
# the links between variables and similarities between individuals in its calculation

options(ggrepel.max.overlaps = Inf) # solution to the error: 
# ggrepel: 3 unlabeled data points (too many overlaps). Consider increasing max.overlaps 
# for the next function

# we can now run a normal pca on the complete data set
res.pca <- PCA(comp$completeObs)
#head(summary(res.pca, nbelements = Inf)) #including all the elements
```



```{r dimensions, include = FALSE, echo = FALSE, warnings = FALSE}
#Description of the dimensions
# We get a correlation between each variable and the first dimension
dimdesc(res.pca)
```

Caution: When imputing data, the percentages of inertia associated with the first dimensions will be overestimated.

Another problem: the imputed data are, when the pca is performed considered like real observations.
But they are estimations!! 

Visualizing uncertainty due to issing data:

--> mulrimple imputation: generate several plausible values for each missing data point

We here visualize the variability, that is uncertainty on the plane defined by two pca axes. 


```{r error_visualization_pca_gene, echo = FALSE, out.width="50%"}
#Visualizing uncertainty due to missing data:

mi <- MIPCA(g1, scale = TRUE, ncp = 5)
plot(mi)

```
Individuals lying on the axis have no missing data, but individuals that far away have many missing data. 
big ellipse = big uncertainty
tight elipse (line) = low uncertainty

Variable representation: 
Poins tight together )look like one) - have no missing variables --> low uncertainty
Points spread -- > higher variability -- > higher uncertainty


High uncertainty--> we should interpret the result with care

The individuals with many missing data values make the axes move, 
and thus the positions of all individuals

Therefore in the last plots every individual is getting an eclipse as they are as well influenced by the missing data of the others. 


THe plot with the dimensions shows the projections of the pca dimensions of each imputed table on the pca plane obtained using the original imputed data table

As all of the arrows are close to either the first or second axes, 
this means that the axes are stable with respect to the set of imputed tables --> we don't have evidence of instability here.


```{r, echo = FALSE, include = FALSE}
str(res.pca)
```



```{r, echo = FALSE}

# extract pc scores for first two component and add to dat dataframe
g2 <- g %>% 
  pivot_wider(names_from = "Gene", values_from = "gene_expression") %>%
  filter(EH_ID %in% colnames(heatmap_data)) %>%
  filter(Position == "mLN")
  
g2$pc1 <- res.pca$ind$coord[, 1] # indexing the first column

g2$pc2 <- res.pca$ind$coord[, 2]  # indexing the second column

imputed_gene$pc1 <- res.pca$ind$coord[, 1]
imputed_gene$pc2 <- res.pca$ind$coord[, 2] 

### join the pc1 and pc2 to the imputed data
imputed_expr <- g2 %>% left_join(imputed_gene, by = c("pc1", "pc2"), 
                                suffix = c("_x", "_y"))

#remove all columns of the non-imputed data
imputed_expr = imputed_expr[,!grepl("_x$",names(imputed_expr))]

#remove the suffix y
colnames(imputed_expr) = gsub("_y", "", colnames(imputed_expr))

#We also need to extract the data for the variable contributions to each of the pc axes.
pca.vars <- res.pca$var$coord %>% data.frame
pca.vars$vars <- rownames(pca.vars)
pca.vars.m <- melt(pca.vars, id.vars = "vars")

source("r_scripts/functions/circle_fun.R")

circ <- circleFun(c(0,0),2,npoints = 500)

```


Biplot of the imputed gene pca


```{r biplot_pca_genes}

#Now we can make our initial plot of the PCA.
imputed_expr %>% 
  pivot_longer(cols = all_of(Genes), names_to = "Gene", values_to = "gene_expression")  %>%
  ggplot(aes(x = pc1, y = pc2, color = Parasite_challenge, shape = Parasite_challenge)) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_vline(xintercept = 0, lty = 2) +
  geom_point(alpha = 0.8) +
  stat_ellipse(geom="polygon", aes(fill = challenge_infection), alpha = 0.2, show.legend = FALSE,
               level = 0.95) +
  theme_minimal() +
  theme(panel.grid = element_blank(), panel.border = element_rect(fill= "transparent")) 
  
```


```{r correlations_genes_dimensions, echo = FALSE}
#It’s possible to use the function corrplot() [corrplot package] to highlight the most contributing variables for each dimension:
var.contrib <- res.pca$var$contrib
corrplot(var.contrib, is.corr=FALSE) 
```

The function fviz_contrib() [factoextra package] can be used to draw a bar plot of variable contributions. If your data contains many variables, you can decide to show only the top contributing variables. The R code below shows the top 10 variables contributing to the principal components:


```{r contr_var_pc_genes, echo = FALSE}
# Contributions of variables to PC1
fviz_contrib(res.pca, choice = "var", axes = 1, top = 18)

```


```{r}
# Contributions of variables to PC2
fviz_contrib(res.pca, choice = "var", axes = 2, top = 18)
```


```{r contr_var_pc1_2_genes, echo = FALSE}
fviz_contrib(res.pca, choice = "var", axes = 1:2, top = 18)
```
The red dashed line on the graph above indicates the expected average contribution. If the contribution of the variables were uniform, the expected value would be 1/length(variables) = 1/10 = 10%. For a given component, a variable with a contribution larger than this cutoff could be considered as important in contributing to the component.

Note that, the total contribution of a given variable, on explaining the variations retained by two principal components, say PC1 and PC2, is calculated as contrib = [(C1 * Eig1) + (C2 * Eig2)]/(Eig1 + Eig2), where

C1 and C2 are the contributions of the variable on PC1 and PC2, respectively
Eig1 and Eig2 are the eigenvalues of PC1 and PC2, respectively. Recall that eigenvalues measure the amount of variation retained by each PC.
In this case, the expected average contribution (cutoff) is calculated as follow: As mentioned above, if the contributions of the 10 variables were uniform, the expected average contribution on a given PC would be 1/10 = 10%. The expected average contribution of a variable for PC1 and PC2 is : [(10* Eig1) + (10 * Eig2)]/(Eig1 + Eig2)



```{r pca_contribution_genes, echo = FALSE}

#The most important (or, contributing) variables can be highlighted on the correlation plot as follow:
fviz_pca_var(res.pca, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")
             )
```



To visualize the contribution of individuals to the first two principal components:

```{r contr_individuals_genes, echo = FALSE}
# Total contribution on PC1 and PC2
fviz_contrib(res.pca, choice = "ind", axes = 1:2)
```

PCA + Biplot combination

```{r pca_biplot_genes, echo = FALSE}

fviz_pca_biplot(res.pca, 
                col.ind = imputed_expr$Parasite_challenge, palette = "jco", 
                addEllipses = TRUE, label = "var",
                col.var = "black", repel = TRUE,
                legend.title = "Parasite in the challenge infection") 
```
 In the following example, we want to color both individuals and variables by groups. The trick is to use pointshape = 21 for individual points. This particular point shape can be filled by a color using the argument fill.ind. The border line color of individual points is set to “black” using col.ind. To color variable by groups, the argument col.var will be used.



################## Linear models: 

```{r lm_weight_pc, echo = FALSE}
weight_lm <- lm(max_WL ~ pc1 + pc2 + Parasite_challenge, data = imputed_expr)
summary(weight_lm)
AIC(weight_lm)
```




```{r lm_pc_parasite_hybrid, echo = FALSE}

weight_lm2 <- lm(max_WL ~ pc1 + pc2 + Parasite_challenge + hybrid_status, data = imputed_expr)
summary(weight_lm2)
AIC(weight_lm2)
```


Try instead: 
LLR test (likelihood ration)  (LM4 package )? 

https://www.rdocumentation.org/packages/lmtest/versions/0.9-38/topics/lrtest

In this way you compare each model, with the different variables usesd to predict. 

Another way is to compare the AIC. (function : step)

```{r}


weight_lm3 <- lm(max_WL ~ pc1 + pc2 + hybrid_status, data = imputed_expr)
weight_no_pc1 <- lm(max_WL ~ pc2 + hybrid_status, data = imputed_expr)
weight_no_pc2 <- lm(max_WL ~ pc1  + hybrid_status, data = imputed_expr)
weight_no_hybrid <- lm(max_WL ~ pc1 + pc2, data = imputed_expr)
lrtest(weight_lm3, weight_no_pc1)
lrtest(weight_lm3, weight_no_pc2)
lrtest(weight_lm3, weight_no_hybrid)
```


```{r lm_pc_hybrid, echo = FALSE}
weight_lm3 <- lm(max_WL ~ pc1 + pc2 + hybrid_status, data = imputed_expr)
summary(weight_lm3)
AIC(weight_lm3)
```



```{r lm_inf_histoy, echo=FALSE}
weight_lm4 <- lm(max_WL ~ pc1 + pc2 + infection_history, data = imputed_expr)
summary(weight_lm4)
AIC(weight_lm4)

```


```{r lm_pc, echo = FALSE}
weight_lm_exp_only <- lm(max_WL ~ pc1 + pc2, data = imputed_expr)
summary(weight_lm_exp_only)
AIC(weight_lm, weight_lm_exp_only)
```

### repeating the heatmap on the now imputed data
```{r}

gene <- imputed_expr %>% dplyr::select(c(EH_ID, all_of(Genes)))
 
 # turn the data frame into a matrix and transpose it. We want to have each cell 
 # type as a row name 
 gene <- t(as.matrix(gene))
 
 #switch the matrix back to a data frame format
 gene <- as.data.frame(gene)
 
 # turn the first row into column names
 gene %>%
     row_to_names(row_number = 1) -> heatmap_data
 
 
 table(rowSums(is.na(heatmap_data)) == nrow(heatmap_data))

 
 # turn the columns to numeric other wise the heatmap function will not work
 heatmap_data[] <- lapply(heatmap_data, function(x) as.numeric(as.character(x)))

 # remove columns with only NAs 
 heatmap_data <- Filter(function(x)!all(is.na(x)), heatmap_data) 
 
 #remove rows with only Nas
 heatmap_data <-  heatmap_data[, colSums(is.na(heatmap_data)) != nrow(heatmap_data)]

rownames(annotation_df) <- colnames(heatmap_data)
```


Heatmap on gene expression data: 


```{r, echo = FALSE}
pheatmap(heatmap_data, annotation_col = annotation_df, scale = "row")
```



### repeating the heatmap on the significant genes imputed data
```{r}

gene <- imputed_expr %>% dplyr::select(c(EH_ID, c("IFNy", "IL.13", "PRF1", 
                                                  "TICAM1")))
 
 # turn the data frame into a matrix and transpose it. We want to have each cell 
 # type as a row name 
 gene <- t(as.matrix(gene))
 
 #switch the matrix back to a data frame format
 gene <- as.data.frame(gene)
 
 # turn the first row into column names
 gene %>%
     row_to_names(row_number = 1) -> heatmap_data
 
 
 table(rowSums(is.na(heatmap_data)) == nrow(heatmap_data))

 
 # turn the columns to numeric other wise the heatmap function will not work
 heatmap_data[] <- lapply(heatmap_data, function(x) as.numeric(as.character(x)))

 # remove columns with only NAs 
 heatmap_data <- Filter(function(x)!all(is.na(x)), heatmap_data) 
 
 #remove rows with only Nas
 heatmap_data <-  heatmap_data[, colSums(is.na(heatmap_data)) != nrow(heatmap_data)]

  
annotation_df <- as_tibble(Challenge) %>%
  dplyr::filter(infection == "challenge", dpi == dpi_max) %>%
  dplyr::group_by(EH_ID) %>%
  dplyr::select(c("EH_ID", "Parasite_challenge",
                  "hybrid_status")) %>%
  dplyr::filter(EH_ID %in% colnames(heatmap_data))
  
annotation_df <- unique(annotation_df)
 

annotation_df <- as.data.frame(unique(annotation_df)) %>%
  dplyr::select(-EH_ID)

### Prepare the annotation columns for the heatmap
rownames(annotation_df) <- annotation_df$EH_ID

rownames(annotation_df) <- colnames(heatmap_data)
```


Heatmap on gene expression data: 


```{r, echo = FALSE}
pheatmap(heatmap_data, annotation_col = annotation_df, scale = "row")
```

# Testing alice's Balard package

HIs ranged from 0 to 1, HI of 0 indicating a pure Mmd and HI of 1 a pure Mmm (Baird et al., 2012; Macholán et al., 2007).

```{r}
#create a new variable for a hybrid index
g2 <- g2 %>% 
  mutate(HI = case_when(
    hybrid_status == "F1 hybrid" ~ 0.5,
    hybrid_status == "F0 M. m. domesticus" ~ 0,
    hybrid_status == "F1 M. m. domesticus" ~ 0,
    hybrid_status == "F1 M. m. musculus" ~ 1,
    hybrid_status == "F0 M. m. musculus" ~ 1,
    hybrid_status == "other" ~ 100,
    TRUE ~ 100
  ))

#filter out the nmri mice
g2 <- g2 %>%
  filter(!HI == 100) %>% 
  dplyr::mutate(max_WL = 100 - max_WL) 

g2 <- g2 %>% 
  mutate(hybrid_status = case_when(
    HI == 0.5 ~ "F1_hybrid",
    HI == 0 ~"M.m.domesticus",
    HI == 1 ~ "M.m.musculus"
  ))



g2 <- g2 %>%
  mutate(Eim_MC = case_when(
    Eim_MC == "TRUE" ~ "infected",
    Eim_MC == "FALSE" ~ "uninfected"
  ))


g2$Eim_MC <- as.factor(g2$Eim_MC)


require(devtools)
devtools::install_github("alicebalard/parasiteLoad@v2.0", force = TRUE)


library(parasiteLoad)

str(g2$max_WL)

parasiteLoad::getParamBounds("weibull", data = g2, response = "max_WL")

speparam <- c(L1start = 7.158694975,
                     L1LB = 0.000000001,
                     L1UB = 26.554054054,
                     L2start = 7.158694975,
                     L2LB = 0.000000001,
                     L2UB = 26.554054054,
                     alphaStart = 0, alphaLB = -5, alphaUB = 5,
                     myshapeStart = 1, myshapeLB = 0.000000001, myshapeUB = 5)
##All
#fitWL_MC <- parasiteLoad::analyse(data = g2,
                       # response = "max_WL",
                       # model = "weibull",
                       # group = "Eim_MC")

##fitWL_MC


#plot_WL_MC<- bananaPlot(mod = fitWL_MC$H3,
       #      data = g2,
     #        response = "max_WL",
    #         group = "Eim_MC") +
  #  scale_fill_manual(values = c("blue", "red")) +
#  scale_color_manual(values = c("blue", "red")) +
#  theme_bw()
#plot_WL_MC




HI_weight <- lm(max_WL ~ hybrid_status, data = g2)
summary(HI_weight)
HI_weight$coefficients

ggplot(data = g2, aes(x = hybrid_status, y = max_WL, color = hybrid_status)) +
  geom_boxplot()

AIC(HI_weight)



one.way <- aov(max_WL ~ hybrid_status, data = g2)

summary(one.way)
```
The Df column displays the degrees of freedom for the independent variable (the number of levels in the variable minus 1), and the degrees of freedom for the residuals (the total number of observations minus one and minus the number of levels in the independent variables).
The Sum Sq column displays the sum of squares (a.k.a. the total variation between the group means and the overall mean).
The Mean Sq column is the mean of the sum of squares, calculated by dividing the sum of squares by the degrees of freedom for each parameter.
The F-value column is the test statistic from the F test. This is the mean square of each independent variable divided by the mean square of the residuals. The larger the F value, the more likely it is that the variation caused by the independent variable is real and not due to chance.
The Pr(>F) column is the p-value of the F-statistic. This shows how likely it is that the F-value calculated from the test would have occurred if the null hypothesis of no difference among group means were true.
The p-value of the fertilizer variable is low (p < 0.001), so it appears that the type of fertilizer used has a real impact on the final crop yield.

```{r}
write.csv(imputed_expr, "output_data/gene_expression/data_products/imputed_gene_expression.csv", row.names = FALSE)

write.csv(g2, "output_data/gene_expression/data_products/clean_gene_expression.csv", row.names = FALSE)

```

