---
title: "2.Gene_expresion"
author: "Fay"
date: '2022-05-27'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Libraries:
```{r libraries}
library(tidyverse)
library(tidyr)
library(dplyr)
library(cowplot)
library(randomForest)
library(ggplot2)
library(caret)
```


#### Import the data:

```{r import}
g <- read.csv("output_data/gene_expression/data_products/clean_gene_expression.csv")


# vectors for selecting gene columns
Genes <- c("IFNy", "CXCR3_bio", "IL.6", "IL.10", "IL.13", "IL.10", "IL.13", "IL1RN", 
           "CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", "MUC2", "MUC5AC", "MYD88", 
           "NCR1", "PRF1", "RETNLB", "SOCS1", "TICAM1", "TNF")

```



#### Data cleaning

```{r cleaning}
g$Parasite_challenge <- as.factor(g$Parasite_challenge)
g$Eim_MC <- as.factor(g$Eim_MC)

```

#### Imputing missing data + cleaning
```{r imputing}
g <- g %>%
  dplyr::mutate(Infection = case_when(
    Parasite_challenge == "E_ferrisi" & Eim_MC == "TRUE" ~ "E_ferrisi",
    Parasite_challenge == "E_ferrisi" & Eim_MC == "FALSE" ~ "uninfected",
    Parasite_challenge == "E_falciformis" & Eim_MC == "TRUE" ~ "E_falciformis",
    Parasite_challenge == "E_falciformis" & Eim_MC == "FALSE" ~ "uninfected",
    Parasite_challenge == "uninfected" & Eim_MC == "TRUE" ~ "infected_eimeria",
    Parasite_challenge == "uninfected" & Eim_MC == "FALSE" ~ "uninfected",
    TRUE ~ ""
  ))

#create new variable end weight 
# since we will test it in the wild, we will also only have access to the end weight 
# of the wild mice 
g <- g %>% 
  dplyr::mutate(end_weight = (weight / weight_dpi0) * 100)


# how to impute delta? Replacing with 0 the ones with negative melting curve
g <- g %>%
  dplyr::mutate(Intensity = case_when(
    Eim_MC == "TRUE" ~ delta,
    Eim_MC == "FALSE" ~ 0))

g.1 <- g %>%
  dplyr::select(c(end_weight, all_of(Genes)))

# to get reproducible results we use a seed
set.seed(42)

# We want the maximum weight loss to be predicted by the data ina ll of the other columns

# iter = how many random forests are needed, in theory 6 are enough
g.imputed <- rfImpute(end_weight ~ ., data = g.1, iter = 6)

g.imputed <- g.imputed %>% dplyr::select(-end_weight)

g <- g %>% 
  dplyr::select(-all_of(Genes)) 

#full data set containing the imputed gene expression data
g.imputed <- cbind(g, g.imputed)




```

How many mice are in the infection planning?
```{r infection_plan}
g.imputed %>% 
  filter(infection == "challenge") %>%
  group_by(Parasite_challenge) %>%
  summarize(length(EH_ID))
  
```
How many mice are indeed infected?

```{r infected}
g.imputed %>% 
  filter(infection == "challenge") %>%
  group_by(Infection) %>%
  summarize(length(EH_ID))
  
```
I gues mice got mixed up here?

#### Splitting data into training and testing sets 
Splitting between training and testing:
- Assess model performance on unseen data
- Avoid over-fitting 


```{r}
#select the relevant columns:
g.imputed <- g.imputed %>%
  dplyr::select(c(end_weight, Infection, Intensity, all_of(Genes)))

# split data into training and test

set.seed(123) # this will help us reproduce this random assignment

# in this way we can pick the random numbers

training.samples <- g.imputed$end_weight%>%
  createDataPartition(p = .7, # this is the partiicition! In this case 0.7 = training data and 0.3 = testing
                      list = FALSE) # we don't want to get a list in return

train.data <- g.imputed[training.samples, ] #include all the randomly selected rows
test.data <- g.imputed[-training.samples, ] 


```

#### Building the model

```{r random_forest}
#train the model
model <- randomForest(end_weight ~., data = train.data, proximity = TRUE)

```

#### Making predictions
```{r predictións}
predictions <- predict(model, test.data)


result <- test.data
result['prediction'] <- predictions


#add the results to a data frame containing test data and the prediction
result <- cbind(g[row.names(result), ], result$prediction)



```

#### Visualizations
```{r prediction_scatter}
result   %>%
  ggplot() +
  geom_point(aes(x = `result$prediction`, y = end_weight)) 
  

```
```{r}
plot(y =result$end_weight, x = result$`result$prediction`, 
     # Draw plot using Base R
     xlab = "Predicted Values",
     ylab = "Observed Values", 
     abline(a = 0,                                        # Add straight line
       b = 1,
       col = "red",
       lwd = 2))
```


