---
title: "3.Gene_expressions_analysis -  Applying random forests on field data"
author: "Fay Webster"
date: '2022-07-18'
geometry: margin=2cm
output:
  pdf_document: default
  theme: journal
  toc: yes
  toc_float: yes
  html_document: null
  word_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Aim: 

- Applying the models established in the script: 
2.Gene_expression_analysis on the field data.
- How are hybrid mice different to the parental species?

#  Load necessary libraries:

```{r libraries, message = FALSE, warnings = FALSE}
#install.packages("optimx", version = "2021-10.12") # this package is required for 
#the parasite load package to work
library(tidyverse)
library(tidyr)
library(dplyr)
library(cowplot)
library(randomForest)
library(ggplot2)
library(VIM) # visualizing missing data
library(mice) # imputing missing data without predictors
library(ggpubr)
library(optimx)
library(rfUtilities) # Implements a permutation test cross-validation for 
# Random Forests models
library(mice) #imputations
library(fitdistrplus) #testing distributions
library(logspline)
library(caret)
```

# Field data

## Import field data


```{r}
Field <- read.csv("https://raw.githubusercontent.com/derele/Mouse_Eimeria_Field/master/data_products/SOTA_Data_Product.csv")

```


## Clean data 

```{r summary_stats_field}
Field %>% summarise(length(Mouse_ID))
Field <- Field %>%
    drop_na(HI)
```

We have 1921 mice in total.


#### Prepare columns for selecting

```{r genes}
EqPCR.cols      <- c("delta_ct_cewe_MminusE", "MC.Eimeria", "Ct.Eimeria") #,"Ct.Mus""delta_ct_ilwe_MminusE", )

EimGeno.cols    <- c("n18S_Seq", "COI_Seq", "ORF470_Seq", "eimeriaSpecies")

Gene.Exp.cols   <- c("IFNy",  "CXCR3", "IL.6", #"GBP2", "IL.12", "IRG6",
                     "IL.10", "IL.13", "IL1RN",
                     "CXCR3", "CASP1", "CXCL9", 
                     "IDO1", "IRGM1", "MPO", "MUC2", "MUC5AC", "MYD88", 
                     "NCR1", "PRF1", "RETNLB", "SOCS1", "TICAM1", "TNF")

House.Keeping.cols <- c("GAPDH", "PPIB", "B.actin", "B-actin")
```


### Actual Cleaning

```{r}
#which are the numbers of the columns of Field
names <- data.frame(colnames(Field))

#how many nas in each column
sapply(Field, function(x) sum(is.na(x)))

f <- Field %>%
  dplyr::select(all_of(Gene.Exp.cols))

#remove rows with only nas
f <- f[rowSums(is.na(f)) != ncol(f), ]

# subset only the rows without any nas everywhere from the first data frame 
Field <- Field[row.names(f), ]

# select columns to be included in imputation
# Including data on body size, adult / pregnant, and on parasites other
# than eimeria


f.1 <- Field %>% dplyr::select(c("Body_Weight", "Body_Length", "Tail_Length",
                                 "Aspiculuris_sp", "Syphacia_sp", 
                                 "Trichuris_muris", "Taenia_sp", 
                                 "Heterakis_sp", "Mastophorus_muris", 
                                 "Hymenolepis_sp", "Catenotaenia_pusilla",
                                 all_of(Gene.Exp.cols)))
                                 
#how many nas in each column
sapply(f.1, function(x) sum(is.na(x)))


f.1  <- f.1 %>% mutate_if(is.character, as.factor)
f.1  <- f.1 %>% mutate_if(is.integer, as.numeric)

```


`

## Imputing missing data 

For the lab data I have used the function rfimpute from the package random 
forest. I can't use the same function for our lab data as the function requires
the data set to contain predictor variable and response variables. 

Therefore I will be using the package MICE (multivariate Imputation by chained 
Equations) which only requires a data frame of missing observations.

Description: *Multiple imputation using Fully Conditional Specification (FCS)*

implemented by the MICE algorithm as described in Van Buuren and
Groothuis-Oudshoorn (2011) <doi:10.18637/jss.v045.i03>. Each variable has
its own imputation model. Built-in imputation models are provided for
continuous data (predictive mean matching, normal), binary data (logistic
regression), unordered categorical data (polytomous logistic regression)
and ordered categorical data (proportional odds). MICE can also impute
continuous two-level data (normal model, pan, second-level variables).
Passive imputation can be used to maintain consistency between variables.
Various diagnostic plots are available to inspect the quality of the
imputations.

https://www.jstatsoft.org/article/view/v045i03

tutorial: https://www.youtube.com/watch?v=WPiYOS3qK70

https://datascienceplus.com/imputing-missing-data-with-r-mice-package/


https://datascienceplus.com/handling-missing-data-with-mice-package-a-simple-approach/



### Missing data can be classified into three categories:
#### 1. Missing completely at random (MCAR)
  
We can't probably predict that value from any other value in the data. 
MCAR implies the reason for the missingness of a field is completely random, 
and that we probably can't predict that value from any other value in the data. 
  
#### 2. Missing at Random (MAR)

Missingess can be explained by other values in other columns, but not from
that column.
  
#### 3. Missing NOT at random (MNAR)


The basic MICE assumption is that the data is missing at random, and that we can 
make a guess about its true value by looking at other data samples.



## Step1 : cleaning and checking the missing data points in our field data. 

```{r impute_missing}

# check the data for missing values
sapply(f.1, function(x) sum(is.na(x)))

f.1 %>%
  aggr(col = c('navyblue', 'red'), numbers = TRUE, sortVars = TRUE, 
       labels=names(f), cex.axis=.7, gap=3, 
       ylab=c("Histogram of missing data","Pattern"))
       
       
marginplot(f.1[c(1,2)])

```



## Now let's coninue by ussing the package MICE to impute the data

```{r imputing_mice}

# The frequency distribution of the missing cases per variable can be obtained 
# as:
init <- mice(f.1, maxit = 0)

# table of amount of variables with the amount of missing values 
table(init$nmis)

# which method is used for imputation? In this case the package mice 
# uses the default method for continuous variable, 
# which is pmm, or predictive mean matching

meth <- init$method


# now impute the immune gene expression for the field and save it as the oject: 
# igf
# m=5 refers to the number of imputed datasets. Five is the default value.
igf <- mice(f.1, method = meth, m = 5, seed = 500)
summary(igf)

# to check each column with imputed data
## igf$imp$IFNy

#Now we can get back the completed dataset using the complete()
completeField <- complete(igf, 1)

```

Predictive mean matching with \(d = 5\) is the default in mice() for continuous 
data. The method is robust against misspecification of the imputation model, 
yet performs as well as theoretically superior methods. In the context of missing 
covariate data, Marshall, Altman, and Holder (2010) concluded that predictive 
mean matching “produced the least biased estimates and better model performance 
measures.” Another simulation study that addressed skewed data concluded that 
predictive mean matching “may be the preferred approach provided that less than 
50% of the cases have missing data and the missing data are not MNAR” 
(Marshall et al. 2010). Kleinke (2017) found that the method works well across a 
wide variety of scenarios, but warned the default cannot address severe 
skewness or small samples.


Let’s compare the distributions of original and imputed data using a some useful 
plots.First of all we can use a scatterplot and plot Ozone against all the other 
variables. Let's first plot the variables for which we have few missing values.



```{r distr_orig_pred}

xyplot(igf,IFNy ~ SOCS1 + IRGM1 + MUC2, pch=18,cex=1)


```


What we would like to see is that the shape of the magenta points (imputed) 
matches the shape of the blue ones (observed). The matching shape tells us that 
the imputed values are indeed “plausible values”.

Now let's plot the variables with many missing data points.


```{r distr_orig_pred2}

xyplot(igf,IFNy ~ IL.10 + PRF1 + CASP1, pch=18,cex=1)
```



```{r densityplotfield}
densityplot(igf)
```

The density of the imputed data for each imputed dataset is showed in magenta 
while the density of the observed data is showed in blue. Again, under our 
previous assumptions we expect the distributions to be similar.

Another useful visual take on the distributions can be obtained using the 
stripplot() function that shows the distributions of the variables as individual 
points


```{r strirrplot}
stripplot(igf, pch = 20, cex = 1.2)
```


# Predicting weight loss in our imputed field data

Start by making the predictions for the field data. 


```{r predicting_field }
#merge the imputations with the field data and save the output file
Field_imputed <- Field %>%
  dplyr::select(-colnames(completeField))

Field_imputed <- cbind(Field_imputed, completeField)

write.csv(Field_imputed, 
          "output_data/gene_expression/data_products/field_imputed_gene_expression.csv", 
          row.names = FALSE)


completeField <- completeField %>%
  dplyr::select(intersect(colnames(completeField), c("IFNy", "IL.6", "IL.10", 
                                                     "IL.13", "IL1RN", "CASP1",
                                                     "CXCL9", "IDO1", "IRGM1", 
                                                     "MPO", "MUC2", "MUC5AC", 
                                                     "MYD88", "NCR1", "PRF1", 
                                                     "RETNLB", "SOCS1", "TICAM1",
                                                     "TNF")))


# load predicting weight loss model
load("r_scripts/models/predict_weight_loss.RData")

set.seed(540)


#The predict() function in R is used to predict the values based on the input data.
predictions_field <- predict(weight_loss_predict, completeField)

# assign test.data to a new object, so that we can make changes
result_field <- completeField

#add the new variable of predictions to the result object
result_field <- cbind(result_field, predictions_field)

# add it to the field data 
Field <- cbind(Field, predictions_field)
```


# It is time to apply the package of Alice Balard et al. on our predictions!


Let's see if we indeed have differences across the hybrid index with our predicted 
weight loss. 

## Install the package

```{r, warning=FALSE, echo=FALSE, message=FALSE}
require(devtools)

devtools::install_github("alicebalard/parasiteLoad@v2.0", force = TRUE)

#force = TRUE)

library(parasiteLoad)
```

## Data diagnostics

### Visualizations 

#### What is the distribution of the predicted weight loss?


```{r}

Field %>% ggplot(aes(x = predictions_field)) +
  geom_histogram(binwidth = 1.5)

```


#### Rough graph of our predictions against the hybrid index and against the 
#### body length



```{r}
Field %>%
    ggplot(aes(x = HI , y = predictions_field , color = Sex)) +
    geom_smooth() +
    geom_point()


Field %>%
    ggplot(aes(x = Body_Length , y = predictions_field , color = Sex)) +
    geom_smooth() +
    geom_point()
```




### Fitting distributions??

Ratios / Percentages are not normally distributed. Weibull is a good distributions. 

Alice used weibull for the qpcr data. (paper) 

```{r}
Field <- Field %>%
dplyr::mutate(WL = predictions_field)

x <- Field$WL

descdist(data = x, discrete = FALSE)
descdist(data = x, discrete = FALSE, #data is continuous
         boot = 1000)


```

### Test for binomial distribution

```{r}
set.seed(10)
n = 25
size = 27
prob = .4
data = rbinom(x, size = size, prob = prob)
fit = fitdist(data = data, dist="binom", 
                   fix.arg=list(size = size), 
                   start=list(prob = 0.1))

summary(fit)


plot(fit)
```


#### Functions for testing distributions
```{r}
normal_ <- fitdist(x, "norm")
weibull_ <- fitdist(x, "weibull")
gamma_ <- fitdist(x, "gamma")


# Define function to be used to test, get the log lik and aic
tryDistrib <- function(x, distrib){
  # deals with fitdistr error:
  fit <- tryCatch(MASS::fitdistr(x, distrib), error=function(err) "fit failed")
  return(list(fit = fit,
              loglik = tryCatch(fit$loglik, error=function(err) "no loglik computed"), 
              AIC = tryCatch(fit$aic, error=function(err) "no aic computed")))
}



findGoodDist <- function(x, distribs, distribs2){
  l =lapply(distribs, function(i) tryDistrib(x, i))
  names(l) <- distribs
  print(l)
  listDistr <- lapply(distribs2, function(i){
    if (i %in% "t"){
      fitdistrplus::fitdist(x, i, start = list(df =2))
    } else {
      fitdistrplus::fitdist(x,i)
    }}
  ) 
  par(mfrow=c(2,2))
  denscomp(listDistr, legendtext=distribs2)
  cdfcomp(listDistr, legendtext=distribs2)
  qqcomp(listDistr, legendtext=distribs2)
  ppcomp(listDistr, legendtext=distribs2)
  par(mfrow=c(1,1))
}
```


```{r}
tryDistrib(x, "normal")
tryDistrib(x, "binomial")
tryDistrib(x, "student")
tryDistrib(x, "weibull")
tryDistrib(x, "weibullshifted")

```

```{r}
findGoodDist(x, "normal", "weibull")
```

```{r normal}
plot(normal_)
summary(normal_)
plot(gamma_)
summary(gamma_)
plot(weibull_)
summary(weibull_)
```


We have a weibull distribution!



### Is alpha significant for each hypothesis?



```{r}
Field$Sex <- as.factor(Field$Sex)



parasiteLoad::getParamBounds("weibull", data = Field, response = "WL")


speparam <- c(L1start = 10,
                     L1LB = 1e-9,
                     L1UB = 20,
                     L2start = 10,
                     L2LB = 1e-9,
                     L2UB = 20,
                     alphaStart = 0, alphaLB = -5, alphaUB = 5,
                     myshapeStart = 1, myshapeLB = 1e-9, myshapeUB = 5)

##All
fitWL_Sex <- parasiteLoad::analyse(data = Field,
                        response = "WL",
                        model = "weibull",
                        group = "Sex")

fitWL_Sex

plot_WL_Sex<- bananaPlot(mod = fitWL_Sex$H3,
             data = Field,
             response = "WL",
             group = "Sex") +
    scale_fill_manual(values = c("blue", "red")) +
  scale_color_manual(values = c("blue", "red")) +
  theme_bw()

plot_WL_Sex

```

<p style="color:blue">
H0: the expected load for the subspecies and between 2 groups is the same

H1: the mean load across 2 groups is the same, but can differ across subspecies

H2: the mean load across subspecies is the same, but can differ between the 
2 groups

H3: the mean load can differ both across subspecies and between 2 groups

## Summary stats for the field 

Can we test the hybrid index, WL and the infection ? 

```{r}
Field %>%
  dplyr::group_by(MC.Eimeria) %>%
  dplyr::summarise(length(Mouse_ID))
    
```


```{r}
Field %>%
    dplyr::group_by(eimeriaSpecies) %>%
    dplyr::summarize(length(Mouse_ID))
```

## Reproducing for melting curve


```{r}

Field_mc <- Field %>% 
    drop_na("MC.Eimeria")

parasiteLoad::getParamBounds("weibull", data = Field_mc, response = "WL")


speparam <- c(L1start = 10,
                     L1LB = 1e-9,
                     L1UB = 20,
                     L2start = 10,
                     L2LB = 1e-9,
                     L2UB = 20,
                     alphaStart = 0, alphaLB = -5, alphaUB = 5,
                     myshapeStart = 1, myshapeLB = 1e-9, myshapeUB = 5)


Field_mc <- Field_mc %>%
    dplyr::mutate(Eimeria = case_when(
        MC.Eimeria == "TRUE" ~ "positive",
        MC.Eimeria == "FALSE" ~ "negative",
        TRUE ~ ""
    ))

Field_mc$Eimeria <- as.factor(Field_mc$Eimeria)


##All
fitWL_Eimeria <- parasiteLoad::analyse(data = Field_mc,
                        response = "WL",
                        model = "weibull",
                        group = "Eimeria")
fitWL_Eimeria

plot_WL_Eimeria <- bananaPlot(mod = fitWL_Eimeria$H0,
             data = Field_mc,
             response = "WL",
             group = "Eimeria") +
  scale_fill_manual(values = c("blue", "red")) +
  scale_color_manual(values = c("blue", "red")) +
  theme_bw()

plot_WL_Eimeria

```



## Applying the classification model on the field data


### Predicting melting curve
```{r }
#
# load model
#load("r_scripts/models/predicting_mc.RData")

#The predict() function in R is used to predict the values based on the input 
# data.
#predictions_mc_field <- predict(model_mc, completeField)


# assign test.data to a new object, so that we can make changes
#result_field_mc <- completeField

#add the new variable of predictions to the result object
#result_field_mc <- cbind(result_field_mc, predictions_mc_field)

# add it to the field data 
#Field <- cbind(Field, result_field_mc)



#levels(Field$predictions_mc_field) <- c("TRUE", "FALSE")


#turn the variable of melting curve into a factor so that you can compare to 
# the predictions
#Field$predictions_mc_field <- 
 # as.factor(as.character(Field$predictions_mc_field))

#Field$MC.Eimeria <- 
 # as.factor(as.character(Field$MC.Eimeria))

```




### Confusion matrix

```{r include = FALSE }


# model
#Field$predictions_mc_field <- 
 # gsub("uninfected", FALSE, Field$eimeriaSpecies)




#conf_matrix_mc <- 
#  confusionMatrix(Field_mc$predictions_mc_field, 
 #                 reference = Field_mc$MC.Eimeria)

#print(conf_matrix_mc)

#conf_matrix_mc$table

#plt <- as.data.frame(conf_matrix_mc$table)

#plt$Prediction <- factor(plt$Prediction, levels=rev(levels(plt$Prediction)))
#ggplot(plt, aes(x = Prediction, y =  Reference, fill= Freq)) +
 #       geom_tile() + geom_text(aes(label=Freq)) +
  #      scale_fill_gradient(low="white", high="forestgreen") +
   #     labs(x = "Predictions",y = "Reference") 
```

## Test the other model for parasite species
## Making predictions


```{r }
load("r_scripts/models/predict_infecting_parasite.RData")

#The predict() function in R is used to predict the values based on 
# the input data.
predictions_parasite_field <- predict(model_Parasite, completeField)

# edit the infecting parasite names to correspond the parasite names in the 
# model
Field$eimeriaSpecies <- 
  gsub("Negative", "uninfected", Field$eimeriaSpecies)


Field$eimeriaSpecies <- as.factor(Field$eimeriaSpecies)

#add the new variable of predictions to the result object
Field_Eim <- cbind(Field, predictions_parasite_field)

# drop the rows with nas in the parasite 
Field_Eim <- Field_Eim %>% 
  drop_na(eimeriaSpecies)


```




## Visualizations



```{r }

conf_matrix_parasite <- 
  confusionMatrix(Field_Eim$predictions_parasite_field, 
                  reference = Field_Eim$eimeriaSpecies)

print(conf_matrix_parasite)

conf_matrix_parasite$table

plt <- as.data.frame(conf_matrix_parasite$table)
plt$Prediction <- factor(plt$Prediction, levels=rev(levels(plt$Prediction)))

ggplot(plt, aes(x = Prediction, y =  Reference, fill= Freq)) +
        geom_tile() + geom_text(aes(label=Freq)) +
        scale_fill_gradient(low="white", high="forestgreen") +
        labs(x = "Predictions",y = "Reference") 

```
# Re-testing infection intensities across the hybrid zone

```{r}
Field_delta <- Field %>% 
    drop_na("delta_ct_cewe_MminusE")

x <- Field_delta$delta_ct_cewe_MminusE

tryDistrib(x, "normal")
tryDistrib(x, "binomial")
tryDistrib(x, "student")
tryDistrib(x, "weibull")
tryDistrib(x, "weibullshifted")

```
We have a normal distribution.


```{r}

parasiteLoad::getParamBounds("normal", data = Field_delta, 
                             response = "delta_ct_cewe_MminusE")


speparam <- c(L1start = -9,
                     L1LB = -14,
                     L1UB = 4,
                     L2start = -9,
                     L2LB = -14,
                     L2UB = 4,
                     alphaStart = 0, alphaLB = -5, alphaUB = 5,
                     myshapeStart = 1, myshapeLB = 1e-9, myshapeUB = 10)


##All
fit_delta <- parasiteLoad::analyse(data = Field_delta,
                        response = "delta_ct_cewe_MminusE",
                        model = "normal",
                        group = "Sex")
fit_delta

plot_delta <- bananaPlot(mod = fit_delta$H0,
             data = Field_delta,
             response = "delta_ct_cewe_MminusE",
             islog10 = FALSE, group = "Sex") +
  scale_fill_manual(values = c("blue", "red")) +
  scale_color_manual(values = c("blue", "red")) +
  theme_bw() 



plot_delta

```

