---
title: "1. Merging_cleaning_heatmap_gene_expression"
author: "Fay"
date: "2022-09-15"
output:
  pdf_document:
    keep_md: yes
    fig_width: 12
    fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

GAPDH	HKG				
B-actin	HKG				
Ppia	HKG				
Ppip	HKG				
CDC42	HKG	susceptible to DNA contamination			
					
Housekeeping genes selected: GAPDH and PPIB					
					
Relm-b	mucosal defense factor (goblet cells)				
Muc2	the major secretory mucin within the gastrointestinal tract				
TFF3	mucosal defense factor (goblet cells)				
Muc5ac	similar to MUC2, produced by surface goblet cells				
NKp46	NK marker				
F4/80	macropahge marker (distinguish by immune response trend)				
Mpo	myeloperoxidase in Neutrophils				
MyD88	TLR protein, NF-kB IRAK protein, inflammation marker by TLR			MyD88-Dependent Pathway	
caspase-1	inflammasome marker (IL-1b and IL-18 production)				
IL-1Ra	natuaral IL-1b antagonist for infection control (if not increase in Tregs is seen)				
CXCL9,	immune cell migration marker + Th1 activator (confirm FACS)				
CXCR3,	CXCL9 and CXCL11 receptor				
IL-6	TNF inhibitor, 				
IL-12ra	T-cell marker Th1				
IFN-y	compare with IFN-y producing cells and IFN-y ELISAs, should correlate wit PRF1, NKp46 and F4/80. One of these cell types just have to be doing the job!				
IRG6A	autonomous cell defense (opsonization)				
TNF-a	upregulated in eimeria but not well explained. Could be present and driving infection where IFN-y isn't				
IL-17	in case IFN-y isn't coming up but pathogenicity is				
TRIF	Type I IFN production			TRIF Dependent Pathway	
Socs1	JAK/STAT signaling pathway, proinflammatory regulating + T-cell differentiation, could explain severity				
IDO1	DC, monocyte and MC protein regulating T-cell activity				
Prf1	perforin, should be dominant in primary infections, but must be correlated between T-cell and NK cell expresions				
CD56	CD56bright = more cytokine producing NKs, CD56dim = more direct cytotoxic killing				
IL-4					
IL-13					
IL-10		

# Load libraries

```{r load_libraries, echo=FALSE, include = FALSE}
library(pheatmap)
library(tidyverse)
library(tidyr)
library(dplyr)
library(janitor)
library(visdat)
```


# Import data 


Challenge = experimental challenge infection data

SOTA = State of the Art, of wild data

```{r}
Challenge <- read.csv("input_data/Challenge_infections.csv")
SOTA <- read.csv("input_data/SOTA_Data_Product.csv")
# Vectors for selecting genes
#Lab genes
# The measurements of IL.12 and IRG6 are done with an other assay and will 
#ignore for now
Gene_lab   <- c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10",
                "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                "TICAM1", "TNF") #"IL.12", "IRG6")
Genes_wild   <- c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10", 
                  "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                  "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                  "TICAM1", "TNF") #, "IL.12", "IRG6")
Facs_lab <- c("Position", "CD4", "Treg", "Div_Treg", "Treg17", "Th1", 
                    "Div_Th1", "Th17", "Div_Th17", "CD8", "Act_CD8", 
                    "Div_Act_CD8", "IFNy_CD4", "IFNy_CD8","Treg_prop", 
                    "IL17A_CD4")  
Facs_wild <- c( "Treg", "CD4", "Treg17", "Th1", "Th17", "CD8",
                     "Act_CD8", "IFNy_CD4", "IL17A_CD4", "IFNy_CD8")
```


# Cleaning

## Challenge

```{r, echo = FALSE}
Challenge <- Challenge %>%
    dplyr::mutate(Parasite_primary = case_when(
        primary_infection == "E64" ~ "E_ferrisi",
        primary_infection == "E88" ~ "E_falciformis",
        primary_infection == "Eflab" ~ "E_falciformis",
        primary_infection == "E139" ~ "E_ferrisi",
        primary_infection == "UNI" ~ "uninfected",
        TRUE ~ ""))
Challenge <- Challenge %>%
    dplyr::mutate(Parasite_challenge = case_when(    
        challenge_infection == "E64" ~ "E_ferrisi",
        challenge_infection == "E88" ~ "E_falciformis",
        challenge_infection == "Eflab" ~ "E_falciformis",
        challenge_infection == "E139" ~ "E_ferrisi",
        challenge_infection == "UNI" ~ "uninfected",
        TRUE ~ ""))
Challenge <- Challenge %>%
  dplyr::mutate(infection_history = case_when(
    Parasite_primary == "uninfected" & 
        Parasite_challenge == "uninfected" ~ "uninfected",
    Parasite_primary == "uninfected" & 
        Parasite_challenge == "E_ferrisi" ~ "uninfected_ferrisi",
    Parasite_primary == "uninfected" & 
        Parasite_challenge == "E_falciformis" ~ "uninfected_falciformis",
    Parasite_primary == "E_falciformis" & 
        Parasite_challenge == "E_falciformis" ~ "falciformis_falciformis",
    Parasite_primary == "E_falciformis" & 
        Parasite_challenge == "E_ferrisi" ~ "falciformis_ferrisi",
    Parasite_primary == "E_falciformis" & 
        Parasite_challenge == "uninfected" ~ "falciformis_uninfected",
    Parasite_primary == "E_ferrisi" & 
        Parasite_challenge == "E_falciformis" ~ "ferrisi_falciformis",
    Parasite_primary == "E_ferrisi" & 
        Parasite_challenge == "E_ferrisi" ~ "ferrisi_ferrisi",
    Parasite_primary == "E_ferrisi" &
        Parasite_challenge == "uninfected" ~ "ferrisi_uninfected",
        TRUE ~ ""))
### Add the variable end weight (relative weight at day of sacrifice)
# start by adding the variable dpi_max which inficates the last day of each mouse
Challenge <- Challenge %>% 
  dplyr::filter(!weight == "NA") %>%
  dplyr::group_by(EH_ID, infection) %>%
  dplyr::mutate(dpi_max = max(dpi), WL_max = (min(relative_weight) - 100))
#somehow case when dplyr ways didn't work for me and this is the only solution 
#that is functional
#let's filter for the challenge mice
chal <- Challenge %>% filter(infection == "challenge")
#now only select the rows where the dpi is equal to the dpi max for each mouse
chal <- chal[chal$dpi == chal$dpi_max, ] 
#now we can easily add the variable end weight to each mouse (which in now equal
#to the weight on the dpi = dpi_max)
chal <- chal %>% dplyr::mutate(end_rel_weight = (weight/weight_dpi0) * 100)
#let'repeat for the prim 
#let's filter for the challenge mice
prim <- Challenge %>% filter(infection == "primary")
#now only select the rows where the dpi is equal to the dpi max for each mouse
prim <- prim[prim$dpi == prim$dpi_max, ] 
#now we can easily add the variable end weight to each mouse (which in now equal
#to the weight on the dpi = dpi_max)
prim <- prim %>% 
  dplyr::mutate(end_rel_weight = (weight/weight_dpi0) * 100)
c <- rbind(chal, prim)
#now jon it to the challenge infections
c %>% 
  dplyr::select(EH_ID, end_rel_weight) %>%
  right_join(Challenge) -> Challenge
Challenge <- unique(Challenge)
#There are two measuremts for CXCR3
# We want to here keep the CXCR3_bio 
Challenge <- Challenge %>% 
  dplyr::select(-CXCR3)
#Now rename the CXCR3_bio to CXCR3
Challenge <- Challenge %>%
  dplyr::rename(CXCR3 = CXCR3_bio)
rm(c, chal, prim)
```


# Join wild and lab data 

```{r, message = FALSE, echo = FALSE}
length(intersect(colnames(Challenge), colnames(SOTA)))
#37 intersecting columns
# create a function that is the opposite of intersect
outersect <- function(x, y) {
  sort(c(setdiff(x, y),
         setdiff(y, x)))
}
length(outersect(colnames(Challenge), colnames(SOTA)))
# 149 dissimilar columns
# add a column that indicates where the samples are from 
Challenge <- Challenge %>%
    dplyr::mutate(origin = "Lab")
SOTA <- SOTA %>%
    dplyr::mutate(origin = "Field")
# Adjust the parasite names to fit the lab
SOTA <- SOTA %>%
    dplyr::mutate(eimeriaSpecies = case_when(
    eimeriaSpecies == "Negative" ~ "uninfected",
    eimeriaSpecies == "" ~ "NA",
    TRUE ~ eimeriaSpecies))
    
# Rename column names to match each other
Challenge <- Challenge %>% 
    dplyr::rename(Mouse_ID = EH_ID, delta_ct_cewe_MminusE = delta, 
                  MC.Eimeria = Eim_MC, Feces_Weight = feces_weight)
#expected columns:
37 + 149 #186
# now join the two data sets
data <- full_join(Challenge, SOTA, 
                  by = intersect(colnames(SOTA), colnames(Challenge)))
write.csv(data, "output_data/1.MICE_cleaned_data.csv", row.names = FALSE)
```

# Lab 

## Heatmap on lab gene expression: data preparation

```{r}
gene <- as_tibble(data) %>%
    dplyr::filter(origin == "Lab", infection == "challenge", dpi == dpi_max) %>%
    dplyr::group_by(Mouse_ID) %>%
    dplyr::select(c(Mouse_ID, MC.Eimeria, all_of(Gene_lab)))
gene <- unique(gene)
# turn the data frame into a matrix and transpose it. We want to have each cell 
 # type as a row name 
 gene <- t(as.matrix(gene %>% 
                         dplyr::select(c(Mouse_ID, all_of(Gene_lab)))))
 
 #switch the matrix back to a data frame format
 gene <- as.data.frame(gene)
 
 # turn the first row into column names
 gene %>%
     row_to_names(row_number = 1) -> heatmap_data
 
 table(rowSums(is.na(heatmap_data)) == nrow(heatmap_data))
 
 # turn the columns to numeric other wise the heatmap function will not work
 heatmap_data[] <- lapply(heatmap_data, function(x) as.numeric(as.character(x)))
 
 # remove columns with only NAs 
 heatmap_data <- Filter(function(x)!all(is.na(x)), heatmap_data) 
 
 #remove rows with only Nas
 heatmap_data <-  heatmap_data[, colSums(is.na(heatmap_data)) != 
                                 nrow(heatmap_data)]
 
#Prepare the annotation data frame
annotation_df <- as_tibble(data) %>%
    dplyr::filter(origin == "Lab", infection == "challenge", dpi == dpi_max)%>%
    dplyr::group_by(Mouse_ID) %>%
    dplyr::select(c("Mouse_ID", "Parasite_challenge", "infection_history",
                    "mouse_strain", "max_WL")) 
  
annotation_df <- unique(annotation_df) %>%
    dplyr::filter(Mouse_ID %in% colnames(heatmap_data))
annotation_df <- as.data.frame(annotation_df)
 ### Prepare the annotation columns for the heatmap
rownames(annotation_df) <- annotation_df$Mouse_ID
# Match the row names to the heatmap data frame
rownames(annotation_df) <- colnames(heatmap_data)
#remove the unecessary column
annotation_df <- annotation_df %>% dplyr::select(-Mouse_ID, )
heatmap_lab <- heatmap_data
```

## Heatmap on lab gene expression data: plot

```{r pheatmap_genes_lab, echo = FALSE}
pheatmap(heatmap_data, annotation_col = annotation_df, scale = "row")
```


# Field 

## Heatmap on Field gene expression: data preparation

```{r}
gene <- as_tibble(data) %>%
    dplyr::filter(origin == "Field") %>%
    dplyr::group_by(Mouse_ID) %>%
  drop_na(HI) %>%
    dplyr::select(c(Mouse_ID, all_of(Genes_wild)))
gene <- unique(gene)
# I will remove nas. See towards the end of this chunk why I did it in this way
gene <- gene %>% drop_na(MYD88)
#
# turn the data frame into a matrix and transpose it. We want to have each cell 
 # type as a row name 
 gene <- t(as.matrix(gene))
 
 #switch the matrix back to a data frame format
 gene <- as.data.frame(gene)
 
 # turn the first row into column names
 gene %>%
     row_to_names(row_number = 1) -> heatmap_data
 
 table(rowSums(is.na(heatmap_data)) == nrow(heatmap_data))
 
 # turn the columns to numeric other wise the heatmap function will not work
 heatmap_data[] <- lapply(heatmap_data, function(x) as.numeric(as.character(x)))
 
 # remove columns with only NAs 
 heatmap_data <- Filter(function(x)!all(is.na(x)), heatmap_data) 
 
 #remove rows with only Nas
 heatmap_data <-  heatmap_data[, colSums(is.na(heatmap_data)) != 
                                 nrow(heatmap_data)]
 
#Prepare the annotation data frame
annotation_df <- as_tibble(data) %>%
    dplyr::filter(origin == "Field")%>%
    dplyr::group_by(Mouse_ID) %>%
    dplyr::select(c("Mouse_ID", "Sex", "HI")) %>% 
  drop_na()
  
annotation_df <- unique(annotation_df) %>%
    dplyr::filter(Mouse_ID %in% colnames(heatmap_data))
annotation_df <- as.data.frame(annotation_df)
 ### Prepare the annotation columns for the heatmap
rownames(annotation_df) <- annotation_df$Mouse_ID
# Match the row names to the heatmap data frame
rownames(annotation_df) <- colnames(heatmap_data)
#remove the unecessary column
annotation_df <- annotation_df %>% dplyr::select(-Mouse_ID, )
# heatmap gives error: Error in hclust(d, method = method) : 
#  NA/NaN/Inf in foreign function call (arg 10)
#Let's figure this out
g_field <- data %>%
    dplyr::filter(origin == "Field") %>%
    dplyr::group_by(Mouse_ID) %>%
    dplyr::select(all_of(Genes_wild))
# How many nas?
is.na(g_field) %>% table()
is.infinite(unlist(g_field)) %>% table() #there are no infinite values
dim(g_field)
is.na(annotation_df) %>% table() # there is an na value in the annotation
# df! that is what is causing the clustering error probably
annotation_df <- annotation_df %>% drop_na()
heatmap_data <- heatmap_data %>%
  dplyr::select(row.names(annotation_df))
# I will have to reduce the nas in order to get some clustering going on
# MYD88 is the most complete gene. I will remove the nas from the nas rows of 
#this genes further up the analysis and see if it works out
heatmap_field <- heatmap_data
```

## Heatmap on lab gene expression data: plot

```{r pheatmap_genes, echo = FALSE}
#removed many rows with missing values 
pheatmap(heatmap_data, annotation_col = annotation_df, scale = "row")
```
# Lab + Field Heatmap combination

```{r}
heatmap_lab_field <- cbind(heatmap_lab, heatmap_field)
 
#Prepare the annotation data frame
annotation_df <- as_tibble(data) %>%
    dplyr::select(origin, Mouse_ID) 
  
annotation_df <- unique(annotation_df) %>%
    dplyr::filter(Mouse_ID %in% colnames(heatmap_lab_field))
annotation_df <- as.data.frame(annotation_df)
 ### Prepare the annotation columns for the heatmap
rownames(annotation_df) <- annotation_df$Mouse_ID
# Match the row names to the heatmap data frame
rownames(annotation_df) <- colnames(heatmap_lab_field)
#remove the unecessary column
annotation_df <- annotation_df %>% dplyr::select(-Mouse_ID, )
```
## Lab - Field data heatmap - plot


```{r}
pheatmap(heatmap_lab_field, annotation_col = annotation_df, scale = "row")
```





