---
title: "13.Random_forest_lab_facs"
author: "Fay"
date: '2022-11-08'
output:
  pdf_document:
    keep_md: yes 
    fig_width: 12
    fig_height: 8
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Aim: 
- Predicting health impact of infections utilizing immune parameters as predictors 
- Predicted variable: WL as a proxy of health 
- To do that we are using immune data from experimental lab infections. 
- We are training random forest models on the immune data from experimental lab infections 
- And we test them on the field.
- We then compare the differences in the predicted health impact among non-hybrid and hybrid mice. 


In this document I am preparing the models using the lab data only. 

#  Load necessary libraries:

```{r libraries, message = FALSE, warnings = FALSE}
#install.packages("optimx", version = "2021-10.12") # this package is required for 
#the parasite load package to work
library(tidyverse)
library(tidyr)
library(dplyr)
library(cowplot)
library(randomForest)
library(ggplot2)
library(caret)
library(ggpubr)
library(rfUtilities) # Implements a permutation test cross-validation for 
# Random Forests models
library(FactoMineR)

```


# Laboratory data 

## Importing the data

We start with the data from experimental lab infections. 

```{r import_data,  message = FALSE, warnings = FALSE }
#import data
hm <- read.csv("output_data/2.imputed_MICE_data_set.csv")
```

# vectors for selecting

```{r}
Facs_lab <- c("CD4", "Treg", "Div_Treg", "Treg17", "Th1", 
                    "Div_Th1", "Th17", "Div_Th17", "CD8", "Act_CD8", 
                    "Div_Act_CD8", "IFNy_CD4", "IFNy_CD8") #"Treg_prop", removed due to many missing values
                    #"IL17A_CD4"
Facs_wild <- c( "Treg", "CD4", "Treg17", "Th1", "Th17", "CD8",
                     "Act_CD8", "IFNy_CD4", "IFNy_CD8") #"IL17A_CD4", 
```



## Data cleaning / preparation

```{r cleaning}
# we need to change the  in challenge infections to a factor
hm$Parasite_challenge <- as.factor(hm$Parasite_challenge)
hm$MC.Eimeria <- as.factor(hm$MC.Eimeria)
# Here I create a new column, where we get the actual infection status
# According to the melting curve for eimeria 
hm <- hm %>%
  dplyr::mutate(current_infection = case_when(
    Parasite_challenge == "E_ferrisi" & MC.Eimeria == "TRUE" ~ "E_ferrisi",
    Parasite_challenge == "E_ferrisi" & MC.Eimeria == "FALSE" ~ "uninfected",
    Parasite_challenge == "E_falciformis" & MC.Eimeria == "TRUE" ~ "E_falciformis",
    Parasite_challenge == "E_falciformis" & MC.Eimeria == "FALSE" ~ "uninfected",
    Parasite_challenge == "uninfected" & MC.Eimeria == "TRUE" ~ "infected_eimeria",
    Parasite_challenge == "uninfected" & MC.Eimeria == "FALSE" ~ "uninfected",
    TRUE ~ ""
  ))
```




## Splitting data into training and testing sets 
Splitting between training and testing:
- Assess model performance on unseen data
- Avoid over-fitting 


# Random forest for predicting percentage of maximum weight loss

## Dividing data into training and testing 

```{r}
#select the genes and lab muce
lab <- hm %>%
  dplyr::filter(origin == "Lab", Position == "mLN") #selecting for mln to avoid

# duplicates

lab <- unique(lab)

facs_mouse <- lab %>%
  dplyr::select(c(Mouse_ID, all_of(Facs_lab)))


facs <- facs_mouse[, -1]

#remove rows with only nas
facs <- facs[,colSums(is.na(facs))<nrow(facs)]

#remove colums with only nas 
facs <- facs[rowSums(is.na(facs)) != ncol(facs), ]



#select same rows in the first table
facs_mouse <- facs_mouse[row.names(facs), ]


# we can now run a normal pca on the complete data set
res.pca <- PCA(facs)


# select the same rows from the lab data
lab <- lab[row.names(facs),]
M_W <- lab %>%
  dplyr::select(c(Mouse_ID, WL_max)) 
facs <- cbind(M_W, facs)
facs <- unique(facs) %>%
  dplyr::select(-Mouse_ID)
# split data into training and test
set.seed(123) # this will help us reproduce this random assignment
# in this way we can pick the random numbers
training.samples <- facs$WL_max%>%
  createDataPartition(p = .7, 
                      list = FALSE) 
# this is the partiicition! In this case 0.7 = training data and 0.3 = testing
# we don't want to get a list in return
train.data <- facs[training.samples, ] 
test.data <- facs[-training.samples, ] 
```



## Building the model

```{r  predicting_weight_loss_model}
#train the model
weight_loss_predict <- randomForest(WL_max ~., data = train.data, 
                                    proximity = TRUE, ntree = 1000) 
# ntree = number of trees     
# save the model 
save(weight_loss_predict, file =  "r_scripts/models/predict_weight_loss.RData")
print(weight_loss_predict)
```

## Model - quality testing

### Cross-validation 

MSE: As a brief explanation, mean squared error (MSE) is the average of the 
summation of the squared difference between the actual output value and the 
predicted output value. Our goal is to reduce the MSE as much as possible. 

Variance explained: %explained variance is a measure of how well out-of-bag 
predictions explain the target variance of the training set.

```{r}
predict_WL_cv <- rf.crossValidation(x = weight_loss_predict, xdata = train.data, 
                                    p = 0.10, n = 99, ntree = 501)
predict_WL_cv$fit.var.exp
par(mfrow=c(2,2))
plot(predict_WL_cv) 
# Root Mean Squared Error (observed vs. predicted) from each Bootstrap 
# iteration (cross-validation)
plot(predict_WL_cv, stat = "mse")
#Percent variance explained from specified fit model
plot(predict_WL_cv, stat = "var.exp")
#Mean Absolute Error from each Bootstrapped model
plot(predict_WL_cv, stat = "mae")
```



```{r}
plot(weight_loss_predict)
```
The plotted error rate above is based on the OOB sample error 
and can be accessed directly at m1$mse. Thus, we can find which number of trees 
providing the lowest error rate

```{r}
# number of trees with lowest MSE
which.min(weight_loss_predict$mse)
# RMSE of this optimal random forest
sqrt(weight_loss_predict$mse[which.min(weight_loss_predict$mse)])
```

### https://uc-r.github.io/s
RandomForest also allows us to use a validation set to measure predictive
accuracy if we did not want to use the OOB samples. 



Tutorial: 
https://hackernoon.com/random-forest-regression-in-r-code-and-interpretation

Random forest regression in R provides two outputs: decrease in mean square 
error (MSE) and node purity. Prediction error described as MSE is based on 
permuting out-of-bag sections of the data per individual tree and predictor, 
and the errors are then averaged. In the regression context, Node purity is the 
total decrease in residual sum of squares when splitting on a variable averaged 
over all trees (i.e. how well a predictor decreases variance). MSE is a more 
reliable measure of variable importance. If the two importance metrics show 
different results, listen to MSE. If all of your predictors are numerical, then 
it shouldn’t be too much of an issue


Mean Decrease Gini (IncNodePurity) - This is a measure of variable importance 
based on the Gini impurity index used for the calculating the splits in trees.

Improving Your Model
Your model depends on the quality of your dataset and the type of Machine 
Learning algorithm used. Therefore, to improve the accuracy of your model, 
you should:

Check what attributes affect our model the most and what variables to leave out 
in future analysis
Find out what other attributes affect a person's wage; we can use as predictors 
in future analysis
Tweak the algorithm (e.g. change the ntree value)
Use a different machine learning algorithm
If any of these reduces the RMSE significantly, you have succeeded in improving 
your model!

```{r include = FALSE, message = FALSE, echo = FALSE}
### Visualize variable importance ----------------------------------------------
#Call importance() function on the model model to check how the attributes used 
# as predictors affect our weight_loss_predict
importance(weight_loss_predict)
#weight_loss_predict$mse
## S3 method for class 'randomForest'
plot(weight_loss_predict, type = "l", main=deparse(substitute(x)))
varImpPlot(weight_loss_predict)
# Get variable importance from the weight_loss_predict fit
ImpData <- as.data.frame(importance(weight_loss_predict))
ImpData$Var.Names <- row.names(ImpData)
```



# Application of weight_loss_predict
## Using the testing data

Let's now make some predictions using our test data.


```{r }
#The predict() function in R is used to predict the values based on the 
# input data.
predictions <- predict(weight_loss_predict, test.data)
# assign test.data to a new object, so that we can make changes
result <- test.data
#add the new variable of predictions to the result object
result <- cbind(result, predictions)
# what is the correlation between predicted and actual data?
cor(result$WL_max, result$predictions, 
    method = c("pearson", "kendall", "spearman"))
test_lab <- lab %>%
  left_join(result, by = c("WL_max", "Treg", "CD4", "Treg17", "Th1", "Th17", "CD8",
                     "Act_CD8", "IFNy_CD4", "IFNy_CD8"))
test_lab <- test_lab %>%
  drop_na(predictions)
```



## Visualizing the predictions


```{r predictions_weight_loss_predict_wL }
# trying to find a way to represent the delta ct for the negative ones
# please find a better way to do this
test_lab <- test_lab %>% 
    dplyr::mutate(Infection_intensity = case_when(
       Parasite_challenge == "uninfected" ~ -9,
       TRUE ~ delta_ct_cewe_MminusE
    ))
cor(test_lab$predictions, test_lab$WL_max)
test_lab   %>%
  ggplot(aes(x = predictions, y = WL_max, color = current_infection, 
                 size = delta_ct_cewe_MminusE)) +
  labs(x = "Predictions: Maximum weight loss", 
       y = "Observed: Maximum weight loss") +
  geom_point(aes(x = predictions, y = WL_max, 
                 color = Parasite_challenge, size = Infection_intensity)) +
  labs(x = "Predictions: Maximum weight loss", 
       y = "Observed: Maximum weight loss") +
    theme_bw()
cor.test(result$predictions, result$WL_max, method = "spearman")
  
```

# Repeating the process using the whole data set as a training data set

