---
title: "3. Exploration Gene "
author: "Fay"
date: '2022-10-05'
output:
  pdf_document: 
    keep_md: yes
    fig_width: 12
    fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


```{r load_libraries, echo=FALSE, include = FALSE}
library(tidyverse)
library(tidyr)
library(dplyr)
library(janitor)
library(visdat)
library(corrplot)
library(RColorBrewer)
library(ggplot2)
```

# Import data
```{r}
MICE <- read.csv("output_data/1.MICE_cleaned_data.csv")
```

# Vectors for selecting genes
```{r}
Gene_lab   <- c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10",
                "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                "TICAM1", "TNF") #"IL.12", "IRG6")

Genes_wild   <- c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10", 
                  "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                  "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                  "TICAM1", "TNF") #, "IL.12", "IRG6")
```


# Lab data 
## Correlations between the genes 


```{r gene_correlations, echo = FALSE}
gene_correlation <- MICE %>% 
  filter(origin == "Lab", infection == "challenge", dpi == dpi_max) %>%
  ungroup() %>%
  dplyr::select(all_of(Gene_lab))

# draw correlation between the genes
gene_correlation <- as.matrix(cor(gene_correlation, 
                                  use="pairwise.complete.obs"))

# load the function to calculate the p value for correlations
source("r_scripts/functions/p_value_for_correlations.R")

# matrix of the p-value of the correlatio
p.mat <- cor.mtest(gene_correlation)
```


## Corrplot of correlations - Laboratory gene expression data

Here is a corrplot of the correlations between the genes.

```{r corrplot_correlations_genes, echo = FALSE}
corrplot(gene_correlation, 
         method = "circle",  #method of the plot, "color" would show colour gradient
         tl.col = "black", tl.srt=45, #colour of labels and rotation
         col = brewer.pal(n = 8, name ="RdYlBu"), #colour of matrix
         order="hclust", #hclust reordering
         p.mat = p.mat, sig.level = 0.01, insig = "blank",
         addCoef.col = 'black',
         number.cex=0.5) 
  #Add significance level to the correlogram
#remove the values that are insignificant
```



## Gene expression versus infection intensities - laboratory infections

```{r, echo = FALSE, message = FALSE}

g <- MICE %>% 
  filter(origin == "Lab", infection == "challenge", dpi == dpi_max) %>%
  dplyr::group_by(Mouse_ID) %>%
  pivot_longer(cols = all_of(Gene_lab), names_to = "Gene", 
               values_to = "gene_expression")

g <- unique(g)


g %>%
  ggplot(aes(x = Parasite_challenge, y = gene_expression, 
             color = Parasite_challenge)) + 
  geom_boxplot() +
  geom_jitter() +
  facet_wrap(~ Gene) +
  labs(x = "Infection groups, E64 = E. ferrisi, E88 = E.falciformis, 
         UNI = Uninfected", y = "Gene expression",
         title = "Gene expression in response to infection group, 
         laboratory infections") +
  theme_bw() +
  theme(axis.text.x=element_blank(),
        plot.title = element_text(size=30))



```








```{r}

g %>%
    filter(!Parasite_challenge == "uninfected", MC.Eimeria == "TRUE") %>%
    ggplot(aes(x = delta_ct_cewe_MminusE, y = gene_expression, 
               color = Parasite_challenge)) +
    geom_jitter() +
    facet_wrap(~ Gene, scales = "free") +
    theme_light() +
    labs(x = "Delta Ct, Infection intensity", y = "Gene expression",
         title = "Gene expression in response to infection intensity - 
         laboratory infections") +
    theme_bw() +
  theme(plot.title = element_text(size= 30))

```









```{r histogram_genes, echo = FALSE, message=FALSE}
g  %>%
  ggplot(aes(x = gene_expression, color = challenge_infection)) + 
  geom_histogram(echo = FALSE) +
  labs(x = "Delta Ct, Infection intensity", y = "Frequency",
         title = "Gene expression in response to infecting parasite - 
         laboratory infections") +
  theme(plot.title = element_text(size= 30))
```




###############################
# Repeating for field infections

# Field data 
## Correlations between the genes 


```{r, echo = FALSE}
gene_correlation <- MICE %>% 
  filter(origin == "Field") %>%
  ungroup() %>%
  dplyr::select(all_of(Genes_wild))

# draw correlation between the genes
gene_correlation <- as.matrix(cor(gene_correlation, use="pairwise.complete.obs"))


# matrix of the p-value of the correlatio
p.mat <- cor.mtest(gene_correlation)
```


## Corrplot of correlations - Field gene expression data

Here is a corrplot of the correlations between the genes.
I am using the non-normalized genes

```{r, echo = FALSE}

corrplot(gene_correlation, 
         method = "circle",  #method of the plot, "color" would show colour gradient
         tl.col = "black", tl.srt=45, #colour of labels and rotation
         col = brewer.pal(n = 8, name ="RdYlBu"), #colour of matrix
         order="hclust", #hclust reordering
         p.mat = p.mat, sig.level = 0.01, insig = "blank",
         addCoef.col = 'black',
         number.cex=0.5) #Add significance level to the correlogram
#remove the values that are insignificant
```

## Gene expression versus infection intensities - Field infections



```{r , echo = FALSE, message = FALSE}

g <- MICE %>% 
  filter(origin == "Field") %>%
  dplyr::group_by(Mouse_ID) %>%
  pivot_longer(cols = all_of(Genes_wild), names_to = "Gene", 
               values_to = "gene_expression")

g <- unique(g)


g %>%
  ggplot(aes(x = eimeriaSpecies, y = gene_expression, 
             color = eimeriaSpecies)) + 
  geom_boxplot() +
  geom_jitter() +
  facet_wrap(~ Gene) +
    labs(x = "Infective parasite species", y = "Gene expression",
         title = "Gene expression in response to parasite species -
         field data") +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        plot.title = element_text(size= 30))



```




```{r}

g %>%
    filter(!eimeriaSpecies == "uninfected", MC.Eimeria == "TRUE") %>%
    ggplot(aes(x = delta_ct_cewe_MminusE, y = gene_expression, 
               color = eimeriaSpecies)) +
    geom_jitter() +
    facet_wrap(~ Gene, scales = "free") +
    theme_light() +
    labs(x = "Delta Ct, Infection intensity", y = "Gene expression",
         title = "Gene expression in response to infection intensity - 
         field data") +
    theme_bw() +
  theme(plot.title = element_text(size= 30))


```






```{r , echo = FALSE, message=FALSE}
g  %>%
  ggplot(aes(x = gene_expression, color = eimeriaSpecies)) + 
  geom_histogram(echo = FALSE) +
  labs(x = "Delta Ct, Infection intensity", y = "Frequency",
         title = "Gene expression in response to infecting parasite - 
         field data") +
  theme(plot.title = element_text(size= 30))

```

#####################
# Gene expression correlations in lab and field data


```{r, echo = FALSE}
gene_correlation1 <- MICE %>% 
  filter(origin == "Lab", infection == "challenge", dpi == dpi_max) %>%
  ungroup() %>%
  dplyr::select(c(Mouse_ID, all_of(Gene_lab)))

gene_correlation2 <- MICE %>% 
  filter(origin == "Field") %>%
  ungroup() %>%
  dplyr::select(c(Mouse_ID, all_of(Genes_wild)))

gene_correlation <- rbind(gene_correlation1, gene_correlation2)

gene_correlation <- gene_correlation %>% 
  dplyr::select(-Mouse_ID)

# draw correlation between the genes
gene_correlation <- as.matrix(cor(gene_correlation, 
                                  use="pairwise.complete.obs"))


# matrix of the p-value of the correlatio
p.mat <- cor.mtest(gene_correlation)
```


## Corrplot of correlations - *Laboratory and field* gene expression data

Here is a corrplot of the correlations between the genes.

```{r, echo = FALSE}
corrplot(gene_correlation, 
         method = "circle",  #method of the plot, "color" would show colour gradient
         tl.col = "black", tl.srt=45, #colour of labels and rotation
         col = brewer.pal(n = 8, name ="RdYlBu"), #colour of matrix
         order="hclust", #hclust reordering
         p.mat = p.mat, sig.level = 0.01, insig = "blank",
         addCoef.col = 'black',
         number.cex=0.5) 
  #Add significance level to the correlogram
#remove the values that are insignificant
```