---
title: "4.1_gene_normalization.rmd"
author: "Fay"
date: '2022-11-01'
output:
  pdf_lab_document:
    keep_md: yes 
    fig_width: 12
    fig_height: 8

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




# Load libraries



```{r}
library(mice)
library(tidyr)
library(tidyverse)
library(dplyr)
```

# Load data

# Import data

```{r}
hm <- read.csv("output_data/2.imputed_MICE_data_set.csv")
```

I only include GAPDH as a housekeeping gene, as PPIB is missing in a large number

```{r}
# Vectors for selecting genes
#Lab genes
# The measurements of IL.12 and IRG6 are done with an other assay and will 
#ignore for now
Gene_lab   <- c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10",
                "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                "TICAM1", "TNF") #"IL.12", "IRG6")

Genes_wild   <- c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10", 
                  "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                  "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                  "TICAM1", "TNF") #, "IL.12", "IRG6")

Facs_lab <- c("CD4", "Treg", "Div_Treg", "Treg17", "Th1", 
                    "Div_Th1", "Th17", "Div_Th17", "CD8", "Act_CD8", 
                    "Div_Act_CD8", "IFNy_CD4", "IFNy_CD8","Treg_prop", 
                    "IL17A_CD4")  

Facs_wild <- c( "Treg", "CD4", "Treg17", "Th1", "Th17", "CD8",
                     "Act_CD8", "IFNy_CD4", "IL17A_CD4", "IFNy_CD8")
```





Do we have all the measurements of our housekeeping genes?

```{r}
#
field <- hm %>% 
  filter(origin == "Field")
lab <- hm %>%
  filter(origin == "Lab")

#dplyr::select(-Mouse_ID)
# looking at patterns of nas)
#pattern_na <-as.data.frame(md.pattern(field_genes))
sapply(field %>%
         dplyr::select(c(all_of(Genes_wild), "PPIB", "GAPDH")), 
                      function(x) sum(is.na(x)))

sapply(lab %>%
         dplyr::select(c(all_of(Gene_lab), "PPIB", "GAPDH")), 
                      function(x) sum(is.na(x)))


# remove old wrongly normalized measures
hm <- hm %>% 
  dplyr::select(-contains("_N"))
         
```


# Gene normalization 



For the field samples the missing values in the house-keeping genes are:
PPIB: 212
GAPDH: 10

For the lab samples the missing values in the house-keeping genes are:
PPIB: 1
GAPDH: 96

Downlad and install the package for Gene normalization


Step 1. Normalize to (REF): ∆Cq = Cq (TAR) – Cq (REF)
Step 2. Exponential expression transform: ∆Cq Expression = 2–∆Cq
Step 3. Average replicates and calculate standard deviation
Step 4. Normalize to treatment control
Step 5. % KD = (1 – ∆∆Cq ) × 100

```{r}
# create a count data set, rows are sample names and columns the genes
count_data <- hm %>%
  dplyr::select(c(Mouse_ID, origin, all_of(Gene_lab), -IL.10 , GAPDH, PPIB))

####################### field ##########################
# select first the field samples 
count_field <- count_data %>%
  dplyr::filter(origin == "Field")

df <- count_field

############### IL.13 
# dct
df <- df %>%
  mutate(IL.13_dct = IL.13 - GAPDH)
# mean of dct
dct_mean <- mean(df$IL.13_dct, na.rm = TRUE)
#fold gene expression
 df <- df %>%
   mutate(IL.13_N = 2^ - (IL.13_dct - dct_mean)) %>%
   mutate(IL.13_N = round(IL.13_N, digits = 2))
 
############# I tried writing a nice function but I failed so now I am
 # going t repeat this many many times
 
 ############### IFNy
 # dct
df <- df %>%
  mutate(IFNy_dct = IFNy - GAPDH)

# mean of dct
dct_mean <- mean(df$IFNy_dct, na.rm = TRUE)
 
#fold gene expression
 df <- df %>%
   mutate(IFNy_N = 2^ - (IFNy_dct - dct_mean)) %>%
   mutate(IFNy_N = round(IFNy_N, digits = 2))
 

 ############### CXCR3
 df <- df %>%
  mutate(CXCR3_dct = CXCR3 - GAPDH)

# mean of dct
dct_mean <- mean(df$CXCR3_dct, na.rm = TRUE)
 
#fold gene expression
 df <- df %>%
   mutate(CXCR3_N = 2^ - (CXCR3_dct - dct_mean)) %>%
   mutate(CXCR3_N = round(CXCR3_N, digits = 2))
 
 ############### IL.6
 df <- df %>%
  mutate(IL.6_dct = IL.6 - GAPDH)

# mean of dct
dct_mean <- mean(df$IL.6_dct, na.rm = TRUE)
 
#fold gene expression
 df <- df %>%
   mutate(IL.6_N = 2^ - (IL.6_dct - dct_mean)) %>%
   mutate(IL.6_N = round(IL.6_N, digits = 2))

 ##############  IL1RN
 df <- df %>%
  mutate(IL1RN_dct = IL1RN - GAPDH)

# mean of dct
dct_mean <- mean(df$IL1RN_dct, na.rm = TRUE)
 
#fold gene expression
 df <- df %>%
   mutate(IL1RN_N = 2^ - (IL1RN_dct - dct_mean)) %>%
   mutate(IL1RN_N = round(IL1RN_N, digits = 2))
 
 
 ##############  CASP1
 df <- df %>%
  mutate(CASP1_dct = CASP1 - GAPDH)

# mean of dct
dct_mean <- mean(df$CASP1_dct, na.rm = TRUE)
 
#fold gene expression
 df <- df %>%
   mutate(CASP1_N = 2^ - (CASP1_dct - dct_mean)) %>%
   mutate(CASP1_N = round(CASP1_N, digits = 2))
 
 
 ##############  CXCL9
 df <- df %>%
  mutate(CXCL9_dct = CXCL9 - GAPDH)

# mean of dct
dct_mean <- mean(df$CXCL9_dct, na.rm = TRUE)
 
#fold gene expression
 df <- df %>%
   mutate(CXCL9_N = 2^ - (CXCL9_dct - dct_mean)) %>%
   mutate(CXCL9_N = round(CXCL9_N, digits = 2))
 
  ##############  IDO1
 df <- df %>%
  mutate(IDO1_dct = IDO1 - GAPDH)
# mean of dct
dct_mean <- mean(df$IDO1_dct, na.rm = TRUE)
#fold gene expression
 df <- df %>%
   mutate(IDO1_N = 2^ - (IDO1_dct - dct_mean)) %>%
   mutate(IDO1_N = round(IDO1_N, digits = 2))
 
  ##############  IRGM1
 df <- df %>%
  mutate(IRGM1_dct = IRGM1 - GAPDH)
# mean of dct
dct_mean <- mean(df$IRGM1_dct, na.rm = TRUE)
#fold gene expression
 df <- df %>%
   mutate(IRGM1_N = 2^ - (IRGM1_dct - dct_mean)) %>%
   mutate(IRGM1_N = round(IRGM1_N, digits = 2))
 
  ##############  MPO
 df <- df %>%
  mutate(MPO_dct = MPO - GAPDH)
# mean of dct
dct_mean <- mean(df$MPO_dct, na.rm = TRUE)
#fold gene expression
 df <- df %>%
   mutate(MPO_N = 2^ - (MPO_dct - dct_mean)) %>%
   mutate(MPO_N = round(MPO_N, digits = 2))
 
 
  ##############  MUC2
 df <- df %>%
  mutate(MUC2_dct = MUC2 - GAPDH)
# mean of dct
dct_mean <- mean(df$MUC2_dct, na.rm = TRUE)
#fold gene expression
 df <- df %>%
   mutate(MUC2_N = 2^ - (MUC2_dct - dct_mean)) %>%
   mutate(MUC2_N = round(MUC2_N, digits = 2))
 
  ##############  MUC5AC
 df <- df %>%
  mutate(MUC5AC_dct = MUC5AC - GAPDH)
# mean of dct
dct_mean <- mean(df$MUC5AC_dct, na.rm = TRUE)
#fold gene expression
 df <- df %>%
   mutate(MUC5AC_N = 2^ - (MUC5AC_dct - dct_mean)) %>%
   mutate(MUC5AC_N = round(MUC5AC_N, digits = 2))
 
  ##############  MYD88
 df <- df %>%
  mutate(MYD88_dct = MYD88 - GAPDH)
# mean of dct
dct_mean <- mean(df$MYD88_dct, na.rm = TRUE)
#fold gene expression
 df <- df %>%
   mutate(MYD88_N = 2^ - (MYD88_dct - dct_mean)) %>%
   mutate(MYD88_N = round(MYD88_N, digits = 2))
 
  ##############  NCR1
 df <- df %>%
  mutate(NCR1_dct = NCR1 - GAPDH)
# mean of dct
dct_mean <- mean(df$NCR1_dct, na.rm = TRUE)
#fold gene expression
 df <- df %>%
   mutate(NCR1_N = 2^ - (NCR1_dct - dct_mean)) %>%
   mutate(NCR1_N = round(NCR1_N, digits = 2))
 
 
  ##############  PRF1
 df <- df %>%
  mutate(PRF1_dct = PRF1 - GAPDH)
# mean of dct
dct_mean <- mean(df$PRF1_dct, na.rm = TRUE)
#fold gene expression
 df <- df %>%
   mutate(PRF1_N = 2^ - (PRF1_dct - dct_mean)) %>%
   mutate(PRF1_N = round(PRF1_N, digits = 2))
 
  ##############  RETNLB
 df <- df %>%
  mutate(RETNLB_dct = RETNLB - GAPDH)
# mean of dct
dct_mean <- mean(df$RETNLB_dct, na.rm = TRUE)
#fold gene expression
 df <- df %>%
   mutate(RETNLB_N = 2^ - (RETNLB_dct - dct_mean)) %>%
   mutate(RETNLB_N = round(RETNLB_N, digits = 2))
 
 
 
  ##############  SOCS1
 df <- df %>%
  mutate(SOCS1_dct = SOCS1 - GAPDH)
# mean of dct
dct_mean <- mean(df$SOCS1_dct, na.rm = TRUE)
#fold gene expression
 df <- df %>%
   mutate(SOCS1_N = 2^ - (SOCS1_dct - dct_mean)) %>%
   mutate(SOCS1_N = round(SOCS1_N, digits = 2))
 
  ##############  TICAM1
 df <- df %>%
  mutate(TICAM1_dct = TICAM1 - GAPDH)
# mean of dct
dct_mean <- mean(df$TICAM1_dct, na.rm = TRUE)
#fold gene expression
 df <- df %>%
   mutate(TICAM1_N = 2^ - (TICAM1_dct - dct_mean)) %>%
   mutate(TICAM1_N = round(TICAM1_N, digits = 2))
 
 ##############  TNF
 df <- df %>%
  mutate(TNF_dct = TNF - GAPDH)
# mean of dct
dct_mean <- mean(df$TNF_dct, na.rm = TRUE)
#fold gene expression
 df <- df %>%
   mutate(TNF_N = 2^ - (TNF_dct - dct_mean)) %>%
   mutate(TNF_N = round(TNF_N, digits = 2))
  
 
 
 ################################# lab 
 # select first the field samples 
count_Lab <- count_data %>%
  dplyr::filter(origin == "Lab")

df_lab <- count_Lab

  ############### IFNy
  df_lab <- df_lab %>%
  mutate(IFNy_dct = IFNy - PPIB)
# mean of dct
dct_mean <- mean(df_lab$IFNy_dct, na.rm = TRUE)
#fold gene expression
 df_lab <- df_lab %>%
   mutate(IFNy_N = 2^ - (IFNy_dct - dct_mean)) %>%
   mutate(IFNy_N = round(IFNy_N, digits = 2))
 
 
 
 ############### CXCR3
  df_lab <- df_lab %>%
  mutate(CXCR3_dct = CXCR3 - PPIB)
# mean of dct
dct_mean <- mean(df_lab$CXCR3_dct, na.rm = TRUE)
#fold gene expression
 df_lab <- df_lab %>%
   mutate(CXCR3_N = 2^ - (CXCR3_dct - dct_mean)) %>%
   mutate(CXCR3_N = round(CXCR3_N, digits = 2))
 
 ############### IL.6
   df_lab <- df_lab %>%
  mutate(IL.6_dct = IL.6 - PPIB)
# mean of dct
dct_mean <- mean(df_lab$IL.6_dct, na.rm = TRUE)
#fold gene expression
 df_lab <- df_lab %>%
   mutate(IL.6_N = 2^ - (IL.6_dct - dct_mean)) %>%
   mutate(IL.6_N = round(IL.6_N, digits = 2))
 
 
 ############## IL.13
   df_lab <- df_lab %>%
  mutate(IL.13_dct = IL.13 - PPIB)
# mean of dct
dct_mean <- mean(df_lab$IL.13_dct, na.rm = TRUE)
#fold gene expression
 df_lab <- df_lab %>%
   mutate(IL.13_N = 2^ - (IL.13_dct - dct_mean)) %>%
   mutate(IL.13_N = round(IL.13_N, digits = 2))
 
 ##############  IL1RN
   df_lab <- df_lab %>%
  mutate(IL1RN_dct = IL1RN - PPIB)
# mean of dct
dct_mean <- mean(df_lab$IL1RN_dct, na.rm = TRUE)
#fold gene expression
 df_lab <- df_lab %>%
   mutate(IL1RN_N = 2^ - (IL1RN_dct - dct_mean)) %>%
   mutate(IL1RN_N = round(IL1RN_N, digits = 2))
 
 ##############  CASP1
   df_lab <- df_lab %>%
  mutate(CASP1_dct = CASP1 - PPIB)
# mean of dct
dct_mean <- mean(df_lab$CASP1_dct, na.rm = TRUE)
#fold gene expression
 df_lab <- df_lab %>%
   mutate(CASP1_N = 2^ - (CASP1_dct - dct_mean)) %>%
   mutate(CASP1_N = round(CASP1_N, digits = 2))
 
 
 ##############  CXCL9
   df_lab <- df_lab %>%
  mutate(CXCL9_dct = CXCL9 - PPIB)
# mean of dct
dct_mean <- mean(df_lab$CXCL9_dct, na.rm = TRUE)
#fold gene expression
 df_lab <- df_lab %>%
   mutate(CXCL9_N = 2^ - (CXCL9_dct - dct_mean)) %>%
   mutate(CXCL9_N = round(CXCL9_N, digits = 2))
 
 
  ##############  IDO1
   df_lab <- df_lab %>%
  mutate(IDO1_dct = IDO1 - PPIB)
# mean of dct
dct_mean <- mean(df_lab$IDO1_dct, na.rm = TRUE)
#fold gene expression
 df_lab <- df_lab %>%
   mutate(IDO1_N = 2^ - (IDO1_dct - dct_mean)) %>%
   mutate(IDO1_N = round(IDO1_N, digits = 2))
 
  ##############  IRGM1
   df_lab <- df_lab %>%
  mutate(IRGM1_dct = IRGM1 - PPIB)
# mean of dct
dct_mean <- mean(df_lab$IRGM1_dct, na.rm = TRUE)
#fold gene expression
 df_lab <- df_lab %>%
   mutate(IRGM1_N = 2^ - (IRGM1_dct - dct_mean)) %>%
   mutate(IRGM1_N = round(IRGM1_N, digits = 2))
 
  ##############  MPO
   df_lab <- df_lab %>%
  mutate(MPO_dct = MPO - PPIB)
# mean of dct
dct_mean <- mean(df_lab$MPO_dct, na.rm = TRUE)
#fold gene expression
 df_lab <- df_lab %>%
   mutate(MPO_N = 2^ - (MPO_dct - dct_mean)) %>%
   mutate(MPO_N = round(MPO_N, digits = 2))
 
  ##############  MUC2
   df_lab <- df_lab %>%
  mutate(MUC2_dct = MUC2 - PPIB)
# mean of dct
dct_mean <- mean(df_lab$MUC2_dct, na.rm = TRUE)
#fold gene expression
 df_lab <- df_lab %>%
   mutate(MUC2_N = 2^ - (MUC2_dct - dct_mean)) %>%
   mutate(MUC2_N = round(MUC2_N, digits = 2))
 
  ##############  MUC5AC
   df_lab <- df_lab %>%
  mutate(MUC5AC_dct = MUC5AC - PPIB)
# mean of dct
dct_mean <- mean(df_lab$MUC5AC_dct, na.rm = TRUE)
#fold gene expression
 df_lab <- df_lab %>%
   mutate(MUC5AC_N = 2^ - (MUC5AC_dct - dct_mean)) %>%
   mutate(MUC5AC_N = round(MUC5AC_N, digits = 2))
 
 
  ##############  MYD88
   df_lab <- df_lab %>%
  mutate(MYD88_dct = MYD88 - PPIB)
# mean of dct
dct_mean <- mean(df_lab$MYD88_dct, na.rm = TRUE)
#fold gene expression
 df_lab <- df_lab %>%
   mutate(MYD88_N = 2^ - (MYD88_dct - dct_mean)) %>%
   mutate(MYD88_N = round(MYD88_N, digits = 2))
 
 
  ##############  NCR1
   df_lab <- df_lab %>%
  mutate(NCR1_dct = NCR1 - PPIB)
# mean of dct
dct_mean <- mean(df_lab$NCR1_dct, na.rm = TRUE)
#fold gene expression
 df_lab <- df_lab %>%
   mutate(NCR1_N = 2^ - (NCR1_dct - dct_mean)) %>%
   mutate(NCR1_N = round(NCR1_N, digits = 2))
 
  ##############  PRF1
   df_lab <- df_lab %>%
  mutate(PRF1_dct = PRF1 - PPIB)
# mean of dct
dct_mean <- mean(df_lab$PRF1_dct, na.rm = TRUE)
#fold gene expression
 df_lab <- df_lab %>%
   mutate(PRF1_N = 2^ - (PRF1_dct - dct_mean)) %>%
   mutate(PRF1_N = round(PRF1_N, digits = 2))
 
  ##############  RETNLB
   df_lab <- df_lab %>%
  mutate(RETNLB_dct = RETNLB - PPIB)
# mean of dct
dct_mean <- mean(df_lab$RETNLB_dct, na.rm = TRUE)
#fold gene expression
 df_lab <- df_lab %>%
   mutate(RETNLB_N = 2^ - (RETNLB_dct - dct_mean)) %>%
   mutate(RETNLB_N = round(RETNLB_N, digits = 2))
 
  ##############  SOCS1
   df_lab <- df_lab %>%
  mutate(SOCS1_dct = SOCS1 - PPIB)
# mean of dct
dct_mean <- mean(df_lab$SOCS1_dct, na.rm = TRUE)
#fold gene expression
 df_lab <- df_lab %>%
   mutate(SOCS1_N = 2^ - (SOCS1_dct - dct_mean)) %>%
   mutate(SOCS1_N = round(SOCS1_N, digits = 2))
 
  ##############  TICAM1
   df_lab <- df_lab %>%
  mutate(TICAM1_dct = TICAM1 - PPIB)
# mean of dct
dct_mean <- mean(df_lab$TICAM1_dct, na.rm = TRUE)
#fold gene expression
 df_lab <- df_lab %>%
   mutate(TICAM1_N = 2^ - (TICAM1_dct - dct_mean)) %>%
   mutate(TICAM1_N = round(TICAM1_N, digits = 2))
 
 ##############  TNF
   df_lab <- df_lab %>%
  mutate(TNF_dct = TNF - PPIB)
# mean of dct
dct_mean <- mean(df_lab$TNF_dct, na.rm = TRUE)
#fold gene expression
 df_lab <- df_lab %>%
   mutate(TNF_N = 2^ - (TNF_dct - dct_mean)) %>%
   mutate(TNF_N = round(TNF_N, digits = 2))
 
 
 # combine the two data frames
 genes_N <- rbind(df, df_lab)
```

#clean the data frame
```{r}
genes_N <- genes_N %>%
  dplyr::select(-c("IFNy", "CXCR3", "IL.6", "IL.13", 
                "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                "TICAM1", "TNF" , contains(("_dct"))))
hm <- hm %>% 
  dplyr::select(-all_of(Genes_wild)) %>%
  left_join(genes_N, by = c("origin", "Mouse_ID") )
  # remove non-normalized genes 
  

colnames(hm) <- str_remove_all(string = colnames(hm), pattern = "_N")



```







```{r}
 ##save the imputed data 
write.csv(hm, "output_data/2.imputed_MICE_data_set.csv", row.names = FALSE)
```
