---
title: "4.4_Mice_imputation_comb.rmd"
author: "Fay"
date: '2022-11-01'
output:
  pdf_document:
    keep_md: yes 
    fig_width: 12
    fig_height: 8

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




# Load libraries



```{r}
library(mice)
library(tidyr)
library(tidyverse)
library(VIM)
library(fitdistrplus)
library(fitur)
library(visdat)
```

# Load data

# Import data

```{r}
hm <- read.csv("output_data/MICE.csv")
```

```{r}
# Vectors for selecting genes

#Lab genes
# The measurements of IL.12 and IRG6 are done with an other assay and will 
#ignore for now
Gene_lab   <- c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10",
                "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                "TICAM1", "TNF") #"IL.12", "IRG6")

Genes_wild   <- c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10", 
                  "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                  "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                  "TICAM1", "TNF") #, "IL.12", "IRG6")

Facs_lab <- c("CD4", "Treg", "Div_Treg", "Treg17", "Th1", 
                    "Div_Th1", "Th17", "Div_Th17", "CD8", "Act_CD8", 
                    "Div_Act_CD8", "IFNy_CD4", "IFNy_CD8","Treg_prop", 
                    "IL17A_CD4")  

Facs_wild <- c( "Treg", "CD4", "Treg17", "Th1", "Th17", "CD8",
                     "Act_CD8", "IFNy_CD4", "IL17A_CD4", "IFNy_CD8")

```



# data imputation

# Genes
```{r imputing_mice}


field <- hm %>%
  dplyr::filter(origin == "Field") 

field <- unique(field)


genes_mouse_field <- field %>%
  dplyr::select(c(Mouse_ID, all_of(Genes_wild))) 


genes <- genes_mouse_field  %>%
  dplyr::select(-Mouse_ID)


#remove rows with only nas
genes <- genes[,colSums(is.na(genes))<nrow(genes)]

#remove colums with only nas 
genes <- genes[rowSums(is.na(genes)) != ncol(genes), ]


genes_mouse_field <- genes_mouse_field[row.names(genes), ]

##select same rows in the first table
field <- field[row.names(genes), ]


###############lab
#select the genes and lab muce
lab <- hm %>%
  dplyr::filter(origin == "Lab", Position == "mLN") #selecting for mln to avoid
# duplicates

lab <- unique(lab)

gene_lab_mouse <- lab %>%
  dplyr::select(c(Mouse_ID, all_of(Gene_lab))) 

gene_lab_mouse <- unique(gene_lab_mouse)

genes_lab <- gene_lab_mouse[, -1]

#remove rows with only nas
genes_lab <- genes_lab[,colSums(is.na(genes_lab))<nrow(genes_lab)]

#remove colums with only nas 
genes_lab <- genes_lab[rowSums(is.na(genes_lab)) != ncol(genes_lab), ]

genes_lab <- unique(genes_lab)

#select same rows in the first table
gene_lab_mouse <- gene_lab_mouse[row.names(genes_lab), ]

##select same rows in the first table
lab <- lab[row.names(genes_lab), ]

hm_genes <- rbind(gene_lab_mouse, genes_mouse_field)

hm_selection_g <- rbind(lab, field)

genes <- hm_genes %>%
  dplyr::select(-Mouse_ID)


# looking at patterns of nas
#pattern_na <-as.data.frame(md.pattern(field_genes))

sapply(hm_genes, function(x) sum(is.na(x)))



#had to remove as they were disturbing the imputation: Worms_presence, MC.Eimeria.FEC,  Heligmosomoides_polygurus, Zfy2, Y,  MpiC,

#vis_miss(field)


# The frequency distribution of the missing cases per variable can be obtained 
# as:
init <- mice(genes, maxit = 0)


#we want to impute only the specific variables
meth <- init$method



```


```{r}
# m=5 refers to the number of imputed datasets. Five is the default value.
igf <- mice(genes, m = 5, seed = 500) # method = meth,
summary(igf)

# to check each column with imputed data
## igf$imp$IFNy

#Now we can get back the completed dataset using the complete()
complete_genes <- complete(igf, 1)


#sapply(complete_field, function(x) sum(is.na(x)))


#visualize missingness
vis_dat(complete_genes)


#remove the non imputed genes from our data set
hm_selection_g <- hm_selection_g %>%
  dplyr::select(-c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10",
                "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                "TICAM1", "TNF"))

# add the new imputed genes to the data
hm_selection_g <- cbind(hm_selection_g, complete_genes)

```




Let’s compare the distributions of original and imputed data using a some useful 
plots.First of all we can use a scatterplot and plot Ozone against all the other 
variables. Let's first plot the variables for which we have few missing values.



```{r distr_orig_pred}

xyplot(igf, IFNy ~ IL.13 + IRGM1 + MUC2, pch=18,cex=1)


```


What we would like to see is that the shape of the magenta points (imputed) 
matches the shape of the blue ones (observed). The matching shape tells us that 
the imputed values are indeed “plausible values”.

Now let's plot the variables with many missing data points.


```{r distr_orig_pred2}

xyplot(igf,IFNy ~ IL.10 + PRF1 + CASP1, pch=18,cex=1)
```



```{r }
stripplot(igf, pch = 20, cex = 1.2)
```



```{r }
densityplot(igf)
```

The density of the imputed data for each imputed dataset is showed in magenta 
while the density of the observed data is showed in blue. Again, under our 
previous assumptions we expect the distributions to be similar.

Another useful visual take on the distributions can be obtained using the 
stripplot() function that shows the distributions of the variables as individual 
points




# Just add here the lab data and join the two tables 


```{r}



 ##save the imputed data 
write.csv(hm_selection_g, "output_data/imputed_mice.csv", row.names = FALSE)


```

