---
title: "4.Mice_imputation_field.rmd"
author: "Fay"
date: '2022-11-01'
output:
  pdf_document: 
    fig_width: 12
    fig_height: 8

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#################### field


# Load libraries



```{r}
library(mice)
library(tidyr)
library(tidyverse)
library(VIM)
library(fitdistrplus)
library(fitur)
library(visdat)
```

# Load data

# Import data

```{r}
hm <- read.csv("output_data/MICE.csv")
```

```{r}
# Vectors for selecting genes

#Lab genes
# The measurements of IL.12 and IRG6 are done with an other assay and will 
#ignore for now
Gene_lab   <- c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10",
                "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                "TICAM1", "TNF") #"IL.12", "IRG6")

Genes_wild   <- c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10", 
                  "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                  "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                  "TICAM1", "TNF") #, "IL.12", "IRG6")

Facs_lab <- c("CD4", "Treg", "Div_Treg", "Treg17", "Th1", 
                    "Div_Th1", "Th17", "Div_Th17", "CD8", "Act_CD8", 
                    "Div_Act_CD8", "IFNy_CD4", "IFNy_CD8","Treg_prop", 
                    "IL17A_CD4")  

Facs_wild <- c( "Treg", "CD4", "Treg17", "Th1", "Th17", "CD8",
                     "Act_CD8", "IFNy_CD4", "IL17A_CD4", "IFNy_CD8")

```



# Field data imputation

# Genes
```{r imputing_mice}

# field samples
field <- hm %>%  filter(origin == "Field")

gf_field <- field %>% 
  dplyr::select(all_of(c(Genes_wild)))

vis_dat(gf_field)

#remove rows with only nas
gf_field <- gf_field[,colSums(is.na(gf_field))<nrow(gf_field)]

#remove colums with only nas 
gf_field <- gf_field[rowSums(is.na(gf_field)) != ncol(gf_field), ]


#select same rows in the first table
field_genes <- field[row.names(gf_field), ]


#remove wrongly normalized genes
field <- field %>%
  dplyr::select(-ends_with("_N"))

# really removing empty columns
field_genes <- field_genes %>% 
  discard(~all(is.na(.) | . ==""))



# looking at patterns of nas
#pattern_na <-as.data.frame(md.pattern(field_genes))

#sapply(field, function(x) sum(is.na(x)))

#select the relevant columns to use for the imputation
field_genes <- field_genes %>%
  dplyr::select(c(Mouse_ID, MC.Eimeria, delta_ct_cewe_MminusE, Sex, Longitude, Latitude, 
                  Year, mtBamH, YNPAR, X332, X347, X65, Tsx, Btk, Syap1, Es1, 
                  Gpd1, Idh1, Mpi, Np, Sod1, Es1C, Gpd1C, Idh1C, NpC, Sod1C,
                  HI_NLoci,  Spleen, Trichuris_muris, Mastophorus_muris, 
                  Catenotaenia_pusilla, Status, 
                  Heterakis_sp, N_oocysts_sq1, 
                  N_oocysts_sq2, N_oocysts_sq3, N_oocysts_sq4, N_oocysts_sq5, 
                  N_oocysts_sq6, N_oocysts_sq7, N_oocysts_sq8, Region, 
                  Body_Length, Fleas, Tail_Length, eimeriaSpecies, Ct.Eimeria,
                  Ct.Mus, ILWE_Crypto_Ct, Aspiculuris_sp, Syphacia_sp, Taenia_sp,
                  Hymenolepis_sp, FEC_Eim_Ct, 
                  all_of(Genes_wild)))

#had to remove as they were disturbing the imputation: Worms_presence, MC.Eimeria.FEC,  Heligmosomoides_polygurus, Zfy2, Y,  MpiC,

#vis_miss(field)


# The frequency distribution of the missing cases per variable can be obtained 
# as:
init <- mice(field_genes, maxit = 0)


#we want to impute only the specific variables
meth <- init$method



```


```{r}
# m=5 refers to the number of imputed datasets. Five is the default value.
igf <- mice(field_genes, m = 5, seed = 500) # method = meth,
summary(igf)

# to check each column with imputed data
## igf$imp$IFNy

#Now we can get back the completed dataset using the complete()
complete_field <- complete(igf, 1)


#sapply(complete_field, function(x) sum(is.na(x)))


# select the required columns
imp_field <- complete_field %>%
  dplyr::select(all_of(Genes_wild))

#visualize missingness
vis_miss(imp_field)

#add an ending to the imputed columns
colnames(imp_field) <- paste(colnames(imp_field), "imp", sep = "_")

#now join it to the full data set of the laboratory infections
field_genes <- cbind(field_genes, imp_field)

# join the now imputed columns to the field
field <- field_genes %>%
  left_join(field, by = intersect(colnames(field), colnames(field_genes)))

```




Let’s compare the distributions of original and imputed data using a some useful 
plots.First of all we can use a scatterplot and plot Ozone against all the other 
variables. Let's first plot the variables for which we have few missing values.



```{r distr_orig_pred}

xyplot(igf, IFNy ~ IL.13 + IRGM1 + MUC2, pch=18,cex=1)


```


What we would like to see is that the shape of the magenta points (imputed) 
matches the shape of the blue ones (observed). The matching shape tells us that 
the imputed values are indeed “plausible values”.

Now let's plot the variables with many missing data points.


```{r distr_orig_pred2}

xyplot(igf,IFNy ~ IL.10 + PRF1 + CASP1, pch=18,cex=1)
```



```{r }
stripplot(igf, pch = 20, cex = 1.2)
```



```{r }
#densityplot(igf)
```

The density of the imputed data for each imputed dataset is showed in magenta 
while the density of the observed data is showed in blue. Again, under our 
previous assumptions we expect the distributions to be similar.

Another useful visual take on the distributions can be obtained using the 
stripplot() function that shows the distributions of the variables as individual 
points


# FACS

```{r }

gf_field <- field %>% 
  dplyr::select(all_of(c(Facs_wild)))

vis_dat(gf_field)

#remove rows with only nas
gf_field <- gf_field[,colSums(is.na(gf_field))<nrow(gf_field)]

#remove colums with only nas 
gf_field <- gf_field[rowSums(is.na(gf_field)) != ncol(gf_field), ]


#select same rows in the first table
field_facs <- field[row.names(gf_field), ]


# really removing empty columns
field_facs <- field_facs %>% 
  discard(~all(is.na(.) | . ==""))

vis_dat(gf_field)
```


# We have no need to impute the facs data, as nothing is missing from the 95 samples 

# Just add here the lab data and join the two tables 


```{r}
lab <- read.csv("output_data/Lab_imputed.csv")

# join the genes and facs data
imputed_mice <- lab %>%
  full_join(field, by = intersect(colnames(lab), colnames(field)))

 ##save the imputed data 
write.csv(imputed_mice, "output_data/imputed_mice.csv", row.names = FALSE)


```

