---
title: "4.4_Mice_imputation_comb.rmd"
author: "Fay"
date: '2022-11-01'
output:
  pdf_document:
    keep_md: yes 
    fig_width: 12
    fig_height: 8

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




# Load libraries



```{r}
library(mice)
library(tidyr)
library(tidyverse)
library(VIM)
library(fitdistrplus)
library(fitur)
library(visdat)
library(DESeq2)
```

# Load data

# Import data

```{r}
hm <- read.csv("output_data/1.MICE_cleaned_data.csv")
```

I only include GAPDH as a housekeeping gene, as PPIB is missing in a large number

```{r}
# Vectors for selecting genes
#Lab genes
# The measurements of IL.12 and IRG6 are done with an other assay and will 
#ignore for now
Gene_lab   <- c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10",
                "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                "TICAM1", "TNF") #"IL.12", "IRG6")

Genes_wild   <- c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10", 
                  "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                  "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                  "TICAM1", "TNF") #, "IL.12", "IRG6")

Facs_lab <- c("CD4", "Treg", "Div_Treg", "Treg17", "Th1", 
                    "Div_Th1", "Th17", "Div_Th17", "CD8", "Act_CD8", 
                    "Div_Act_CD8", "IFNy_CD4", "IFNy_CD8","Treg_prop", 
                    "IL17A_CD4")  

Facs_wild <- c( "Treg", "CD4", "Treg17", "Th1", "Th17", "CD8",
                     "Act_CD8", "IFNy_CD4", "IL17A_CD4", "IFNy_CD8")
```


# data imputation

# Genes
```{r imputing_mice}
hm$Mouse_ID <- str_replace(hm$Mouse_ID, "_", "")

field <- hm %>%
  dplyr::filter(origin == "Field") 

field <- unique(field)
genes_mouse_field <- field %>%
  dplyr::select(c(Mouse_ID, all_of(Genes_wild), )) 
genes <- genes_mouse_field  %>%
  dplyr::select(-Mouse_ID)
#remove rows with only nas
genes <- genes[,colSums(is.na(genes))<nrow(genes)]
#remove colums with only nas 
genes <- genes[rowSums(is.na(genes)) != ncol(genes), ]
genes_mouse_field <- genes_mouse_field[row.names(genes), ]

##select same rows in the first table
field <- field[row.names(genes), ]


###############lab
#select the genes and lab muce
lab <- hm %>%
  dplyr::filter(origin == "Lab", Position == "mLN") #selecting for mln to avoid
# duplicates
lab <- unique(lab)
gene_lab_mouse <- lab %>%
  dplyr::select(c(Mouse_ID, all_of(Gene_lab))) 

gene_lab_mouse <- unique(gene_lab_mouse)

genes_lab <- gene_lab_mouse[, -1]

#remove rows with only nas
genes_lab <- genes_lab[,colSums(is.na(genes_lab))<nrow(genes_lab)]

#remove colums with only nas 
genes_lab <- genes_lab[rowSums(is.na(genes_lab)) != ncol(genes_lab), ]

genes_lab <- unique(genes_lab)

#select same rows in the first table
gene_lab_mouse <- gene_lab_mouse[row.names(genes_lab), ]

##select same rows in the first table
lab <- lab[row.names(genes_lab), ]

hm_genes <- rbind(gene_lab_mouse, genes_mouse_field)

hm_selection_g <- rbind(lab, field)

genes <- cbind(hm_selection_g %>%
                 dplyr::select(c(origin, GAPDH, PPIB)), hm_genes)

genes <- genes %>%
  dplyr::select(-Mouse_ID)
  
genes$origin <- as.factor(genes$origin)

#dplyr::select(-Mouse_ID)
# looking at patterns of nas)
#pattern_na <-as.data.frame(md.pattern(field_genes))
sapply(genes, function(x) sum(is.na(x)))


```


Do we have all the measurements of our housekeeping genes?

```{r}
#glimpse(hm_selection_g)

#dplyr::select(-Mouse_ID)
# looking at patterns of nas)
#pattern_na <-as.data.frame(md.pattern(field_genes))
sapply(field %>%
         dplyr::select(c(all_of(Genes_wild), "PPIB", "GAPDH")), 
                      function(x) sum(is.na(x)))

sapply(lab %>%
         dplyr::select(c(all_of(Gene_lab), "PPIB", "GAPDH")), 
                      function(x) sum(is.na(x)))
         
```

```{r}


#had to remove as they were disturbing the imputation: Worms_presence, MC.Eimeria.FEC,  Heligmosomoides_polygurus, Zfy2, Y,  MpiC,
#vis_miss(field)
# The frequency distribution of the missing cases per variable can be obtained 
# as:
init <- mice(genes, maxit = 0)

#we want to impute only the specific variables
meth <- init$method
```


```{r}
aggr_plot <- aggr(hm_genes, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, labels=names(hm_genes), cex.axis=.7, gap=3, ylab=c("Histogram of missing data","Pattern"))
```


```{r}
marginplot(hm_genes[c(6,8)])
```



```{r}
# removing il 10
genes <- genes %>%
  dplyr::select(-IL.10)
# m=5 refers to the number of imputed datasets. Five is the default value.
igf <- mice(genes, m = 5, seed = 500) # method = meth,

summary(igf)

# to check each column with imputed data
## igf$imp$IFNy
#Now we can get back the completed dataset using the complete()
complete_genes <- complete(igf, 1)

#sapply(complete_field, function(x) sum(is.na(x)))
#visualize missingness
vis_dat(complete_genes)

#remove the non imputed genes from our data set
hm_selection_g <- hm_selection_g %>%
  dplyr::select(-c("IFNy", "CXCR3", "IL.6", "IL.13", #"IL.10",
                "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                "TICAM1", "TNF", "GAPDH", "PPIB", "origin"))
# add the new imputed genes to the data
hm_selection_g <- cbind(hm_selection_g, complete_genes)
```

inspect the trace lines for convergence: 

```{r}
plot(igf)
```



Let’s compare the distributions of original and imputed data using a some useful 
plots.First of all we can use a scatterplot and plot Ozone against all the other 
variables. Let's first plot the variables for which we have few missing values.



```{r distr_orig_pred}
xyplot(igf, IFNy ~ IL.13 + IRGM1 + MUC2, pch=18,cex=1)
```


What we would like to see is that the shape of the magenta points (imputed) 
matches the shape of the blue ones (observed). The matching shape tells us that 
the imputed values are indeed “plausible values”.

Now let's plot the variables with many missing data points.


```{r distr_orig_pred2}
xyplot(igf,IFNy ~ IL.13 + PRF1 + CASP1, pch=18,cex=1)
```



```{r }
stripplot(igf, pch = c(20,21), cex = 1.2)
#bwplot(igf)
```



```{r }
densityplot(igf)
```

The density of the imputed data for each imputed dataset is showed in magenta 
while the density of the observed data is showed in blue. Again, under our 
previous assumptions we expect the distributions to be similar.

Another useful visual take on the distributions can be obtained using the 
stripplot() function that shows the distributions of the variables as individual 
points





# Facs

```{r}
###############lab
#select the facs and lab muce
lab <- hm %>%
  dplyr::filter(origin == "Lab", Position == "mLN") #selecting for mln to avoid

# duplicates
lab <- unique(lab)

facs_mouse <- lab %>%
  dplyr::select(c(Mouse_ID, all_of(Facs_lab))) #choosing the same with the wild

facs_mouse <- unique(facs_mouse)

facs_lab <- facs_mouse[, -1]

#remove rows with only nas
facs_lab <- facs_lab[,colSums(is.na(facs_lab))<nrow(facs_lab)]
#remove colums with only nas 
facs_lab <- facs_lab[rowSums(is.na(facs_lab)) != ncol(facs_lab), ]

#select same rows in the first table
facs_mouse_lab <- facs_mouse[row.names(facs_lab), ]




#########################Field
###########field
# somehow the field samples have the origin na,
# fix that
field <- hm %>%
  dplyr::filter(origin == "Field") 

field <- unique(field)
facs_mouse <- field %>%
  dplyr::select(c(Mouse_ID, all_of(Facs_wild))) 
facs_field <- facs_mouse[,-1]
#remove rows with only nas
facs_field <- facs_field[,colSums(is.na(facs_field))<nrow(facs_field)]
#remove colums with only nas 
facs_field <- facs_field[rowSums(is.na(facs_field)) != ncol(facs_field), ]

#select same rows in the first table
facs_mouse_field <- facs_mouse[row.names(facs_field), ]

# full join the two tables 
facs_data <- full_join(facs_mouse_lab, facs_mouse_field)

length(intersect(hm_selection_g$Mouse_ID, facs_data$Mouse_ID))

facs_data <- facs_data %>%
  left_join(hm) 


```

We don't need to impute anything for the facs data as we have a complete data
set

# join the gene expression data with the facs data
```{r}
setdiff(facs_data$Mouse_ID, hm_selection_g$Mouse_ID)


facs_data <- facs_data %>%
  dplyr::filter(Mouse_ID %in% setdiff(facs_data$Mouse_ID, hm_selection_g$Mouse_ID))


# We expect 477 mice in the new data frame 
472 + 5 

#now combine the two data frames
hm_select <- rbind(hm_selection_g, facs_data)

hm_select <- unique(hm_select)


```

# Gene normalization 



For the field samples the missing values in the house-keeping genes are:
PPIB: 212
GAPDH: 10

For the lab samples the missing values in the house-keeping genes are:
PPIB: 1
GAPDH: 96

Downlad and install the package for Gene normalization

```{r}

#if (!require("BiocManager", quietly = TRUE))
 #   install.packages("BiocManager")

#BiocManager::install("DESeq2")

# create a count data set, rows are sample names and columns the genes
count_data <- hm_select %>%
  dplyr::select(c(Mouse_ID, origin, all_of(Gene_lab), -IL.10 , GAPDH, PPIB))

# select first the field samples 
count_field <- count_data %>%
  dplyr::filter(origin == "Field")

mouse_id <- count_field$Mouse_ID

# row names are samples names
rownames(count_field) <-count_field[,1]

# remove the id column
#count_field <- count_field %>% dplyr::select(-c(origin, Mouse_ID))

count_field <- unique(count_field)

# transpose the data frame to get the sample ids as columns
count_field <- (t(count_field))

# now create the meta data 

meta_field<- hm_select %>%
  dplyr::select(c(Mouse_ID, origin)) %>%
  dplyr::filter(origin == "Field") %>%
  dplyr::rename(id = Mouse_ID) 

#meta_field$id <- str_remove(meta_field$id, "AA_")

# row names are samples names

#rownames(meta_field) <- meta_field[,1]

meta_field$origin <- as.factor(meta_field$origin)

#meta_field <- meta_field %>% 
 # dplyr::select(-id)

condition <- meta_field$origin


all(rownames(meta_field) %in% mouse_id)

all(rownames(meta_field) %in% colnames(count_field))



# construct deseq data set object
dds <- DESeqDataSetFromMatrix(countData= count_field, 
                              colData= meta_data, 
                              design= ~origin, 
                              tidy = TRUE)

head(sort(count_field[,1]))


?DESeqDataSetFromMatrix

norm_field <- DESeq(count_field)

?DESeq

countData <- matrix(1:100,ncol=4)
condition <- factor(c("A","A","B","B"))
?
```
Step 1. Normalize to (REF): ∆Cq = Cq (TAR) – Cq (REF)
Step 2. Exponential expression transform: ∆Cq Expression = 2–∆Cq
Step 3. Average replicates and calculate standard deviation
Step 4. Normalize to treatment control
Step 5. % KD = (1 – ∆∆Cq ) × 100

```{r}
# create a count data set, rows are sample names and columns the genes
count_data <- hm_select %>%
  dplyr::select(c(Mouse_ID, origin, all_of(Gene_lab), -IL.10 , GAPDH, PPIB))

# select first the field samples 
count_field <- count_data %>%
  dplyr::filter(origin == "Field")


## have to remove rows with only nas 

# We are using GAPDH for the field samples

normalize_gene_field <- function(x, y) { # x = target gene, y = housekeeping gene, o = origin
 df <- count_data %>%
  dplyr::filter(origin == "Field")
 
 dct <- df %>% dplyr::select(all_of(x)) - df %>% dplyr::select(all_of(y))
 
 dct <- unlist(as.vector(df %>% dplyr::select(all_of(x))))
 
 dct_mean <- mean(dct, na.rm = TRUE)
 
 #fold gene expression
 norm_gene <- 2^ - (dct - dct_mean)
 norm_gene <- round(norm_gene, digits = 2)

}

x <- "IL.13"
y <- "GAPDH"
 df <- count_data %>%
  dplyr::filter(origin == "Field")

df <- df %>%
  mutate(IL.13_dct = IL.13 - GAPDH)
 
 
 dct_mean <- mean(df$IL.13_dct, na.rm = TRUE)
 
 #fold gene expression
 df <- df %>%
   mutate(IL.13_N = 2^ - (IL.13_dct - dct_mean)) %>%
   mutate(IL.13_N = round(IL.13_N, digits = 2))
 

 
x <- normalize_gene_field("IL.13", "GAPDH") 
  

library(pcr)
# create a count data set, rows are sample names and columns the genes
count_data <- hm_select %>%
  dplyr::select(c(Mouse_ID, origin, all_of(Gene_lab), -IL.10 , GAPDH, PPIB))

count_field <- count_data %>% filter(origin == "Field")

# row names are samples names
rownames(count_field) <-count_field[,1]
group <- count_field$origin
Mouse_ID <- count_field$Mouse_ID
count_field <- count_field[,-(1:2)]

#count_field <- count_field %>% pivot_wider(names_from = o)
#count_field <- as.data.frame(t(count_field))

x <- pcr_analyze(df = count_field %>% dplyr::select(c("IL.13", "GAPDH")), 
            group_var = group,
            reference_gene = "GAPDH",
            method = "delta_delta_ct",
            reference_group = "Field")

?pcr_analyze
  
```


```{r}
# if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

#BiocManager::install("ddCt")
library(ddCt)

?ddCt
data <- SMDFrame()

?SMDFrame
ddCt(object = count_field, housekeepingGenes = "GAPDH" )
```




```{r}
 ##save the imputed data 
write.csv(hm_select, "output_data/2.imputed_MICE_data_set.csv", row.names = FALSE)
```
