---
title: "4.4_Mice_imputation_comb.rmd"
author: "Fay"
date: '2022-11-01'
output:
  pdf_document:
    keep_md: yes 
    fig_width: 12
    fig_height: 8

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




# Load libraries



```{r}
library(mice)
library(tidyr)
library(tidyverse)
library(VIM)
library(fitdistrplus)
library(fitur)
library(visdat)
```

# Load data

# Import data

Here are the cleaned merged data of our mice

```{r}
hm <- read.csv("output_data/1.MICE_cleaned_data.csv")
```


Let's get the vectors ready for selecting genes

```{r}
# Vectors for selecting genes

#Lab genes
# The measurements of IL.12 and IRG6 are done with an other assay and will 
#ignore for now
Gene_lab   <- c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10",
                "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                "TICAM1", "TNF") #"IL.12", "IRG6")

Genes_wild   <- c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10", 
                  "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                  "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                  "TICAM1", "TNF") #, "IL.12", "IRG6")

Facs_lab <- c("CD4", "Treg", "Div_Treg", "Treg17", "Th1", 
                    "Div_Th1", "Th17", "Div_Th17", "CD8", "Act_CD8", 
                    "Div_Act_CD8", "IFNy_CD4", "IFNy_CD8","Treg_prop", 
                    "IL17A_CD4")  

Facs_wild <- c( "Treg", "CD4", "Treg17", "Th1", "Th17", "CD8",
                     "Act_CD8", "IFNy_CD4", "IL17A_CD4", "IFNy_CD8")

```



# data imputation

## Genes
### Field 

Preparing the field gene samples

```{r imputing_mice}
#select the samples that only originate from the field 
field <- hm %>%
  dplyr::filter(origin == "Field") 

# remove the duplicates
field <- unique(field)

#select the genes and the mouse id
genes_mouse_field <- field %>%
  dplyr::select(c(Mouse_ID, origin, all_of(Genes_wild))) 

# now remove the mouse_id to facilitate the next steps
genes_field <- genes_mouse_field  %>%
  dplyr::select(-c(Mouse_ID, origin))


#remove rows with only nas
genes_field <- genes_field[,colSums(is.na(genes_field))<nrow(genes_field)]

#remove colums with only nas 
genes_field <- genes_field[rowSums(is.na(genes_field)) != ncol(genes_field), ]

#add again a variable showing the origin of the samples
genes_field <- genes_field %>%
  mutate(origin = "Field")

#now select from the moouse 
genes_mouse_field <- genes_mouse_field[row.names(genes_field), ]

##select same rows in the first table
field <- field[row.names(genes_field), ]

```

## Lab

Let's repeat for the lab samples


```{r}

###############lab
#select the genes and lab muce
lab <- hm %>%
  dplyr::filter(origin == "Lab", Position == "mLN") #selecting for mln to avoid
# duplicates

lab <- unique(lab)

gene_lab_mouse <- lab %>%
  dplyr::select(c(Mouse_ID, origin, all_of(Gene_lab))) 

gene_lab_mouse <- unique(gene_lab_mouse)

genes_lab <- gene_lab_mouse %>% 
  dplyr::select(-c(Mouse_ID, origin))

#remove rows with only nas
genes_lab <- genes_lab[,colSums(is.na(genes_lab))<nrow(genes_lab)]

#remove colums with only nas 
genes_lab <- genes_lab[rowSums(is.na(genes_lab)) != ncol(genes_lab), ]

genes_lab <- unique(genes_lab)

genes_lab <- genes_lab %>%
  mutate(origin = "Lab")

genes_lab <- unique(genes_lab)

#select same rows in the first table
gene_lab_mouse <- gene_lab_mouse[row.names(genes_lab), ]


##select same rows in the first table
lab <- lab[row.names(genes_lab), ]

```

Combining the two data sets

```{r}
hm_genes <- rbind(genes_field, genes_lab)

hm_selection_g <- rbind(lab, field)



# looking at patterns of nas
#pattern_na <-as.data.frame(md.pattern(field_genes))

sapply(hm_genes, function(x) sum(is.na(x)))



#had to remove as they were disturbing the imputation: Worms_presence, MC.Eimeria.FEC,  Heligmosomoides_polygurus, Zfy2, Y,  MpiC,

#vis_miss(field)



```


```{r}

aggr_plot <- aggr(hm_genes, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, labels=names(hm_genes), cex.axis=.7, gap=3, ylab=c("Histogram of missing data","Pattern"))
```


```{r}
marginplot(hm_genes[c(4,5)])
```
Ready to impute!!!

## Imputation of genes


```{r}

# The frequency distribution of the missing cases per variable can be obtained 
# as:
init <- mice(hm_genes, maxit = 0)

pred_matrix <- init$predictorMatrix


#we want to impute only the specific variables
meth <- init$method



igf$loggedEvents
#why is it removing the origin as a predictor


# removing the colinear 
igf <- mice(hm_genes, m = 5, seed = 500)

igf$loggedEvents
# didnt work
# to check each column with imputed data
## igf$imp$IFNy

#Now we can get back the completed dataset using the complete()
complete_genes <- complete(igf, 1)


#sapply(complete_field, function(x) sum(is.na(x)))


#visualize missingness
vis_dat(complete_genes)


#remove the non imputed genes from our data set
hm_selection_g <- hm_selection_g %>%
  dplyr::select(-c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10",
                "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                "TICAM1", "TNF"))

# add the new imputed genes to the data
hm_selection_g <- cbind(hm_selection_g, complete_genes)

```

inspect the trace lines for convergence: 

```{r}
plot(igf)
```



Let’s compare the distributions of original and imputed data using a some useful 
plots.First of all we can use a scatterplot and plot Ozone against all the other 
variables. Let's first plot the variables for which we have few missing values.



```{r distr_orig_pred}

xyplot(igf, IFNy ~ IL.13 + IRGM1 + MUC2, pch=18,cex=1)


```


What we would like to see is that the shape of the magenta points (imputed) 
matches the shape of the blue ones (observed). The matching shape tells us that 
the imputed values are indeed “plausible values”.

Now let's plot the variables with many missing data points.


```{r distr_orig_pred2}

xyplot(igf,IFNy ~ IL.10 + PRF1 + CASP1, pch=18,cex=1)
```



```{r }
stripplot(igf[, 1:20], pch = 20, cex = 1.2)

```



```{r }

densityplot(igf, height = 1000, width = 800)
```

The density of the imputed data for each imputed dataset is showed in magenta 
while the density of the observed data is showed in blue. Again, under our 
previous assumptions we expect the distributions to be similar.

Another useful visual take on the distributions can be obtained using the 
stripplot() function that shows the distributions of the variables as individual 
points





# Facs

```{r}
###############lab
#select the facs and lab muce
lab <- hm %>%
  dplyr::filter(origin == "Lab", Position == "mLN") #selecting for mln to avoid
# duplicates

lab <- unique(lab)

facs_mouse <- lab %>%
  dplyr::select(c(Mouse_ID, all_of(Facs_lab))) #choosing the same with the wild

facs_mouse <- unique(facs_mouse)

facs_lab <- facs_mouse[, -1]

#remove rows with only nas
facs_lab <- facs_lab[,colSums(is.na(facs_lab))<nrow(facs_lab)]

#remove colums with only nas 
facs_lab <- facs_lab[rowSums(is.na(facs_lab)) != ncol(facs_lab), ]


#select same rows in the first table
facs_mouse <- facs_mouse[row.names(facs_lab), ]

##select same rows in the first table
lab <- lab[row.names(facs_mouse), ]

#########################Field
###########field
# somehow the field samples have the origin na,
# fix that

field <- hm %>%
  dplyr::filter(origin == "Field") 

field <- unique(field)


facs_mouse <- field %>%
  dplyr::select(c(Mouse_ID, all_of(Facs_wild))) 

facs_field <- facs_mouse[,-1]

#remove rows with only nas
facs_field <- facs_field[,colSums(is.na(facs_field))<nrow(facs_field)]

#remove colums with only nas 
facs_field <- facs_field[rowSums(is.na(facs_field)) != ncol(facs_field), ]


##select same rows in the first table
field <- field[row.names(facs_field), ]

facs_data <- full_join(lab, field, by = intersect(colnames(lab), colnames(field)))

facs_data <- unique(facs_data) %>%
  dplyr::select(-c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10",
                "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                "TICAM1", "TNF"))
```

# Now put everything togetehr


```{r}
hm_selection_g <- hm_selection_g %>% 
  full_join(facs_data, by = intersect(colnames(facs_data), colnames(hm_selection_g)))

hm_selection_g <- unique(hm_selection_g)

```


```{r}



 ##save the imputed data 
write.csv(hm_selection_g, "output_data/2.imputed_MICE_data_set.csv", row.names = FALSE)


```

