---
title: "6.  PCA FACS -lab"
author: "Fay"
date: '2022-11-03'
output:
  pdf_document:
    keep_md: yes 
    fig_width: 12
    fig_height: 8
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Always change the knitting directory to the working directory!
# Load libraries

```{r message = FALSE, warnings = FALSE}
library(tidyverse)
library(dplyr)
library(stringr)
library(FactoMineR)
library(reshape2)
library(corrplot)
library(factoextra)
library(lmtest)
library(ggpubr)
library(janitor)
library(pheatmap)
library(visdat)
```


# Load data

```{r}
hm <- read.csv("output_data/2.imputed_MICE_data_set.csv")
```


# vectors for selecting

```{r}
Gene_lab   <- c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10",
                "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                "TICAM1", "TNF") # "IL.12", "IRG6")

#add a suffix to represent changes in data file
Gene_lab_imp <- paste(Gene_lab, "imp", sep = "_")

Genes_wild   <- c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10", 
                  "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                  "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                  "TICAM1", "TNF", "IL.12", "IRG6")

Genes_wild_imp <- paste(Genes_wild, "imp", sep = "_")

Facs_lab <- c("CD4", "Treg", "Div_Treg", "Treg17", "Th1", 
                    "Div_Th1", "Th17", "Div_Th17", "CD8", "Act_CD8", 
                    "Div_Act_CD8", "IFNy_CD4", "IFNy_CD8") #"Treg_prop", removed due to many missing values
                    #"IL17A_CD4"


Facs_wild <- c( "Treg", "CD4", "Treg17", "Th1", "Th17", "CD8",
                     "Act_CD8", "IFNy_CD4", "IL17A_CD4", "IFNy_CD8")


```


# FACS
## Lab

# PCA on the lab genes -*imputed*
```{r}
#select the genes and lab muce
lab <- hm %>%
  dplyr::filter(origin == "Lab", Position == "mLN") #selecting for mln to avoid

# duplicates

lab <- unique(lab)

facs_mouse <- lab %>%
  dplyr::select(c(Mouse_ID, all_of(Facs_lab)))


facs <- facs_mouse[, -1]

#remove rows with only nas
facs <- facs[,colSums(is.na(facs))<nrow(facs)]

#remove colums with only nas 
facs <- facs[rowSums(is.na(facs)) != ncol(facs), ]

vis_dat(facs)

#select same rows in the first table
facs_mouse <- facs_mouse[row.names(facs), ]


# we can now run a normal pca on the complete data set
res.pca <- PCA(facs)
```
```{r , echo = FALSE}

#The most important (or, contributing) variables can be highlighted on the 
#correlation plot as follow:
fviz_pca_var(res.pca, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")
             )
```
```{r}
fviz_pca_ind(res.pca, col.ind = "cos2", 
                  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
                  repel = TRUE)
```


### Dimensions of the pca


```{r dimensions, include = FALSE, echo = FALSE, warnings = FALSE}
#Description of the dimensions
# We get a correlation between each variable and the first dimension
dimdesc(res.pca)
```

Caution: When imputing data, the percentages of inertia associated with the first dimensions will be overestimated.

Another problem: the imputed data are, when the pca is performed considered like real observations.
But they are estimations!! 

Visualizing uncertainty due to issing data:

--> mulrimple imputation: generate several plausible values for each missing data point

We here visualize the variability, that is uncertainty on the plane defined by two pca axes. 

```{r, echo = FALSE, include = FALSE}
str(res.pca)
```



```{r, echo = FALSE}

# take the mouse column to add the pc1 and pc2
mouse_id <- as.data.frame(facs_mouse$Mouse_ID)

mouse_id <- mouse_id %>%
  rename(Mouse_ID = "facs_mouse$Mouse_ID")

mouse_id$pc1 <- res.pca$ind$coord[, 1] # indexing the first column

mouse_id$pc2 <- res.pca$ind$coord[, 2]  # indexing the second column

# join it to the lab data

lab <- lab %>% 
  left_join(mouse_id, by = "Mouse_ID")


#We also need to extract the data for the variable contributions to each of the pc axes.
pca.vars <- res.pca$var$coord %>% data.frame


pca.vars$vars <- rownames(pca.vars)

pca.vars.m <- melt(pca.vars, id.vars = "vars")

source("r_scripts/functions/circle_fun.R")

circ <- circleFun(c(0,0),2,npoints = 500)

```


Biplot of the imputed facs pca


```{r biplot_pca_facss}

#Now we can make our initial plot of the PCA.
lab %>%
  ggplot(aes(x = pc1, y = pc2, 
             color = Parasite_challenge, 
             shape = Parasite_challenge)) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_vline(xintercept = 0, lty = 2) +
  geom_point(alpha = 0.8) +
  stat_ellipse(geom="polygon", 
               aes(fill = challenge_infection), 
               alpha = 0.2, show.legend = FALSE,
               level = 0.95) +
  theme_minimal() +
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(fill= "transparent")) 
  
```
```{r correlations_genes_dimensions, echo = FALSE}
#It’s possible to use the function corrplot() [corrplot package] to highlight 
#the most contributing variables for each dimension:
var.contrib <- res.pca$var$contrib
corrplot(var.contrib, is.corr=FALSE) 
```

The function fviz_contrib() [factoextra package] can be used to draw a bar plot of variable contributions. If your data contains many variables, you can decide to show only the top contributing variables. The R code below shows the top 10 variables contributing to the principal components:


```{r contr_var_pc_genes, echo = FALSE}
# Contributions of variables to PC1
fviz_contrib(res.pca, choice = "var", axes = 1, top = 18)

```


```{r}
# Contributions of variables to PC2
fviz_contrib(res.pca, choice = "var", axes = 2, top = 18)
```


```{r contr_var_pc1_2_genes, echo = FALSE}
fviz_contrib(res.pca, choice = "var", axes = 1:2, top = 18)
```
The red dashed line on the graph above indicates the expected average contribution. If the contribution of the variables were uniform, the expected value would be 1/length(variables) = 1/10 = 10%. For a given component, a variable with a contribution larger than this cutoff could be considered as important in contributing to the component.

Note that, the total contribution of a given variable, on explaining the variations retained by two principal components, say PC1 and PC2, is calculated as contrib = [(C1 * Eig1) + (C2 * Eig2)]/(Eig1 + Eig2), where

C1 and C2 are the contributions of the variable on PC1 and PC2, respectively
Eig1 and Eig2 are the eigenvalues of PC1 and PC2, respectively. Recall that eigenvalues measure the amount of variation retained by each PC.
In this case, the expected average contribution (cutoff) is calculated as follow: As mentioned above, if the contributions of the 10 variables were uniform, the expected average contribution on a given PC would be 1/10 = 10%. The expected average contribution of a variable for PC1 and PC2 is : [(10* Eig1) + (10 * Eig2)]/(Eig1 + Eig2)



```{r pca_contribution_genes, echo = FALSE}

#The most important (or, contributing) variables can be highlighted on the 
#correlation plot as follow:
fviz_pca_var(res.pca, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")
             )
```



To visualize the contribution of individuals to the first two principal components:

```{r contr_individuals_genes, echo = FALSE}
# Total contribution on PC1 and PC2
fviz_contrib(res.pca, choice = "ind", axes = 1:2)
```

PCA + Biplot combination

```{r pca_biplot_genes, echo = FALSE}
#select same rows in the first table
lab <- lab[row.names(facs), ]

fviz_pca_biplot(res.pca, 
                col.ind = lab$Parasite_challenge, palette = "jco", 
                addEllipses = TRUE, label = "var",
                col.var = "black", repel = TRUE,
                legend.title = "Parasite in the challenge infection") 

```
 In the following example, we want to color both individuals and variables by groups. The trick is to use pointshape = 21 for individual points. This particular point shape can be filled by a color using the argument fill.ind. The border line color of individual points is set to “black” using col.ind. To color variable by groups, the argument col.var will be used.





################## Linear models: 

```{r lm_weight_pc, echo = FALSE}
weight_lm <- lm(max_WL ~ pc1 + pc2 , data = lab)
summary(weight_lm)
AIC(weight_lm)
```




```{r lm_pc_parasite_hybrid, echo = FALSE}

weight_lm2 <- lm(max_WL ~ pc1 + pc2  + hybrid_status, 
                 data = lab)
summary(weight_lm2)
AIC(weight_lm2)
```

```{r}
weight_lm3 <- lm(max_WL ~ pc1  + hybrid_status, data = lab)
summary(weight_lm3)
AIC(weight_lm3)
```

Try instead: 
LLR test (likelihood ration)  (LM4 package )? 

https://www.rdocumentation.org/packages/lmtest/versions/0.9-38/topics/lrtest

In this way you compare each model, with the different variables usesd to predict. 

Another way is to compare the AIC. (function : step)

```{r}


weight_lm3 <- lm(max_WL ~ pc1 + pc2 + hybrid_status, data = lab)
weight_no_pc1 <- lm(max_WL ~ pc2 + hybrid_status, data = lab)
weight_no_pc2 <- lm(max_WL ~ pc1  + hybrid_status, data = lab)
weight_no_hybrid <- lm(max_WL ~ pc1 + pc2, data = lab)
lrtest(weight_lm2, weight_lm3)
lrtest(weight_lm, weight_lm3)
lrtest(weight_lm3, weight_no_hybrid)
```


```{r lm_pc_hybrid, echo = FALSE}
weight_lm3 <- lm(max_WL ~ pc1 + pc2 + hybrid_status, data = lab)
summary(weight_lm3)
AIC(weight_lm3)
```



```{r lm_inf_histoy, echo=FALSE}
weight_lm4 <- lm(max_WL ~ pc1 + pc2 + infection_history, data = lab)
summary(weight_lm4)
AIC(weight_lm4)

```


```{r lm_pc, echo = FALSE}
weight_lm_exp_only <- lm(max_WL ~ pc1 + pc2, data = lab)
summary(weight_lm_exp_only)
AIC(weight_lm, weight_lm_exp_only)
```


```{r}
weight_lm5 <- lm(max_WL ~ hybrid_status, data = lab)
summary(weight_lm5)
AIC(weight_lm5)
```




### repeating the heatmap on the now imputed data
```{r}
 # turn the data frame into a matrix and transpose it. We want to have each cell 
 # type as a row name 
 facs_mouse <- t(as.matrix(facs_mouse))
 
 # turn the first row into column names
 facs_mouse %>%
     row_to_names(row_number = 1) -> heatmap_data
 
 heatmap_data <- as.data.frame(heatmap_data)
 
 table(rowSums(is.na(heatmap_data)) == nrow(heatmap_data))

 
# turn the columns to numeric other wise the heatmap function will not work
 heatmap_data[] <- lapply(heatmap_data, function(x) as.numeric(as.character(x)))

 # remove columns with only NAs 
 heatmap_data <- Filter(function(x)!all(is.na(x)), heatmap_data) 
 
 #remove rows with only Nas
 heatmap_data <-  heatmap_data[, colSums(is.na(heatmap_data)) != 
                                   nrow(heatmap_data)]
 
  
#Prepare the annotation data frame
annotation_df <- as_tibble(lab) %>%
    dplyr::select(c("Mouse_ID", "Parasite_challenge", "infection_history",
                    "mouse_strain", "max_WL")) 
  
annotation_df <- unique(annotation_df) 

annotation_df <- as.data.frame(annotation_df)




### Prepare the annotation columns for the heatmap
rownames(annotation_df) <- annotation_df$Mouse_ID


# Match the row names to the heatmap data frame
rownames(annotation_df) <- colnames(heatmap_data)

#remove the unecessary column
annotation_df <- annotation_df %>% dplyr::select(-Mouse_ID, )


```


Heatmap on facs expression data: 


```{r, echo = FALSE}
pheatmap(heatmap_data, annotation_col = annotation_df, scale = "row")
```