---
title: "facs_gene_analysis"
author: "Fay"
date: '2022-05-19'
output:
  pdf_document: 
    fig_width: 12
    fig_height: 8
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

################## FACS - genes Imputation and pca ############################


```{r, echo=FALSE, include = FALSE}
library(pheatmap)
library(tidyverse)
library(matrixStats)
library(tidyr)
library(janitor)
library(tibble)
library(corrplot)
library(RColorBrewer)
library(ggplot2)
library(dplyr)
library(missMDA) 
library(FactoMineR)
library(PerformanceAnalytics)
library(factoextra)
library(MASS)
library(reshape2)
library(cowplot)
```

Load the data sets
```{r loading_data_set}
facs <- read.csv("output_data/facs/data_products/FACS_clean.csv")
gene <- read.csv("output_data/gene_expression/data_products/clean_gene_expression.csv")
```


# 1. Start by combining the data sets 
```{r merging_data}
## Adding prefixes to the columns of each data frame and joining

#Adding the suffix G to the genes
colnames(gene) <- paste("G_", colnames(gene), sep = "")

gene <- gene %>% rename(EH_ID = G_EH_ID,
                                  primary_infection = G_primary_infection,
                                  challenge_infection = G_challenge_infection,
                                  infection_history = G_infection_history,
                                  mouse_strain = G_mouse_strain,
                                  max_WL = G_max_WL,
                                  delta = G_delta, 
                                  Parasite_challenge =  G_Parasite_challenge,
                                  hybrid_status = G_hybrid_status)

#Adding the suffix f to the facs data
colnames(facs) <- paste("F_", colnames(facs), sep = "")

facs <- facs %>% rename(EH_ID = F_EH_ID,
                        infection_history = F_infection_history,
                        max_WL = F_max_WL,
                        Parasite_challenge =  F_Parasite_challenge,
                        hybrid_status = F_hybrid_status,
                        delta = F_delta)

immune_data <- gene %>% full_join(facs, by = intersect(colnames(gene), colnames(facs)))

immune_data <- unique(immune_data)
```


### Now we go on to see the correlations between our data 


```{r correlations_immune_data_prep, echo = FALSE}
# draw correlations
immune <- immune_data  %>% 
  dplyr::select(-c(EH_ID, primary_infection, challenge_infection, infection_history, 
                   mouse_strain, max_WL, delta, Parasite_challenge, hybrid_status, 
                   F_Position, F_Treg_prop, F_IL17A_CD4))
                
immune_correlation <-  as.data.frame(cor(immune, use="pairwise.complete.obs"))

# matrix of the p-value of the correlatio
p.mat <- cor.mtest(immune_correlation)


immune_correlation <- as.matrix(immune_correlation)

# I am removing F_Treg_prop, F_IL17A_CD4 from the corrplot as it causes problems with plotting 
# reason: too many nas in IL13 and F_Treg_prop, F_IL17A_CD4

# load the function to calculate the p value for correlations
source("r_scripts/functions/p_value_for_correlations.R")


# matrix of the p-value of the correlatio
p.mat <- cor.mtest(immune_correlation)
```

```{r immune_correlation}
corrplot(immune_correlation, 
         method = "circle",  #method of the plot, "color" would show colour gradient
         tl.col = "black", tl.srt = 45, #colour of labels and rotation
         col = brewer.pal(n = 8, name ="RdYlBu"), #colour of matrix
         order="hclust", #hclust reordering
         p.mat = p.mat, sig.level = 0.01, insig = "blank",
         addCoef.col = 'black',
         number.cex = 0.5)

```







We will now continue by using an iterative pca to impute missing data 
 A. Initialization: impute using the mean
 B. Step lampda:
  # a. do pca on imputed data table S dimensions retained
  # b. missing data imputed using pca
  # c. means (and standard deviations) updated
 C. Iterate the estimation and imputation steps (until convergence)
(convergence: the act of converging and especially moving toward union or uniformity)

Overfitting is a common problem due to believing too much in links between variables. 
--> regularized iterative PCA (This version is what is being implented in missMDA)
This is a way of taking less risk when imputing the missing data. 
The algorithm estimates the missing data values with values that have no influence
on the PCA results, i.e., no influence on the coordinates of the individals or variables. 

```{r pca_imputation_immune, echo = FALSE}
## 1. Estimate the number of dimensions necessary to impute the dataset 
# with the estim_ncpPCA
nb <- estim_ncpPCA(immune, scale = TRUE) 
# Each observed value is removed, then imputed using ncp components
# The difference between the ovserved value and the imputed value is calculated 
# ncp = 2

## 2. Imputation of the missing values using the function impute PCA
comp <- imputePCA(immune, ncp = 2, scale = TRUE) #impute the table
#head(comp$completeObs) # resulting imputed data table
# this estimate should be better than simply using the mean, beacause we've taken into account
# the links between variables and similarities between individuals in its calculation
imputed_immune <- as.data.frame(comp$completeObs)

#options(ggrepel.max.overlaps = Inf) # solution to the error: 
# ggrepel: 3 unlabeled data points (too many overlaps). Consider increasing max.overlaps 
# for the next function

# we can now run a normal pca on the complete data set
res.pca <- PCA(comp$completeObs)

#head(summary(res.pca, nbelements = Inf)) #including all the elements
```
```{r dimensions, echo = FALSE, include = FALSE, message=FALSE, warning = FALSE}
#Description of the dimensions
# We get a correlation between each variable and the first dimension
dimdesc(res.pca)
```


Caution: When imputing data, the percentages of inertia associated with the first dimensions will be overestimated.

Another problem: the imputed data are, when the pca is performed considered like real observations.
But they are estimations!! 

Visualizing uncertainty due to issing data:

--> mulrimple imputation: generate several plausible values for each missing data point

We here visualize the variability, that is uncertainty on the plane defined by two pca axes.



```{r visu_uncertainty_immune, echo = FALSE}
#Visualizing uncertainty due to missing data:

mi <- MIPCA(immune, scale = TRUE, ncp = 5)
plot(mi)
```




Individuals lying on the axis have no missing data, but individuals that far away have many missing data. 
big ellipse = big uncertainty
tight elipse (line) = low uncertainty

Variable representation: 
Poins tight together )look like one) - have no missing variables --> low uncertainty
Points spread -- > higher variability -- > higher uncertainty


High uncertainty--> we should interpret the result with care

The individuals with many missing data values make the axes move, 
and thus the positions of all individuals

Therefore in the last plots every individual is getting an eclipse as they are as well influenced by the missing data of the others. 


THe plot with the dimensions shows the projections of the pca dimensions of each imputed table on the pca plane obtained using the original imputed data table

As all of the arrows are close to either the first or second axes, 
this means that the axes are stable with respect to the set of imputed tables --> we don't have evidence of instability here.

```{r structure_res.pca, echo = FALSE, include = FALSE}
str(res.pca)
```



```{r merging_immune_data, echo = FALSE}
# extract pc scores for first two component and add to dat dataframe
immune_data$pc1 <- res.pca$ind$coord[, 1] # indexing the first column

immune_data$pc2 <- res.pca$ind$coord[, 2]  # indexing the second column

imputed_immune$pc1 <- res.pca$ind$coord[, 1]
imputed_immune$pc2 <- res.pca$ind$coord[, 2] 

#We also need to extract the data for the variable contributions to each of the pc axes.
pca.vars <- res.pca$var$coord %>% data.frame
pca.vars$vars <- rownames(pca.vars)
pca.vars.m <- melt(pca.vars, id.vars = "vars")

#source the function required
source("r_scripts/functions/circle_fun.R")

circ <- circleFun(c(0,0),2,npoints = 500)

#Joining the imputed data to the information columns

### join the imputed data 
imputed_immune <- immune_data %>% left_join(imputed_immune, by = c("pc1", "pc2"), 
                                suffix = c("_x", "_y"))

#remove all columns of the non-imputed data
imputed_immune = imputed_immune[,!grepl("_x$",names(imputed_immune))]

#remove the suffix y
colnames(imputed_immune) = gsub("_y", "", colnames(imputed_immune))

```

```{r biplot_pca_immune}

imputed_immune %>%
  pivot_longer(cols = 13:47, names_to = "Immune_parameter", values_to = "measurement")  %>%
  ggplot(aes(x = pc1, y = pc2, color = Parasite_challenge, shape = Parasite_challenge)) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_vline(xintercept = 0, lty = 2) +
  geom_point(alpha = 0.8) +
  stat_ellipse(geom="polygon", aes(fill = Parasite_challenge), alpha = 0.2, show.legend = FALSE,
               level = 0.95) +
  theme_minimal() +
  theme(panel.grid = element_blank(), panel.border = element_rect(fill= "transparent"))

```


```{r contr_pc1, echo = FALSE}
# Contributions of variables to PC1
fviz_contrib(res.pca, choice = "var", axes = 1, top = 18)

```


```{r contr_pc2}
# Contributions of variables to PC2
fviz_contrib(res.pca, choice = "var", axes = 2, top = 18)
```



The total contribution to PC1 and PC2 is obtained with the following R code:
```{r contr_pc1_pc2, echo = FALSE}
fviz_contrib(res.pca, choice = "var", axes = 1:2, top = 18)
```




################## Linear models: 

```{r lm_weight_pc, echo = FALSE}
weight_lm <- lm(max_WL ~ pc1 + pc2 + Parasite_challenge, data = imputed_immune)
summary(weight_lm)
AIC(weight_lm)
```




```{r lm_pc_parasite_hybrid, echo = FALSE}
weight_lm2 <- lm(max_WL ~ pc1 + pc2 + Parasite_challenge + hybrid_status, data = imputed_immune)
summary(weight_lm2)
AIC(weight_lm2)
```



```{r lm_pc_hybrid, echo = FALSE}
weight_lm3 <- lm(max_WL ~ pc1 + pc2 + hybrid_status, data = imputed_immune)
summary(weight_lm3)
AIC(weight_lm3)
```



```{r lm_inf_histoy, echo=FALSE}
weight_lm4 <- lm(max_WL ~ pc1 + pc2 + infection_history, data = imputed_immune)
summary(weight_lm4)
AIC(weight_lm4)
```


```{r lm_pc, echo = FALSE}
weight_lm_exp_only <- lm(max_WL ~ pc1 + pc2, data = imputed_immune)
summary(weight_lm_exp_only)
AIC(weight_lm, weight_lm_exp_only)
```




### repeating the heatmap on the now imputed data



```{r, echo = FALSE}


i <- imputed_immune %>% dplyr::select(c(3, 13:47))

# turn the data frame into a matrix and transpose it. We want to have each cell 
# type as a row name 
i <- t(as.matrix(i))

#switch the matrix back to a data frame format
i <- as.data.frame(i)

# turn the first row into column names
i %>%
  row_to_names(row_number = 1) -> heatmap_data


# turn the columns to numeric other wise the heatmap function will not work
heatmap_data[] <- lapply(heatmap_data, function(x) as.numeric(as.character(x)))

annotation_df <- imputed_immune %>%
  dplyr::select(c("EH_ID", "Parasite_challenge", "infection_history",  
                  "hybrid_status", "max_WL"))

### Prepare the annotation columns for the heatmap
rownames(annotation_df) <- annotation_df$EH_ID

# Match the row names to the heatmap data frame
rownames(annotation_df) <- colnames(i)

#remove the unecessary column
annotation_df <- annotation_df %>% dplyr::select(-EH_ID, )

```




### Heatmap on imputed combined data:

```{r heatmap_imputed_facs_genes}
#plot the heatmap

heatmap_data %>% 
  pheatmap(annotation_col = annotation_df, scale = "row")
```

