---
title: "TAC_report_2022"
author: "Fay"
date: '2022-12-12'
bibliography: mybibfile.bib
output:
  pdf_document:
    keep_md: yes 
    fig_width: 12
    fig_height: 8
  output:
  html_document:
    df_print: paged
    fig_width: 12
    fig_height: 8
  word_document: default
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, # By default, hide code; set to TRUE to see code
  fig.pos = 'th', # Places figures at top or here
  out.width = '100%', dpi = 300, # Figure resolution and size
  fig.env="figure"
) # Latex figure environment

```

```{r load libraries, message = FALSE, include=FALSE, warning=FALSE, echo=FALSE}
library(mice)
library(tidyr)
library(tidyverse)
library(VIM)
library(fitdistrplus)
library(fitur)
library(visdat)
library(corrplot)
library(dplyr)
library(randomForest)
library(ggplot2)
library(caret)
library(ggpubr)
library(rfUtilities) # Implements a permutation test cross-validation for 
# Random Forests models

#install.packages("optimx", version = "2021-10.12") # this package is required for 
#the parasite load package to work


```

# Abstract 

Parasites in hybrid zones can give insight into species barriers, as they are modulating the fitness of hybrid hosts. Recent findings have demonstrated lower infection intensities with parasites in hybrids in the European House Mouse Hybrid zone (HMHZ), indicating higher disease resistance. However, tolerance has not yet been addressed in depth, as it is impractical to measure in wild populations. In an attempt to predict and evaluate the health impact of parasite infections and extrapolate tolerance in the HMHZ, we use a machine learning method. A random forest model was trained on immune parameters measured in experimental lab infections with Eimeria and then applied to data obtained from field sampling. Our predictions revealed that these infections are more detrimental to hybrid male mice. This approach represents an initial step in assessing tolerance in field studies.

# Introduction

# Methods




## Laboratory infection experiments


## Wild mice 

During the years 2016 to 2019, we sampled 1889 mice in the House Mouse Hybrid zone. We used live traps to capture mice in farms and houses, during September every year. To catch mice of *M. musculus musculus*, *M. musculus domesticus* and hybrid origin, we selected a large geographic area in Brandenburg and Neuruppin.  For each mouse, we have collected information during dissections, on their body length, their weight and their pathophysiology. Tissue samples were used to genotype the hosts. Further information on the set-up of the experiment can be found in previous publications of the group @balard2020intensity From 1889 mice, 336 were selected with a varying genotype for Fluidigm BioMark assay. Another 95 were selected for fluorescence-activated cell sorting. 

## Statistical Analysis

### Imputation of missing data 

```{r MICE_imputation, warning=FALSE, message=FALSE, include=FALSE}

setwd("~/GitHub/Eimeria_mouse_immunity/")

hm <- read.csv("output_data/1.MICE_cleaned_data.csv")

# Vectors for selecting genes
#Lab genes
# The measurements of IL.12 and IRG6 are done with an other assay and will 
#ignore for now
Gene_lab   <- c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10",
                "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                "TICAM1", "TNF") #"IL.12", "IRG6")
Genes_wild   <- c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10", 
                  "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                  "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                  "TICAM1", "TNF") #, "IL.12", "IRG6")
Facs_lab <- c("CD4", "Treg", "Div_Treg", "Treg17", "Th1", 
                    "Div_Th1", "Th17", "Div_Th17", "CD8", "Act_CD8", 
                    "Div_Act_CD8", "IFNy_CD4", "IFNy_CD8","Treg_prop", 
                    "IL17A_CD4")  
Facs_wild <- c( "Treg", "CD4", "Treg17", "Th1", "Th17", "CD8")

# data imputation

# Genes
field <- hm %>%
  dplyr::filter(origin == "Field") 

field <- unique(field)

genes_mouse_field <- field %>%
  dplyr::select(c(Mouse_ID, all_of(Genes_wild)))

genes <- genes_mouse_field  %>%
  dplyr::select(-Mouse_ID)

#remove rows with only nas
genes <- genes[,colSums(is.na(genes))<nrow(genes)]

#remove colums with only nas 
genes <- genes[rowSums(is.na(genes)) != ncol(genes), ]

genes_mouse_field <- genes_mouse_field[row.names(genes), ]

##select same rows in the first table
field <- field[row.names(genes), ]

###############lab
#select the genes and lab muce
lab <- hm %>%
  dplyr::filter(origin == "Lab", Position == "mLN") #selecting for mln to avoid
# duplicates

lab <- unique(lab)

gene_lab_mouse <- lab %>%
  dplyr::select(c(Mouse_ID, all_of(Gene_lab))) 

gene_lab_mouse <- unique(gene_lab_mouse)

genes_lab <- gene_lab_mouse[, -1]

#remove rows with only nas
genes_lab <- genes_lab[,colSums(is.na(genes_lab))<nrow(genes_lab)]

#remove colums with only nas 
genes_lab <- genes_lab[rowSums(is.na(genes_lab)) != ncol(genes_lab), ]

genes_lab <- unique(genes_lab)

#select same rows in the first table
gene_lab_mouse <- gene_lab_mouse[row.names(genes_lab), ]

##select same rows in the first table
lab <- lab[row.names(genes_lab), ]

hm_genes <- rbind(gene_lab_mouse, genes_mouse_field)

hm_selection_g <- rbind(lab, field)

genes <- hm_genes %>%
 dplyr::select(-Mouse_ID)

# looking at patterns of nas
#pattern_na <-as.data.frame(md.pattern(field_genes))
sapply(hm_genes, function(x) sum(is.na(x)))
#had to remove as they were disturbing the imputation: Worms_presence, MC.Eimeria.FEC,  Heligmosomoides_polygurus, Zfy2, Y,  MpiC,
#vis_miss(field)
# The frequency distribution of the missing cases per variable can be obtained 
# as:
init <- mice(genes, maxit = 0)
#we want to impute only the specific variables
meth <- init$method

aggr_plot <- aggr(hm_genes, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, labels=names(hm_genes), cex.axis=.7, gap=3, ylab=c("Histogram of missing data","Pattern"))

marginplot(hm_genes[c(4,5)])

# m=5 refers to the number of imputed datasets. Five is the default value.
igf <- mice(genes, m = 5, seed = 500) # method = meth,
summary(igf)
# to check each column with imputed data
## igf$imp$IFNy
#Now we can get back the completed dataset using the complete()
complete_genes <- complete(igf, 1)
#sapply(complete_field, function(x) sum(is.na(x)))
#visualize missingness
vis_dat(complete_genes)
#remove the non imputed genes from our data set
hm_selection_g <- hm_selection_g %>%
  dplyr::select(-c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10",
                "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                "TICAM1", "TNF"))
# add the new imputed genes to the data
hm_selection_g <- cbind(hm_selection_g, complete_genes)

xyplot(igf, IFNy ~ IL.13 + IRGM1 + MUC2, pch=18,cex=1)

xyplot(igf,IFNy ~ IL.10 + PRF1 + CASP1, pch=18,cex=1)

###############lab
#select the facs and lab muce
lab <- hm %>%
  dplyr::filter(origin == "Lab", Position == "mLN") #selecting for mln to avoid
# duplicates
lab <- unique(lab)
facs_mouse <- lab %>%
  dplyr::select(c(Mouse_ID, all_of(Facs_lab))) #choosing the same with the wild
facs_mouse <- unique(facs_mouse)
facs_lab <- facs_mouse[, -1]
#remove rows with only nas
facs_lab <- facs_lab[,colSums(is.na(facs_lab))<nrow(facs_lab)]
#remove colums with only nas 
facs_lab <- facs_lab[rowSums(is.na(facs_lab)) != ncol(facs_lab), ]
#select same rows in the first table
facs_mouse <- facs_mouse[row.names(facs_lab), ]
##select same rows in the first table
lab <- lab[row.names(facs_mouse), ]
#########################Field
###########field
# somehow the field samples have the origin na,
# fix that
field <- hm %>%
  dplyr::filter(origin == "Field") 

field <- unique(field)

facs_mouse <- field %>%
  dplyr::select(c(Mouse_ID, all_of(Facs_wild))) 

facs_field <- facs_mouse[,-1]

#remove rows with only nas
facs_field <- facs_field[,colSums(is.na(facs_field))<nrow(facs_field)]

#remove colums with only nas 
facs_field <- facs_field[rowSums(is.na(facs_field)) != ncol(facs_field), ]

##select same rows in the first table
field <- field[row.names(facs_field), ]

facs_data <- full_join(lab, field, by = intersect(colnames(lab), colnames(field)))

facs_data <- unique(facs_data) %>%
  dplyr::select(-c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10",
                "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                "TICAM1", "TNF"))

hm_selection_g <- hm_selection_g %>% 
  full_join(facs_data, by = intersect(colnames(facs_data), colnames(hm_selection_g)))

hm_selection_g <- unique(hm_selection_g)

hm <- hm_selection_g

rm(hm_selection_g)
```

To make the most of our data collection, we aimed to resolve missingness. Missing data were imputed using multiple imputations by chained equations. We used the package MICE in R @van2011mice, with five imputed data sets and five iterations. Data generated by FACS or the Gene Expression / Biomarker assay were regarded as missing if each mouse had measurements for some variables. For each continuous variable, we specified a predictive mean matching model. All the remaining variables were used as predictors in the imputation. To control the quality of our imputations, we evaluated the distribution plot of the existing data and the imputed data for all measurements \ref{fig:fig1} \ref{fig:fig2}. Further, we tested for convergence. We assume data is "missing completely at random" or "missing at random". For both types of missingness, multiple imputation is a suggested method to impute missing variables @van2018flexible. 


```{r fig1, fig.cap="Stripplot of observed and imputed data", fig.env="figure*"}
stripplot(igf, pch = 20, cex = 1.2)
```



```{r fig2, fig.cap="Density plot of observed and imputed data",fig.env="figure*"}
densityplot(igf, height = 1000, width = 800)
```

```{r fig3, fig.cap="Inspecting to evaluate trace lines for convergence", fig.env="figure*", echo=FALSE, message = FALSE, include = FALSE}
plot(igf)
```


#### Questions / To-dos:
 
1. Should I log-transform the data prior to imputation?
2. Increasing produced data sets / iterations
3. sensitivity analyses using complete cases only 

### Random forest 

In an attempt to predict the health impact of infections in mice, we used a random forest model @breiman2001random. We chose the maximum weight loss during the experimental laboratory infections of mice with the parasite *Eimeria spp.* as a response variable. Maximum weight loss is here used as a proxy describing the health impact caused by infections. The random forest was constructed utilizing the expression data of 20 genes, obtained by the biomarker assay (utilizing the R package “randomForest,” ntree = 1000). The data set was split into a training data set of 70 % and to a testing data set of 30 % to avoid over-fitting and to assess the performance of the model on "unseen" data. 

As a quality assessment, we used k- fold cross validation, where the data set is divided into k subsets. Each time, the model is assessed by using one of the k subsets as the test data set and the other as the training data set. The model was then fitted on each k-subset and afterwards evaluated on the test set. Last the evaluation score was noted and the model then discarded. Further, a permutation test cross-validation was implemented (using the function "rf.crossValidation" (using the R package rfUtilities, version 2.1-5). The percent variance explained from the specified fit model was 27.7%. Moreover, the mean squared error from each bootstrapped model was 44.14. Next, the variable importance was calculated according to the total decrease in node impurities from splitting on each variable, averaged over all trees. In this case of regression, the node impurity is measured by residual sum of squares. Variables with comparatively higher importance have a greater impact on the predictions of the model.




```{r cleaning prepping,  message = FALSE, include=FALSE, warning=FALSE, echo=FALSE}

############################################################# DATA Cleaning / Prep


# we need to change the  in challenge infections to a factor
hm$Parasite_challenge <- as.factor(hm$Parasite_challenge)
hm$MC.Eimeria <- as.factor(hm$MC.Eimeria)

# Here I create a new column, where we get the actual infection status
# According to the melting curve for eimeria 
hm <- hm %>%
  dplyr::mutate(current_infection = case_when(
    Parasite_challenge == "E_ferrisi" & MC.Eimeria == "TRUE" ~ "E_ferrisi",
    Parasite_challenge == "E_ferrisi" & MC.Eimeria == "FALSE" ~ "uninfected",
    Parasite_challenge == "E_falciformis" & MC.Eimeria == "TRUE" ~ "E_falciformis",
    Parasite_challenge == "E_falciformis" & MC.Eimeria == "FALSE" ~ "uninfected",
    Parasite_challenge == "uninfected" & MC.Eimeria == "TRUE" ~ "infected_eimeria",
    Parasite_challenge == "uninfected" & MC.Eimeria == "FALSE" ~ "uninfected",
    TRUE ~ ""
  ))

############################# Splitting data into training and testing sets 

# prepare the lab data
lab <- hm %>% 
  dplyr::filter(origin == "Lab")


#select the imputed gene columns
gene <-  lab %>%
  dplyr::select(c(Mouse_ID, all_of(Gene_lab)))

gene <- unique(gene)

genes <- gene %>%
  dplyr::select(-Mouse_ID)

#remove rows with only nas
genes <- genes[,colSums(is.na(genes))<nrow(genes)]

#remove colums with only nas 
genes <- genes[rowSums(is.na(genes)) != ncol(genes), ]

# select the same rows from the gene data
gene <- gene[row.names(genes),]

# select the same rows from the lab data
lab <- lab[row.names(genes),]


gene <- lab %>%
  dplyr::select(c(Mouse_ID, WL_max)) %>%
  right_join(gene, by = "Mouse_ID")

gene <- unique(gene) %>%
  dplyr::select(-Mouse_ID)


############################################# Cross validation

repeat_cv <- trainControl(method = "repeatedcv", #repeated cross validation
                           number = 5, # 5 fold cross validation
                           repeats = 3)

####################################### split data into training and test
set.seed(333) # this will help us reproduce this random assignment

# in this way we can pick the random numbers
training.samples <- createDataPartition(y = gene$WL_max, p = .7, list = FALSE) 

# this is the partiicition! In this case 0.7 = training data and 0.3 = testing
# we don't want to get a list in return
train.data <- gene[training.samples, ] 
test.data <- gene[-training.samples, ] 


############################### Building the model
set.seed(333)

#train the model
WL_predict_gene <- randomForest(WL_max ~., data = train.data, 
                                    proximity = TRUE, ntree = 1000) 

print(WL_predict_gene)

```

```{r fig4, fig.cap="Variance explained and Root Mean Squared Error", fig.env="figure*"}

###################################### ### Cross-validation 
predict_WL_cv <- rf.crossValidation(x = WL_predict_gene, xdata = train.data, 
                                    p = 0.10, n = 99, ntree = 1000)

predict_WL_cv$fit.var.exp
predict_WL_cv$fit.mse

par(mfrow=c(2,2))

plot(predict_WL_cv) 

# Root Mean Squared Error (observed vs. predicted) from each Bootstrap 
# iteration (cross-validation)
plot(predict_WL_cv, stat = "mse")

#Percent variance explained from specified fit model
plot(predict_WL_cv, stat = "var.exp")

#Mean Absolute Error from each Bootstrapped model


# number of trees with lowest MSE
which.min(WL_predict_gene$mse)

# RMSE of this optimal random forest
sqrt(WL_predict_gene$mse[which.min(WL_predict_gene$mse)])
```
```{r fig5, fig.cap="Variable Importance", fig.env="figure*"}
# Get variable importance from the WL_predict_gene fit
ImpData <- as.data.frame(importance(WL_predict_gene))
ImpData$Var.Names <- row.names(ImpData)


ImpData <- ImpData %>% arrange(desc(IncNodePurity))
print(ImpData)
```


```{r fig6, fig.cap="", fig.env="figure*"}
### Visualize variable importance 
#Call importance() function on the model model to check how the attributes used 
# as predictors affect our WL_predict_gene
#importance(WL_predict_gene)
#WL_predict_gene$mse

## S3 method for class 'randomForest'
#plot(WL_predict_gene, type = "l", main=deparse(substitute(x)))
varImpPlot(WL_predict_gene)

```













# Results 
# Discussion
# Conclusion
# Literature citations
# References



