---
title: "Untitled"
author: "Fay"
date: "28 4 2022"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
CIMouse %>% 
  filter(infection == "challenge") %>%
  group_by(Parasite_challenge) %>%
  dplyr:: summarize(total_mice = length(EH_ID),
            Mean_OOC = mean(na.omit(max_OOC)),
            Median_OOC = median(na.omit(max_OOC)),
            Min_OOC = min(na.omit(max_OOC)),
            Max_OOC = max(na.omit(max_OOC))
            )
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

