---
title: "Descriptive_immune_parameters"
author: "Fay"
date: "06/05/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
How does our immune experimental data look like?


```{r, include=FALSE}
## Load the required libraries
library(dplyr)
library(tidyverse)
library(tibble)
```

```{r}
#### Read the file
CI <- read.csv("https://raw.githubusercontent.com/derele/Eimeria_Lab/master/data_products/Challenge_infections.csv")
```
Let's add a column with the parasite names 
```{r, echo = FALSE}
CI <- CI %>%
  dplyr::mutate(Parasite_primary = case_when(
    primary_infection == "E64" ~ "Eimeria ferrisi",
    primary_infection == "E88" ~ "Eimeria falciformis",
    primary_infection == "Eflab" ~ "Eimeria falciformis",
    primary_infection == "E139" ~ "Eimeria ferrisi",
    primary_infection == "UNI" ~ "uninfected",
    TRUE ~ ""))
    
    
CI <- CI %>%
  dplyr::mutate(Parasite_challenge = case_when(    
    challenge_infection == "E64" ~ "Eimeria ferrisi",
    challenge_infection == "E88" ~ "Eimeria falciformis",
    challenge_infection == "Eflab" ~ "E. falciformis",
    challenge_infection == "E139" ~ "Eimeria ferrisi",
    challenge_infection == "UNI" ~ "uninfected",
    TRUE ~ ""))
```

### Summary statistics on experimental design

#### Summarizing data by each mouse

```{r, echo = FALSE, warning=FALSE, message=FALSE, error=FALSE}

CIMouse_gene <- CI  %>%
  dplyr::group_by(EH_ID, infection) %>%
    summarise(max_OOC = max(OOC, na.rm = TRUE),
            max_WL = min(relative_weight, na.rm = TRUE),
            experiment = unique(experiment),
            mouse_strain = unique(mouse_strain),
            primary_infection = unique(primary_infection),
            challenge_infection =unique(challenge_infection),
            infection_history=unique(infection_history),
            infection = unique(infection),
            experiment=unique(experiment),
            hybrid_status = unique(hybrid_status),
            death = unique(death),
            Parasite_primary = unique(Parasite_primary),
            Parasite_challenge = unique(Parasite_challenge),
            #gene expression
            IFNy = unique(IFNy),
            CXCR3_bio = unique(CXCR3_bio),
            IL.6 = unique(IL.6),
            IL.10 = unique(IL.10),
            IL.13 = unique(IL.13),
            IL1RN = unique(IL1RN),
            CASP1 = unique(CASP1),
            CXCL9 = unique(CXCL9),
            IDO1 = unique(IDO1),
            IRGM1 = unique(IRGM1),
            MPO = unique(MPO),
            MUC2 = unique(MUC2),
            MUC5AC = unique(MUC5AC),
            MYD88 = unique(MYD88),
            NCR1 = unique(NCR1),
            PRF1 = unique(PRF1),
            RETNLB = unique(RETNLB),
            SOCS1 = unique(SOCS1),
            TICAM1 = unique(TICAM1),
            TNF = unique(TNF))

CIMouse_facs <- CI  %>%
  dplyr::group_by(EH_ID, infection) %>%
    filter(Position == "mLN") %>%
    summarise(max_OOC = max(OOC, na.rm = TRUE),
            max_WL = min(relative_weight, na.rm = TRUE),
            experiment = unique(experiment),
            mouse_strain = unique(mouse_strain),
            primary_infection = unique(primary_infection),
            challenge_infection =unique(challenge_infection),
            infection_history=unique(infection_history),
            infection = unique(infection),
            experiment=unique(experiment),
            hybrid_status = unique(hybrid_status),
            death = unique(death),
            Parasite_primary = unique(Parasite_primary),
            Parasite_challenge = unique(Parasite_challenge),
            #cell counts
            Position = unique(Position),
            CD4 = unique(CD4),
            Treg = unique(Treg),
            Div_Treg = unique(Div_Treg),
            Treg17 = unique(Treg17),
            Th1 = unique(Th1),
            Div_Th1 = unique(Div_Th1),
            Th17 = unique(Th17),
            Div_Th17 = unique(Div_Th17),
            CD8 = unique(CD8),
            Act_CD8 = unique(Act_CD8),
            Div_Act_CD8 = unique(Div_Act_CD8),
            IFNy_CD4 = unique(IFNy_CD4),
            IFNy_CD8 = unique(IFNy_CD8),
            Treg_prop = unique(Treg_prop),
            IL17A_CD4 = unique(IL17A_CD4))

```
#### How many mice do we have in each infection ? (primary or challenge infection)
```{r, echo = FALSE}
CIMouse_gene %>% 
  group_by(infection) %>%
  dplyr::summarise(total_mice = length(EH_ID))
```

For How many mice do we have Oocyst data? 
```{r}
CIMouse_gene %>%
    dplyr::group_by(infection) %>%
    drop_na(max_OOC) %>%
    dplyr::summarise(Total_mice = length(EH_ID))
```
For how many mice do we have data on weight?
```{r}
CIMouse_gene %>%
    dplyr::group_by(infection) %>%
    drop_na(max_WL) %>%
    dplyr::summarise(Total_mice = length(EH_ID))
```
For how many mice do we have gene expression data?
```{r}
#Primary: 
CIMouse_gene %>%
    dplyr::group_by(infection) %>%
    dplyr::summarise(IFNy = length(na.omit(IFNy)),
            CXCR3_bio = length(na.omit(CXCR3_bio)),
            IL.6 = length(na.omit(IL.6)),
            IL.10 = length(na.omit(IL.10)),
            IL.13 = length(na.omit(IL.13)),
            IL1RN = length(na.omit(IL1RN)),
            CASP1 = length(na.omit(CASP1)),
            CXCL9 = length(na.omit(CXCL9)),
            IDO1 = length(na.omit(IDO1)),
            IRGM1 = length(na.omit(IRGM1)),
            MPO = length(na.omit(MPO)),
            MUC2 = length(na.omit(MUC2)),
            MUC5AC = length(na.omit(MUC5AC)),
            MYD88 = length(na.omit(MYD88)),
            NCR1 = length(na.omit(NCR1)),
            PRF1 = length(na.omit(PRF1)),
            RETNLB = length(na.omit(RETNLB)),
            SOCS1 = length(na.omit(SOCS1)),
            TICAM1 = length(na.omit(TICAM1)),
            TNF = length(na.omit(TNF)))
```



For how many mice do we have facs data? 

```{r}
CIMouse_facs %>%
    dplyr::group_by(infection) %>%
    dplyr::summarise(CD4 = length(na.omit(CD4)),
                     Treg = length(na.omit(Treg)),
                    Div_Treg = length(na.omit(Div_Treg)),
                    Treg17 = length(na.omit(Treg17)),
                    Th1 = length(na.omit(Th1)),
                    Div_Th1 = length(na.omit(Div_Th1)),
                    Th17 = length(na.omit(Th17)),
                    Div_Th17 = length(na.omit(Div_Th17)),
                    CD8 = length(na.omit(CD8)),
                    Act_CD8 = length(na.omit(Act_CD8)),
                    Div_Act_CD8 = length(na.omit(Div_Act_CD8)),
                    IFNy_CD4 = length(na.omit(IFNy_CD4)),
                    IFNy_CD8 = length(na.omit(IFNy_CD8)),
                    Treg_prop = length(na.omit(Treg_prop)),
                    IL17A_CD4 = length(na.omit(IL17A_CD4)))

```