---
title: "Challenge infections - Descreptive statistics"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Challenge infections with *Eimeria*. Here I am presenting some summary 
### statistics on the laboratory infection of mice with different *Eimeria* strains.


```{r, include=FALSE}
## Load the required libraries
library(dplyr)
library(tidyverse)
library(tibble)
```

Here is a data frame containing all information derived from the experimental infections
with *Eimeria*.
```{r}
#### Read the file
CI <- read.csv("https://raw.githubusercontent.com/derele/Eimeria_Lab/master/data_products/Challenge_infections.csv")
```

### Wrangling
```{r, echo = FALSE}
tibble::as_tibble(CI)
```


#### Summarizing data by each mouse

```{r}
CIMouse <- as_tibble(CI) %>%
  dplyr::group_by(EH_ID, infection) %>%
  dplyr::summarize(max_OOC = max(OOC),
            max_WL = min(relative_weight),
            experiment = unique(experiment),
            mouse_strain = unique(mouse_strain),
            primary_infection = unique(primary_infection),
            challenge_infection =unique(challenge_infection),
            infection_history=unique(infection_history),
            infection = unique(infection),
            experiment=unique(experiment),
            hybrid_status = unique(hybrid_status),
            death = unique(death))
```
#### Summary statistics

How many mice do we have in each infection ? (primary or challenge infection)
```{r}
CIMouse %>% 
  group_by(infection) %>%
  dplyr::summarise(total_mice = length(EH_ID))
```
How many experiments do we have? How many mice are in each experiment?
```{r}
CIMouse %>% 
  group_by(experiment) %>%
  dplyr::summarise(total_mice = length(EH_ID))
```

How many mouse strains do we have? 
```{r}
CIMouse2 <- CIMouse %>% mutate(hybrid_status = case_when(
  mouse_strain == "BUSNA_BUSNA" ~ "parental strains",
  mouse_strain == "NMRI" ~ "other",
  TRUE ~ as.character(hybrid_status)
))

CIMouse %>%
  group_by(mouse_strain, hybrid_status) %>%
  dplyr::summarise(total_mice = length(EH_ID))
```


### How many mice died in each infection
```{r}
CIMouse %>% 
  group_by(death) %>%
  summarise(length(EH_ID))
```
From the mice that died in the primary infection, what was the infection status?

```{r}
CIMouse %>%
  filter(death == "primary", infection == "primary") %>%
  group_by(primary_infection) %>%
  summarise(length(EH_ID))
```


```{r}

# Summaries for weight loss

CIMouse %>%
  #group by infections (primary or challenge infections)
  dplyr::group_by(infection) %>%
  summarize(total_mice = length(EH_ID),
            Mean_weightchange = mean(na.omit(max_WL)))
  
```
We here observe a lower weight loss average in the challenge infections when
compared to the primary infections



